### 2023-02-03 products/manuscript/Manuscript.qmd

---
title: "Example Manuscript Template for a Data Analysis Project"
subtitle: ""
author: Andreas Handel
date: "`r Sys.Date()`"
format:
  docx:
    toc: false
    number-sections: true
    highlight-style: github
bibliography: ../dataanalysis_template_references.bib
csl: ../apa.csl
---

The structure below is one possible setup for a data analysis project (including the course project). For a manuscript, adjust as needed. 
You don't need to have exactly these sections, but the content covering those sections should be addressed.

This uses MS Word as output format. [See here](https://quarto.org/docs/output-formats/ms-word.html) for more information. You can switch to other formats, like html or pdf. See [the Quarto documentation](https://quarto.org/) for other formats.

```{r, echo=FALSE, message=FALSE}
# load a few R packages
library(here)
library(knitr)
library(tidyverse)
library(broom)
library(ggtext)
```

# Summary/Abstract
_Write a summary of your project._

{{< pagebreak >}}

# Introduction

## General Background Information
_Provide enough background on your topic that others can understand the why and how of your analysis_ 
Foodborne pathogens typically cause serious gastrointestinal infections. In the U.S., one in six Americans has been reported to acquire foodborne illness annually, along with approximately 128,000 hospitalizations and 3,000 deaths (U.S. Food and Drug Administration, 2022). Economic loss associated with foodborne illnesses is estimated to be greater than $50 billion a year.  Investigating this dataset containing foodborne outbreak information in the 17-year period may reveal the frequent sources and the common etiologic agents that are related to foodborne illnesses. Ultimately, measures can be conducted or adjusted based on analysis for prevention of outbreaks.
## Description of data and data source
_Describe what the data is, what it contains, where it is from, etc. Eventually this might be part of a methods section._

## Questions/Hypotheses to be addressed
_State the research questions you plan to answer with this analysis._

To cite other work (important everywhere, but likely happens first in introduction), make sure your references are in the bibtex file specified in the YAML header above (here `dataanalysis_template_references.bib`) and have the right bibtex key. Then you can include like this:

Examples of reproducible research projects can for instance be found in [@mckay2020; @mckay2020a]

{{< pagebreak >}}

# Methods

_Describe your methods. That should describe the data, the cleaning processes, and the analysis approaches. You might want to provide a shorter description here and all the details in the supplement._
1. Data wrangling: handling the missing values and cleaning inconsistent data.
2. Selecting the interested predictors and outcomes to plot their relationships.
3. If the graphs indicate any significant or interesting trend, we will conduct a general linear model and omnibus test.
4. Post-hoc test to identify the significant difference between levels.
5. Mapping cases with different predictors across the states.

## Data aquisition
_As applicable, explain where and how you got the data. If you directly import the data from an online source, you can combine this section with the next._

## Data import and cleaning
_Write code that reads in the file and cleans it so it's ready for analysis. Since this will be fairly long code for most datasets, it might be a good idea to have it in one or several R scripts. If that is the case, explain here briefly what kind of cleaning/processing you do, and provide more details and well documented code somewhere (e.g. as supplement in a paper). All materials, including files that contain code, should be commented well so everyone can follow along._

## Statistical analysis
_Explain anything related to your statistical analyses._

{{< pagebreak >}}

# Results

## Exploratory/Descriptive analysis

_Use a combination of text/tables/figures to explore and describe your data. Show the most important descriptive results here. Additional ones should go in the supplement. Even more can be in the R and Quarto files that are part of your project._

@tbl-summarytable shows a summary of the data.

Note the loading of the data providing a **relative** path using the `../../` notation. (Two dots means a folder up). You never want to specify an **absolute** path like `C:\ahandel\myproject\results\` because if you share this with someone, it won't work for them since they don't have that path. You can also use the `here` R package to create paths. See examples of that below.

```{r}
#| label: tbl-summarytable
#| tbl-cap: "Data summary table."
#| echo: FALSE
resulttable=readRDS("../../results/summarytable.rds")
knitr::kable(resulttable)
```

## Basic statistical analysis

_To get some further insight into your data, if reasonable you could compute simple statistics (e.g. simple models with 1 predictor) to look for associations between your outcome(s) and each individual predictor variable. Though note that unless you pre-specified the outcome and main exposure, any "p<0.05 means statistical significance" interpretation is not valid._

@fig-result shows a scatterplot figure produced by one of the R scripts.

```{r}
#| label: fig-result
#| fig-cap: "Height and weight stratified by sex."
#| echo: FALSE
knitr::include_graphics(here("results","height_weight_stratified.png"))
```

## Full analysis

_Use one or several suitable statistical/machine learning methods to analyze your data and to produce meaningful figures, tables, etc. This might again be code that is best placed in one or several separate R scripts that need to be well documented. You want the code to produce figures and data ready for display as tables, and save those. Then you load them here._

Example @tbl-resulttable2 shows a summary of a linear model fit.

```{r}
#| label: tbl-resulttable2
#| tbl-cap: "Linear model fit table."
#| echo: FALSE
resulttable2 = readRDS(here("results","resulttable2.rds"))
knitr::kable(resulttable2)
```

{{< pagebreak >}}

# Discussion

## Summary and Interpretation
_Summarize what you did, what you found and what it means._

## Strengths and Limitations
_Discuss what you perceive as strengths and limitations of your analysis._

## Conclusions
_What are the main take-home messages?_

_Include citations in your Rmd file using bibtex, the list of references will automatically be placed at the end_

This paper [@leek2015] discusses types of analyses.

These papers [@mckay2020; @mckay2020a] are good examples of papers published using a fully reproducible setup similar to the one shown in this template.

Note that this cited reference will show up at the end of the document, the reference formatting is determined by the CSL file specified in the YAML header. Many more style files for almost any journal [are available](https://www.zotero.org/styles). You also specify the location of your bibtex reference file in the YAML. You can call your reference file anything you like, I just used the generic word `references.bib` but giving it a more descriptive name is probably better.

{{< pagebreak >}}

# References

### 2023-02-03 products/manuscript/Manuscript.qmd
Project Update: 2/3/2023 12:51 PM EST

@@
     highlight-style: github
-bibliography: ../dataanalysis_template_references.bib
+bibliography: ../references.bib
 csl: ../apa.csl
+editor: 
+  markdown: 
+    wrap: sentence
 ---
 
-The structure below is one possible setup for a data analysis project (including the course project). For a manuscript, adjust as needed. 
+The structure below is one possible setup for a data analysis project (including the course project).
+For a manuscript, adjust as needed.
 You don't need to have exactly these sections, but the content covering those sections should be addressed.
@@
 # Summary/Abstract
-_Write a summary of your project._
+
+*Write a summary of your project.*
 
@@
 ## General Background Information
-_Provide enough background on your topic that others can understand the why and how of your analysis_ 
-Foodborne pathogens typically cause serious gastrointestinal infections.
-In the U.S., one in six Americans has been reported to acquire foodborne illness annually, along with approximately 128,000 hospitalizations and 3,000 deaths (U.S.
-Food and Drug Administration, 2022).
-Economic loss associated with foodborne illnesses is estimated to be greater than $50 billion a year.
+
+*Provide enough background on your topic that others can understand the why and how of your analysis* Foodborne pathogens typically cause serious gastrointestinal infections.
+In the U.S., one in six Americans has been reported to acquire foodborne illness annually, along with approximately 128,000 hospitalizations and 3,000 deaths @centerforfoodsafetyandappliednutrition2022.
+Economic loss associated with foodborne illnesses is estimated to be greater than \$50 billion a year.
 Investigating this dataset containing foodborne outbreak information in the 17-year period may reveal the frequent sources and the common etiologic agents that are related to foodborne illnesses.
 Ultimately, measures can be conducted or adjusted based on analysis for prevention of outbreaks.
+
 ## Description of data and data source
-_Describe what the data is, what it contains, where it is from, etc. Eventually this might be part of a methods section._
+
+*Describe what the data is, what it contains, where it is from, etc. Eventually this might be part of a methods section.*
+
+This dataset was uploaded onto Kaggle in 2017 and contains data from the Centers for Disease Control and Prevention (CDC) on foodborne disease outbreaks from 1998 to 2015 @centersfordiseasecontrolandprevention2017.
+There are 12 variables and 16,950+ observations.
+The raw data may be viewed within the repository of this project.
 
 ## Questions/Hypotheses to be addressed
-_State the research questions you plan to answer with this analysis._
 
+*State the research questions you plan to answer with this analysis.* We have several questions we would like this data to answer: (1) do foodborne disease outbreaks occur more often in restaurants or in private residences, (2) what pathogen species are most frequently associated with foodborne outbreaks, and (3) what type of food is most commonly involved in foodborne outbreaks.
+The outcomes of interest would therefore be location (for whether more outbreaks occur in restaurants/homes), species of pathogen, and food type most often associated with outbreaks.
+
+We expect to see that outbreaks occur most often in homes rather than restaurants, the pathogen most commonly seen to be E.
+coli, and for the food most often associated with outbreaks to be some form of salad.
+
+### Reminder
 To cite other work (important everywhere, but likely happens first in introduction), make sure your references are in the bibtex file specified in the YAML header above (here `dataanalysis_template_references.bib`) and have the right bibtex key.
@@
 
-_Describe your methods. That should describe the data, the cleaning processes, and the analysis approaches. You might want to provide a shorter description here and all the details in the supplement._
+*Describe your methods.
+That should describe the data, the cleaning processes, and the analysis approaches.
+You might want to provide a shorter description here and all the details in the supplement.* 1.
 1. Data wrangling: handling the missing values and cleaning inconsistent data.
@@
 ## Data aquisition
-_As applicable, explain where and how you got the data. If you directly import the data from an online source, you can combine this section with the next._
+
+*As applicable, explain where and how you got the data. If you directly import the data from an online source, you can combine this section with the next.*
 
 ## Data import and cleaning
-_Write code that reads in the file and cleans it so it's ready for analysis.
+
+*Write code that reads in the file and cleans it so it's ready for analysis.
 Since this will be fairly long code for most datasets, it might be a good idea to have it in one or several R scripts.
@@
 as supplement in a paper).
-All materials, including files that contain code, should be commented well so everyone can follow along._
+All materials, including files that contain code, should be commented well so everyone can follow along.*
 
 ## Statistical analysis
-_Explain anything related to your statistical analyses._
+
+*Explain anything related to your statistical analyses.*
 
@@
 
-_Use a combination of text/tables/figures to explore and describe your data.
+*Use a combination of text/tables/figures to explore and describe your data.
 Show the most important descriptive results here.
 Additional ones should go in the supplement.
-Even more can be in the R and Quarto files that are part of your project._
+Even more can be in the R and Quarto files that are part of your project.*
 
@@
 
-_To get some further insight into your data, if reasonable you could compute simple statistics (e.g.
+*To get some further insight into your data, if reasonable you could compute simple statistics (e.g.
 simple models with 1 predictor) to look for associations between your outcome(s) and each individual predictor variable.
-Though note that unless you pre-specified the outcome and main exposure, any "p<0.05 means statistical significance" interpretation is not valid._
+Though note that unless you pre-specified the outcome and main exposure, any "p\<0.05 means statistical significance" interpretation is not valid.*
 
@@
 
-_Use one or several suitable statistical/machine learning methods to analyze your data and to produce meaningful figures, tables, etc.
+*Use one or several suitable statistical/machine learning methods to analyze your data and to produce meaningful figures, tables, etc.
 This might again be code that is best placed in one or several separate R scripts that need to be well documented.
 You want the code to produce figures and data ready for display as tables, and save those.
-Then you load them here._
+Then you load them here.*
 
@@
 ## Summary and Interpretation
-_Summarize what you did, what you found and what it means._
+
+*Summarize what you did, what you found and what it means.*
 
 ## Strengths and Limitations
-_Discuss what you perceive as strengths and limitations of your analysis._
+
+*Discuss what you perceive as strengths and limitations of your analysis.*
 
 ## Conclusions
-_What are the main take-home messages?_
 
-_Include citations in your Rmd file using bibtex, the list of references will automatically be placed at the end_
+*What are the main take-home messages?*
+
+*Include citations in your Rmd file using bibtex, the list of references will automatically be placed at the end*
 

### 2023-02-03 products/manuscript/Manuscript.qmd
Update Manuscript.qmd

@@
 
-*Provide enough background on your topic that others can understand the why and how of your analysis* Foodborne pathogens typically cause serious gastrointestinal infections.
+*Provide enough background on your topic that others can understand the why and how of your analysis.*
+
+Foodborne pathogens typically cause serious gastrointestinal infections.
 In the U.S., one in six Americans has been reported to acquire foodborne illness annually, along with approximately 128,000 hospitalizations and 3,000 deaths @centerforfoodsafetyandappliednutrition2022.
@@
 
-*State the research questions you plan to answer with this analysis.* We have several questions we would like this data to answer: (1) do foodborne disease outbreaks occur more often in restaurants or in private residences, (2) what pathogen species are most frequently associated with foodborne outbreaks, and (3) what type of food is most commonly involved in foodborne outbreaks.
+*State the research questions you plan to answer with this analysis.*
+
+We have several questions we would like this data to answer: (1) do foodborne disease outbreaks occur more often in restaurants or in private residences, (2) what pathogen species are most frequently associated with foodborne outbreaks, and (3) what type of food is most commonly involved in foodborne outbreaks.
 The outcomes of interest would therefore be location (for whether more outbreaks occur in restaurants/homes), species of pathogen, and food type most often associated with outbreaks.
 
-We expect to see that outbreaks occur most often in homes rather than restaurants, the pathogen most commonly seen to be E.
-coli, and for the food most often associated with outbreaks to be some form of salad.
+We expect to see that outbreaks occur most often in homes rather than restaurants, the pathogen most commonly seen to be *E.
+coli*, and for the food most often associated with outbreaks to be some form of salad.
 
 ### Reminder
+
 To cite other work (important everywhere, but likely happens first in introduction), make sure your references are in the bibtex file specified in the YAML header above (here `dataanalysis_template_references.bib`) and have the right bibtex key.
@@
 
-*Describe your methods.
-That should describe the data, the cleaning processes, and the analysis approaches.
-You might want to provide a shorter description here and all the details in the supplement.* 1.
+*Describe your methods. That should describe the data, the cleaning processes, and the analysis approaches. You might want to provide a shorter description here and all the details in the supplement.*
+
 1.  Data wrangling: handling the missing values and cleaning inconsistent data.
@@
 
-## Data aquisition
+## Data acquisition
 
 *As applicable, explain where and how you got the data. If you directly import the data from an online source, you can combine this section with the next.*
+
+As previously mentioned, the data was acquired from [Kaggle](https://www.kaggle.com/datasets/cdc/foodborne-diseases).
 

### 2023-02-22 code/processing_code/processingfile_v2.qmd
2

---
title: "An example cleaning script with code pulled in"
author: "Andreas Handel"
date: "2023-01-03"
output: html_document
---

# Processing script with code pulled in

This is essentially the same as the other Quarto file, but now the code is not inside this file. Instead, it is pulled in from the R script using the code chunk labels.

# Setup

This needs to run to load the R script so we can include the code chunks below.

Note that you can pull in more than one R script if you want.

```{r, include=FALSE, cache=FALSE}
knitr::read_chunk('processingcode.R')
```

Load needed packages. Instead of having R commands here, this just pulls in the code from the R script we loaded above and labelled.

```{r, packages, message = FALSE, warning = FALSE}
library(readxl) #for loading excel files
library(dplyr) #for data processing/cleaning
library(tidyr) #for data processing/cleaning
library(skimr) #for nice visualization of data 
library(here) #to set paths
library(dlookr)
```

# Data loading

Note that for functions that come from specific packages (instead of base R), I often specify both package and function like so:
package::function() that's not required one could just call the function specifying the package makes it clearer where the function "lives",
but it adds typing. You can do it either way.

```{r, loaddata}
data_location<-here("data","raw_data","NationalOutbreakPublicDataTool.xlsx")
rawdata2=read_excel(data_location)

```

# Explore data

Several ways of looking at the data

```{r, exploredata, fig.height=8, fig.width=8}
glimpse(rawdata2)
summary(rawdata2)
head(rawdata2)
skim(rawdata2)
plot_na_pareto(rawdata2)
```

Looks like we have the following data:

* height (seems like it's in centimeters) 
* weight (seems to be in kilogram)
* Sex

# Cleaning

By inspecting the data as done above, we find some problems that need addressing:

First, there is an entry for height which says "sixty" instead of a number. 
Does that mean it should be a numeric 60? It somehow doesn't make sense since the weight is 60kg, which can't happen for a 60cm person (a baby).
Since we don't know how to fix this, we might decide to remove the person. This "sixty" entry also turned all Height entries into characters instead of numeric. That conversion to character also means that our summary function isn't very meaningful. So let's fix that first.

```{r, cleandata1,fig.height=8, fig.width=8}
d2=rawdata2%>%
  filter(`Primary Mode`=="Food")
plot_na_pareto(d2)
```

Now we see that there is one person with a height of 6. That could be a typo, or someone mistakenly entered their height in feet. Since we unfortunately don't know, we might need to remove this person, which we'll do here.

```{r, cleandata2}
```

Height values seem ok now.

Now let's look at the `Weight` variable. There is a person with weight of 7000, which is impossible, and one person with missing weight.
To be able to analyze the data, we'll remove those individuals as well.

Note that removing anyone who had "faulty" or missing data is one approach. It's often not the best. Based on your question and your analysis approach, you might want to do cleaning differently (e.g. keep individuals with some missing information).

```{r, cleandata3}
```

Now checking the `Sex` variable. Sex should be a categorical/factor variable but is loaded as character. We can fix that with simple base R code to mix things up.

```{r, cleandata4}
```

Now we see that there is another NA, but it's not `NA` from R, instead it was loaded as character and is now considered as a category.
Well proceed here by removing that individual with that NA entry. Since this keeps an empty category for Sex, I'm also using droplevels() to get rid of it.

```{r, cleandata5}
```

# Save data

All done, data is clean now. Time to save.
Let's assign at the end to some final variable, this makes it easier to add further cleaning steps above.

Finally, we save the clean data as RDS file. I suggest you save your processed and cleaned data as RDS or RDA/Rdata files. 
This preserves coding like factors, characters, numeric, etc.  If you save as CSV, that information would get lost.
However, CSV is better for sharing with others since it's plain text. If you do CSV, you might want to write down somewhere what each variable is.

See here for some suggestions on how to store your processed data:
http://www.sthda.com/english/wiki/saving-data-into-r-data-format-rds-and-rdata

```{r, savedata}
```

### 2023-02-23 code/processing_code/processingfile_v1.qmd
3

--- 2023-02-22 code/processing_code/processingfile_v2.qmd
@@
 ---
-title: "An example cleaning script with code pulled in"
+title: "cleaning script for foodborne_outbreaks.csv"
 author: "Andreas Handel"
@@
 
-# Processing script with code pulled in
-
-This is essentially the same as the other Quarto file, but now the code is not inside this file. Instead, it is pulled in from the R script using the code chunk labels.
-
 # Setup
-
-This needs to run to load the R script so we can include the code chunks below.
-
-Note that you can pull in more than one R script if you want.
-
-```{r, include=FALSE, cache=FALSE}
-knitr::read_chunk('processingcode.R')
-```
-
-Load needed packages. Instead of having R commands here, this just pulls in the code from the R script we loaded above and labelled.
-
-```{r, packages, message = FALSE, warning = FALSE}
-library(readxl) #for loading excel files
+```{r}
+library(readr) #for loading csv files
 library(dplyr) #for data processing/cleaning
@@
 
-```{r, loaddata}
-data_location<-here("data","raw_data","NationalOutbreakPublicDataTool.xlsx")
-rawdata2=read_excel(data_location)
-
+```{r}
+# path to data
+# note the use of the here() package and not absolute paths
+data_location <- here::here("data","raw_data","foodborne_outbreaks.csv")
+rawdata1 <- read_csv(data_location)
 ```
 
-# Explore data
+# Check data
 
@@
 
-```{r, exploredata, fig.height=8, fig.width=8}
-glimpse(rawdata2)
-summary(rawdata2)
-head(rawdata2)
-skim(rawdata2)
-plot_na_pareto(rawdata2)
+```{r}
+str(rawdata1)
+summary(rawdata1)
+head(rawdata1)
+skim(rawdata1)
+plot_na_pareto(rawdata1)
 ```
 
-Looks like we have the following data:
+Looks like we have the following predictor variables:
 
-* height (seems like it's in centimeters) 
-* weight (seems to be in kilogram)
-* Sex
+* Year, Month (date)
+* Sate, Location (geographic information)
+* Food, Ingredient
+*Species, Serotype/Genotype
+*Status
+
+Following variables are outcomes:
+*illnesses
+*Hospitalizations
+*Fatalities
 
@@
 
-First, there is an entry for height which says "sixty" instead of a number. 
+First, we will remove the variable "ingredient" and "serotype" since over 80% of data is missing. 
 Does that mean it should be a numeric 60? It somehow doesn't make sense since the weight is 60kg, which can't happen for a 60cm person (a baby).
@@
 
-```{r, cleandata1,fig.height=8, fig.width=8}
-d2=rawdata2%>%
-  filter(`Primary Mode`=="Food")
-plot_na_pareto(d2)
+```{r}
+d1=
 ```
@@
 
-```{r, cleandata2}
+```{r}
+d2 <- d1 %>% dplyr::filter(Height != 6)
+skimr::skim(d2)
 ```
@@
 
-Note that removing anyone who had "faulty" or missing data is one approach.
-It's often not the best.
-Based on your question and your analysis approach, you might want to do cleaning differently (e.g.
-keep individuals with some missing information).
-
-```{r, cleandata3}
+```{r}
+d3 <- d2 %>%  dplyr::filter(Weight != 7000) %>% tidyr::drop_na()
+skimr::skim(d3)
 ```
@@
 
-```{r, cleandata4}
+```{r}
+d3$Sex <- as.factor(d3$Sex)  
+skimr::skim(d3)
 ```
@@
 
-```{r, cleandata5}
+```{r}
+d4 <- d3 %>% dplyr::filter(Sex != "NA") %>% droplevels()
+skimr::skim(d4)
+```
+
+All done, data is clean now.
+
+Let's assign at the end to some final variable, this makes it easier to add further cleaning steps above.
+
+```{r}
+processeddata <- d4
 ```
@@
 # Save data
-
-All done, data is clean now. Time to save.
-Let's assign at the end to some final variable, this makes it easier to add further cleaning steps above.
 
@@
 
-```{r, savedata}
+```{r}
+save_data_location <- here::here("data","processed_data","processeddata.rds")
+saveRDS(processeddata, file = save_data_location)
 ```
+
+# Notes
+
+Removing anyone who had "faulty" or missing data is one approach.
+It's often not the best.
+based on your question and your analysis approach, you might want to do cleaning differently (e.g.
+keep individuals with some missing information).

### 2023-02-23 code/processing_code/processingfile_v2.qmd
3

@@
 ---
-title: "An example cleaning script with code pulled in"
+title: "cleaning script for NationalOutbreakPublicDataTool.xlsx"
 author: "Andreas Handel"
@@
 
-# Processing script with code pulled in
-
-This is essentially the same as the other Quarto file, but now the code is not inside this file. Instead, it is pulled in from the R script using the code chunk labels.
-
 # Setup
 
-This needs to run to load the R script so we can include the code chunks below.
-
-Note that you can pull in more than one R script if you want.
-
 ```{r, include=FALSE, cache=FALSE}
-knitr::read_chunk('processingcode.R')
-```
-
-Load needed packages. Instead of having R commands here, this just pulls in the code from the R script we loaded above and labelled.
-
-```{r, packages, message = FALSE, warning = FALSE}
-library(readxl) #for loading excel files
+library(readr) #for loading csv files
 library(dplyr) #for data processing/cleaning
@@
 library(here) #to set paths
-library(dlookr)
+```
+
+```{r, packages, message = FALSE, warning = FALSE}
 ```
@@
 ```{r, loaddata}
-data_location<-here("data","raw_data","NationalOutbreakPublicDataTool.xlsx")
-rawdata2=read_excel(data_location)
-
+data_location
 ```
@@
 
-```{r, exploredata, fig.height=8, fig.width=8}
-glimpse(rawdata2)
-summary(rawdata2)
-head(rawdata2)
-skim(rawdata2)
-plot_na_pareto(rawdata2)
+```{r, exploredata}
+dplyr::glimpse(rawdata)
+summary(rawdata)
+head(rawdata)
+skimr::skim(rawdata)
 ```
@@
 
-```{r, cleandata1,fig.height=8, fig.width=8}
-d2=rawdata2%>%
-  filter(`Primary Mode`=="Food")
-plot_na_pareto(d2)
+```{r, cleandata1}
 ```

### 2023-02-23 code/processing_code/processingfile_v1.qmd
4

@@
 ```{r}
+library(tidyverse)
 library(readr) #for loading csv files
@@
 
-* Year, Month (date)
-* Sate, Location (geographic information)
-* Food, Ingredient
-*Species, Serotype/Genotype
+* Year, Month (numeric)
+* State, Location (geographic information)
+* Food, -Ingredient
+*Species, -Serotype/Genotype
 *Status
@@
 *Fatalities
+main outcome: illnesses
 
@@
 So let's fix that first.
-
+#potential code for EDA
+#change month to numeric data
 ```{r}
-d1=
+rawdata1%>%filter(grepl("Confirmed", Status))
+unique(rawdata1$Food)
+d2=rawdata1%>%
+  filter(!is.na(Food))%>%
+  mutate(Food=fct_lump(Food,10))%>%
+  count(Food)%>%
+  arrange(desc(n))
+d2
+d3=rawdata1%>%
+  filter(!is.na(Species))%>%
+  mutate(Food=fct_collapse(Food,Norovirus=c("Norovirus","Norovirus genogroup I","Norovirus genogroup II")))%>%
+  mutate(Food=fct_lump(Species,10))%>%
+  count(Species)%>%
+  arrange(desc(n))
+d3
 ```

### 2023-02-23 code/processing_code/processingfile_v1.qmd
5

--- 2023-02-23 code/processing_code/processingfile_v1.qmd
@@
 
+install.packages("impute")
 # Setup
 ```{r}
+library(tidyverse)
 library(readr) #for loading csv files
@@
 library(dlookr)
+library(caret)
+library(randomForest)
 ```
@@
 
-* Year, Month (date)
-* Sate, Location (geographic information)
+* Year, Month (numeric)
+* State, Location (geographic information)
 * Food, Ingredient
-*Species, Serotype/Genotype
-*Status
+* Species, Serotype/Genotype
+* Status
 
 Following variables are outcomes:
-*illnesses
-*Hospitalizations
-*Fatalities
+* illnesses
+* Hospitalizations
+* Fatalities
+
+main outcome: illnesses
 
@@
 
-First, we will remove the variable "ingredient" and "serotype" since over 80% of data is missing. 
-Does that mean it should be a numeric 60? It somehow doesn't make sense since the weight is 60kg, which can't happen for a 60cm person (a baby).
-Since we don't know how to fix this, we might decide to remove the person.
-This "sixty" entry also turned all Height entries into characters instead of numeric.
-That conversion to character also means that our summary function isn't very meaningful.
-So let's fix that first.
+First, we will remove the variable "ingredient" and "serotype/genotype" since over 80% of data is missing. 
+```{r}
+d1=rawdata1%>%
+  select(-c(Ingredient,`Serotype/Genotype`))
+skim(d1)
+```
+#checking the levels under each variable
+#checking the levels under `Food`
+```{r}
+skim(d1$Food)
+#convert Food to factor
+d1$Food=as.factor(d1$Food)
+str(d1$Food)
+#Since there are over 3000 different levels, we need to combine some food into the same category
+d2=
+```
+#checking the levels under `Species`
+```{r}
 
+```
+#checking the levels under `Location`
 ```{r}
-d1=
+
 ```
 
-Now we see that there is one person with a height of 6.
-That could be a typo, or someone mistakenly entered their height in feet.
-Since we unfortunately don't know, we might need to remove this person, which we'll do here.
+#potential code for EDA
+```{r}
+rawdata1%>%filter(grepl("Confirmed", Status))
+food=rawdata1%>%
+  filter(!is.na(Food))%>%
+  group_by(Food)%>%
+  top_n(n=10)
+food
+species=rawdata1%>%
+  filter(!is.na(Species))%>%
+  mutate(Species=fct_collapse(Species,Norovirus=c("Norovirus","Norovirus genogroup I","Norovirus genogroup II","Norovirus unknown")))%>%
+  mutate(Species=fct_lump(Species,11,other_level = NA))%>%
+  count(Species)%>%
+  arrange(desc(n))%>%
+  filter(!is.na(Species))
+species
+```
+#predict missing value
+```{r}
+set.seed(123)
+index=createDataPartition(rawdata1$State,p=0.8,list=FALSE)
+train=rawdata1[index,]
+test=rawdata1[-index,]
+model1=randomForest(State~Year+Illnesses+Food,data=train,na.action = na.roughfix)
+predicted_states <- predict(model, newdata = test)
+actual_states <- test$State
+accuracy <- sum(predicted_states == actual_states) / length(actual_states)
 
-```{r}
-d2 <- d1 %>% dplyr::filter(Height != 6)
-skimr::skim(d2)
 ```
-
-Height values seem ok now.
 
@@
 d3 <- d2 %>%  dplyr::filter(Weight != 7000) %>% tidyr::drop_na()
-skimr::skim(d3)
-```
-
-Now checking the `Sex` variable. Sex should be a categorical/factor variable but is loaded as character. We can fix that with simple base R code to mix things up.
-
-```{r}
-d3$Sex <- as.factor(d3$Sex)  
 skimr::skim(d3)
@@
 
-Finally, we save the clean data as RDS file. I suggest you save your processed and cleaned data as RDS or RDA/Rdata files. 
-This preserves coding like factors, characters, numeric, etc.  If you save as CSV, that information would get lost.
-However, CSV is better for sharing with others since it's plain text. If you do CSV, you might want to write down somewhere what each variable is.
-
-See here for some suggestions on how to store your processed data:
-http://www.sthda.com/english/wiki/saving-data-into-r-data-format-rds-and-rdata
+Finally, we save the clean data as RDS file.
 
 ```{r}
-save_data_location <- here::here("data","processed_data","processeddata.rds")
-saveRDS(processeddata, file = save_data_location)
+save_data_location <- here::here("data","processed_data","processeddata1.rds")
+saveRDS(processeddata1, file = save_data_location)
 ```

### 2023-02-23 code/processing_code/processingfile_v1.qmd
working on part 2 of project

--- 2023-02-23 code/processing_code/processingfile_v1.qmd
@@
 ---
-title: "cleaning script for foodborne_outbreaks.csv"
+title: "Data Pre-Processing/Cleaning Script"
 author: "Andreas Handel"
@@
 
-# Setup
+# Load Libraries
 ```{r}
-library(readr) #for loading csv files
-library(dplyr) #for data processing/cleaning
-library(tidyr) #for data processing/cleaning
-library(skimr) #for nice visualization of data 
-library(here) #to set paths
+# General Pre-Processing
+library(tidyverse)
+library(dplyr)
+
+# Data Structure Visualization
+library(skimr)
+
+# Setting Paths
+library(here)
+
+# Information About NAs
 library(dlookr)
@@
 
-# Data loading
-
-Note that for functions that come from specific packages (instead of base R), I often specify both package and function like so:
-package::function() that's not required one could just call the function specifying the package makes it clearer where the function "lives",
-but it adds typing. You can do it either way.
+# Loading Data
 
 ```{r}
-# path to data
-# note the use of the here() package and not absolute paths
-data_location <- here::here("data","raw_data","foodborne_outbreaks.csv")
-rawdata1 <- read_csv(data_location)
+# Path to Data
+data_location <- here("data","raw_data","foodborne_outbreaks.csv")
+training_data <- read_csv(data_location)
+
+data_location <- here(
+  "data","raw_data","NationalOutbreakPublicDataTool.xlsx")
+analysis_data <- readxl::read_excel(data_location)
 ```
 
-# Check data
+# Check Data
 
-Several ways of looking at the data
+There are several ways we can look at the datasets' structures.
+
+## Training Data
+```{r}
+str(training_data)
+summary(training_data)
+
+# Closer Look at Training Data
+head(training_data)
+skim(training_data)
+
+# Plotting NAs of Training Data
+plot_na_pareto(training_data)
+```
+
+It appears that the training data contains the following variables:
+General
+-   Year, Month (Time)
+-   State (Location)
+-   Food, Ingredient (Vehicle of Transmission)
+-   Species
+-   Status
+Health Outcomes
+-   Illnesses
+-   Hospitalizations
+-   Fatalities
+
+## Analysis Data
+```{r}
+str(analysis_data)
+summary(analysis_data)
+
+# Closer Look at Training Data
+head(analysis_data)
+skim(analysis_data)
+
+# Plotting NAs of Training Data
+plot_na_pareto(analysis_data)
+```
+
+It appears that the training data contains the following variables:
+General
+-   Year, Month (Time)
+-   State (Location)
+-   Primary Mode (Vehicle of Transmission)
+-   Food Vehicle
+-   Food Contaminated Ingredient
+-   Water Exposure
+-   Water Type
+-   Water Status
+-   Animal Type
+-   Animal Type Specify
+-   Etiology
+-   Serotype or Genotype
+-   Etiology Status
+-   Setting
+-   IFSAC Category
+Health Outcomes
+-   Illnesses
+-   Hospitalizations
+-   Info on Hospitalizations
+-   Deaths
+-   Info on Deaths
+
+Because the analysis dataset contains more than just foodborne outbreaks, we want to specify it to just foodborne illnesses.
 
 ```{r}
-str(rawdata1)
-summary(rawdata1)
-head(rawdata1)
-skim(rawdata1)
-plot_na_pareto(rawdata1)
+# Plot of NAs After Limiting to Analysis Data to Food
+plot_na_pareto(analysis_data %>% filter(`Primary Mode` == "Food"))
+```
+## Checking Etiologies
+```{r}
+# Training Data
+unique(training_data$Status)
+
+# Analysis Data
+unique(analysis_data$`Etiology Status`)
 ```
 
-Looks like we have the following predictor variables:
-
-* Year, Month (date)
-* Sate, Location (geographic information)
-* Food, Ingredient
-*Species, Serotype/Genotype
-*Status
-
-Following variables are outcomes:
-*illnesses
-*Hospitalizations
-*Fatalities
+Given that "Status" and "Etiology Status" refer to laboratory confirmations, we will filter our case definitions to be only laboratory-confirmed pathogens.
 
@@
 
-By inspecting the data as done above, we find some problems that need addressing:
+Based on the data inspections completed above, we want to remove "Ingredient" and "Serotype/Genotype" from the training dataset (> 80% data missing), and we want to remove all variables unrelated to foodborne illness for the analysis dataset (ex: animal type, water type, etc.) in addition to columns with many NAs.
 
-First, we will remove the variable "ingredient" and "serotype" since over 80% of data is missing. 
-Does that mean it should be a numeric 60? It somehow doesn't make sense since the weight is 60kg, which can't happen for a 60cm person (a baby).
-Since we don't know how to fix this, we might decide to remove the person.
-This "sixty" entry also turned all Height entries into characters instead of numeric.
-That conversion to character also means that our summary function isn't very meaningful.
-So let's fix that first.
+## Training Data
+```{r}
+training_data_fin <- training_data %>%
+  select(!c(Ingredient, `Serotype/Genotype`)) %>%
+  filter(grepl("Confirmed", Status))
+```
+
+## Analysis Data
+```{r}
+analysis_data_fin <- analysis_data %>%
+  select(!c(`Animal Type`, `Animal Type Specify`, 
+            `Water Exposure`, `Water Type`, `Water Status`, 
+            `Food Contaminated Ingredient`)) %>%
+  filter(grepl("Confirmed", `Etiology Status`))
+```
+
+#potential code for EDA
+
+## Data Manipulation
+### Separate Rows to Address Coinfections
+```{r}
+analysis_data_fin <- analysis_data_fin %>%
+  separate_rows(Etiology, sep = "; ")
+```
+
+### Column for Type of Pathogen
+```{r}
+Pathogen_Bacteria <- c("Bacillus", "Brucella", "Campylobacter", "Clostridium", "Staphylococcus", "Escherichia coli", "Listeria", "Salmonella", "Vibrio", "Cholera", "Enterococcus", "Bacterium", "Shigella", "Yersinia")
+Pathogen_Parasite <- c("Cryptosporidium", "Cyclospora", "Entamoeba", "Giardia", "Parasite", "Toxoplasma", "Trichinella")
+Pathogen_Virus <- c("Atrovirus", "Norovirus", "Rotavirus")
+Pathogen_Poison <- c("Poison", "Toxin", "Heavy Metal")
+
+analysis_data_fin %>%
+  mutate(Pathogen_Type = case_when(
+    grepl(paste(Pathogen_Bacteria, 
+                collapse = "|"), Etiology) ~ "Bacteria",
+    grepl(paste(Pathogen_Parasite, 
+                collapse = "|"), Etiology) ~ "Parasite",
+    grepl(paste(Pathogen_Virus, 
+                collapse = "|"), Etiology) ~ "Virus",
+    grepl(paste(Pathogen_Poison, 
+                collapse = "|"), Etiology) ~ "Poison")) %>%
+  mutate(Etiology = case_when(
+    Pathogen_Type == "Bacteria" ~
+
+      grepl(paste(Pathogen_Bacteria, collapse = "| "), Etiology)
+  ))
+
+```
+
+### Add Month as Numeric Column
+```{r}
+training_data_fin <- training_data_fin %>%
+  mutate(Numeric_Month = case_when(
+    Month == "January" ~ 1,
+    Month == "February" ~ 2,
+    Month == "March" ~ 3,
+    Month == "April" ~ 4,
+    Month == "May" ~ 5,
+    Month == "June" ~ 6,
+    Month == "July" ~ 7,
+    Month == "August" ~ 8,
+    Month == "September" ~ 9,
+    Month == "October" ~ 10,
+    Month == "November" ~ 11,
+    Month == "December" ~ 12)) %>%
+  relocate(Numeric_Month, .after = Month)
+```
 
 ```{r}
-d1=
-```
-
-Now we see that there is one person with a height of 6.
-That could be a typo, or someone mistakenly entered their height in feet.
-Since we unfortunately don't know, we might need to remove this person, which we'll do here.
-
-```{r}
-d2 <- d1 %>% dplyr::filter(Height != 6)
-skimr::skim(d2)
-```
-
-Height values seem ok now.
-
-Now let's look at the `Weight` variable. There is a person with weight of 7000, which is impossible, and one person with missing weight.
-To be able to analyze the data, we'll remove those individuals as well.
-
-```{r}
-d3 <- d2 %>%  dplyr::filter(Weight != 7000) %>% tidyr::drop_na()
-skimr::skim(d3)
-```
-
-Now checking the `Sex` variable. Sex should be a categorical/factor variable but is loaded as character. We can fix that with simple base R code to mix things up.
-
-```{r}
-d3$Sex <- as.factor(d3$Sex)  
-skimr::skim(d3)
-```
-
-Now we see that there is another NA, but it's not `NA` from R, instead it was loaded as character and is now considered as a category.
-Well proceed here by removing that individual with that NA entry. Since this keeps an empty category for Sex, I'm also using droplevels() to get rid of it.
-
-```{r}
-d4 <- d3 %>% dplyr::filter(Sex != "NA") %>% droplevels()
-skimr::skim(d4)
+rawdata1%>%filter(grepl("Confirmed", Status))
+unique(rawdata1$Food)
+d2=rawdata1%>%
+  filter(!is.na(Food))%>%
+  mutate(Food=fct_lump(Food,10))%>%
+  count(Food)%>%
+  arrange(desc(n))
+d2
+d3=rawdata1%>%
+  filter(!is.na(Species))%>%
+  mutate(Food=fct_collapse(Food,Norovirus=c("Norovirus","Norovirus genogroup I","Norovirus genogroup II")))%>%
+  mutate(Food=fct_lump(Species,10))%>%
+  count(Species)%>%
+  arrange(desc(n))
+d3
 ```

### 2023-02-23 code/processing_code/processingfile_v1.qmd
part 2 of project

@@
   mutate(Etiology = case_when(
-    Pathogen_Type == "Bacteria" ~
+    Pathogen_Type == "Bacteria" ~ which(Etiology %in% )
 

### 2023-02-23 code/processing_code/processingfile_v1.qmd
working on part 2 of project

--- 2023-02-23 code/processing_code/processingfile_v1.qmd
@@
 library(dplyr)
+library(stringr)
 
@@
             `Food Contaminated Ingredient`)) %>%
-  filter(grepl("Confirmed", `Etiology Status`))
+  filter(grepl("Confirmed", `Etiology Status`)) %>%
+  filter(`Primary Mode` == "Food")
 ```
@@
 ```{r}
-Pathogen_Bacteria <- c("Bacillus", "Brucella", "Campylobacter", "Clostridium", "Staphylococcus", "Escherichia coli", "Listeria", "Salmonella", "Vibrio", "Cholera", "Enterococcus", "Bacterium", "Shigella", "Yersinia")
+Pathogen_Bacteria <- c("Bacillus", "Brucella", "Campylobacter", "Clostridium", "Staphylococcus", "Escherichia", "Listeria", "Salmonella", "Vibrio", "Cholera", "Enterococcus", "Bacterium", "Shigella", "Yersinia")
 Pathogen_Parasite <- c("Cryptosporidium", "Cyclospora", "Entamoeba", "Giardia", "Parasite", "Toxoplasma", "Trichinella")
-Pathogen_Virus <- c("Atrovirus", "Norovirus", "Rotavirus")
-Pathogen_Poison <- c("Poison", "Toxin", "Heavy Metal")
+Pathogen_Virus <- c("Astrovirus", "Norovirus", "Rotavirus")
+Pathogen_Poison <- c("Poison", "Toxin", "Metal")
 
-analysis_data_fin %>%
+analysis_data_fin <- analysis_data_fin %>%
   mutate(Pathogen_Type = case_when(
+    grepl(paste(Pathogen_Poison, 
+                collapse = "|"), Etiology) ~ "Poison",
     grepl(paste(Pathogen_Bacteria, 
@@
     grepl(paste(Pathogen_Virus, 
-                collapse = "|"), Etiology) ~ "Virus",
-    grepl(paste(Pathogen_Poison, 
-                collapse = "|"), Etiology) ~ "Poison")) %>%
-  mutate(Etiology = case_when(
-    Pathogen_Type == "Bacteria" ~
-
-      grepl(paste(Pathogen_Bacteria, collapse = "| "), Etiology)
-  ))
-
+                collapse = "|"), Etiology) ~ "Virus"
+    )) %>%
+  mutate(Simplified_Etiology = case_when(
+    Pathogen_Type == "Bacteria" ~ str_extract(Etiology, 
+      "Bacillus|Brucella|Campylobacter|Clostridium|Staphylococcus|Escherichia|Listeria|Salmonella|Vibrio|Cholera|Enterococcus|Bacterium|Shigella|Yersinia"),
+    Pathogen_Type == "Parasite" ~ str_extract(Etiology, "Cryptosporidium|Cyclospora|Entamoeba|Giardia|Parasite|Toxoplasma|Trichinella"),
+    Pathogen_Type == "Virus" ~ str_extract(Etiology, "Astrovirus|Norovirus|Rotavirus"),
+    Pathogen_Type == "Poison" ~ str_extract(Etiology, "Poison|Toxin|Heavy")
+    )) %>%
+      mutate(Simplified_Etiology = case_when(
+        Simplified_Etiology == "Escherichia" ~ "E. coli",
+        Simplified_Etiology == "Heavy" ~ "Heavy Metals",
+        TRUE ~ Simplified_Etiology))
 ```

### 2023-02-24 products/manuscript/Manuscript.qmd
3
change in manuscript

@@
 ---
-title: "Example Manuscript Template for a Data Analysis Project"
+title: "Manuscript for foodborne-outbreak Project"
 subtitle: ""
@@
 
-The structure below is one possible setup for a data analysis project (including the course project).
-For a manuscript, adjust as needed.
-You don't need to have exactly these sections, but the content covering those sections should be addressed.
-
-This uses MS Word as output format.
-[See here](https://quarto.org/docs/output-formats/ms-word.html) for more information.
-You can switch to other formats, like html or pdf.
-See [the Quarto documentation](https://quarto.org/) for other formats.
-
+set up
 ```{r, echo=FALSE, message=FALSE}
@@
 
-*Provide enough background on your topic that others can understand the why and how of your analysis.*
-
 Foodborne pathogens typically cause serious gastrointestinal infections.
@@
 
-*Describe what the data is, what it contains, where it is from, etc. Eventually this might be part of a methods section.*
-
-This dataset was uploaded onto Kaggle in 2017 and contains data from the Centers for Disease Control and Prevention (CDC) on foodborne disease outbreaks from 1998 to 2015 @centersfordiseasecontrolandprevention2017.
+We adopted two datasets for this study.
+The first dataset was uploaded onto Kaggle in 2017 and contains data from the Centers for Disease Control and Prevention (CDC) on foodborne disease outbreaks from 1998 to 2015 @centersfordiseasecontrolandprevention2017.
 There are 12 variables and 16,950+ observations.
-The raw data may be viewed within the repository of this project.
+The second dataset was collected from National Outbreak Reporting System (NORS), Centers for Disease Control and Prevention (CDC) website.
+This dataset covers the reports of foodborne and waterborne disease outbreaks and enteric (intestinal) disease outbreaks spread by contact with environmental sources, infected people or animals, and other means from 2009 to 2020.
+The raw datasets may be viewed within the repository of this project.
 
@@
 coli*, and for the food most often associated with outbreaks to be some form of salad.
+Moreover, we want to predict the accuracy of the model trained by first dataset using the second dataset as testing data.
 
@@
 To cite other work (important everywhere, but likely happens first in introduction), make sure your references are in the bibtex file specified in the YAML header above (here `dataanalysis_template_references.bib`) and have the right bibtex key.
-Then you can include like this:
-
-Examples of reproducible research projects can for instance be found in [@mckay2020; @mckay2020a]
-
-{{< pagebreak >}}
 
@@
 5.  Mapping cases with different predictors across the states.
+6.  Applying Machine learning approach to test the model accuracy.
 
@@
 
-As previously mentioned, the data was acquired from [Kaggle](https://www.kaggle.com/datasets/cdc/foodborne-diseases).
+As previously mentioned, the data was acquired from [Kaggle](https://www.kaggle.com/datasets/cdc/foodborne-diseases) and [CDC] (https://wwwn.cdc.gov/norsdashboard/).
 
@@
 All materials, including files that contain code, should be commented well so everyone can follow along.*
+```{r}
+#The detailed cleaning process can be found in the file located in ("code","processing","
+```
 

### 2023-02-24 code/processing_code/processingfile_v1.qmd
part 2 stuff

@@
 ---
+# Load Libraries
 
-# Load Libraries
 ```{r}
@@
 ```
+
 ## Checking Etiologies
@@
 
-#potential code for EDA
+# checking the levels under each variable
+
+:D
+
+```{r}
+skim(d1$Food)
+#convert Food to factor
+d1$Food=as.factor(d1$Food)
+str(d1$Food)
+#Since there are over 3000 different levels, we need to combine some food into the same category
+d2=
+```
+
+checking the levels under `Species`
+```{r}
+
+```
+
+# Checking the levels under `Location`
+```{r}
+
+```
 
@@
   separate_rows(Etiology, sep = "; ")
+```
+
+### Modifying IFSAC Category to Include Salads
+```{r}
+Desserts <- c("cookies", "cake", "chocolate", "brownie")
+
+analysis_data_fin <- analysis_data_fin %>%
+  mutate(`IFSAC Category` = case_when(
+    grepl("Salad", `Food Vehicle`, ignore.case = TRUE) ~ "Salad",
+    grepl(paste(Desserts, 
+                collapse = "|"), `Food Vehicle`) ~ "Dessert",
+    grepl("Chicken", `Food Vehicle`, ignore.case = TRUE) ~ "Poultry",
+    grepl("Turkey", `Food Vehicle`, ignore.case = TRUE) ~ "Poultry",
+    grepl("Poultry", `Food Vehicle`, ignore.case = TRUE) ~ "Poultry",
+    TRUE ~ `IFSAC Category`))
 ```
@@
 
+## Check Data Structure Again
+```{r}
+skim(analysis_data_fin)
+```
+
 ```{r}
 rawdata1%>%filter(grepl("Confirmed", Status))
-unique(rawdata1$Food)
-d2=rawdata1%>%
+food=rawdata1%>%
   filter(!is.na(Food))%>%
-  mutate(Food=fct_lump(Food,10))%>%
-  count(Food)%>%
-  arrange(desc(n))
-d2
-d3=rawdata1%>%
+  group_by(Food)%>%
+  top_n(n=10)
+food
+species=rawdata1%>%
   filter(!is.na(Species))%>%
-  mutate(Food=fct_collapse(Food,Norovirus=c("Norovirus","Norovirus genogroup I","Norovirus genogroup II")))%>%
-  mutate(Food=fct_lump(Species,10))%>%
+  mutate(Species=fct_collapse(Species,Norovirus=c("Norovirus","Norovirus genogroup I","Norovirus genogroup II","Norovirus unknown")))%>%
+  mutate(Species=fct_lump(Species,11,other_level = NA))%>%
   count(Species)%>%
-  arrange(desc(n))
-d3
+  arrange(desc(n))%>%
+  filter(!is.na(Species))
+species
 ```

### 2023-02-24 code/analysis_code/exploratory_analysis.qmd
working on cleaning food variable+move EDA part to analysis code

---
title: "An example exploratory analysis script"
author: "Andreas Handel"
date: "10/28/2021"
output: html_document
---

This Quarto file loads the cleaned data and does some exploring.

I'm only showing it the way where the code is included in the file. 
As described in the `processing_code` materials, I currently prefer the approach of having R code in a separate file and pulling it in.

But I already had this written and haven't yet re-done it that way. Feel free to redo and send a pull request on GitHub :)

Again, it is largely a matter of preference and what makes the most sense to decide if one wants to have code inside Quarto files, or as separate R files.
And sometimes, an R script with enough comments is good enough and one doesn't need a Quarto file.

Also note that while here I split cleaning and exploring, this is iterative. You saw that as part of the processing, we already had to explore the data somewhat to understand how to clean it. In general, as you explore, you'll find things that need cleaning. As you clean, you can explore more. Therefore, at times it might make more sense to combine the cleaning and exploring code parts into a single R or Quarto file. Or split things in any other logical way.

As part of the exploratory analysis, you should produce plots or tables or other summary quantities for the most interesting/important quantities in your data. Depending on the total number of variables in your dataset, explore all or some of the others. Figures produced here might be histograms or density plots, correlation plots, etc. Tables might summarize your data.

Start by exploring one variable at a time. Then continue by creating plots or tables of the outcome(s) of interest and the predictor/exposure/input variables you are most interested in. If your dataset is small, you can do that for all variables.

Plots produced here can be scatterplots, boxplots, violinplots, etc. Tables can be simple 2x2 tables or larger ones.

# Setup

```{r}
#load needed packages. make sure they are installed.
library(here) #for data loading/saving
library(tidyverse)

```

Load the data.

```{r}
#Path to data. Note the use of the here() package and not absolute paths
data_location <- here::here("data","processed_data","processeddata.rds")
#load data
mydata <- readRDS(data_location)
```
#outcome: hospitalization/illness
#predictors:year/month/Food/State/Location
#logistic regression models
#visualizaiton: 
linegraphs for years/month and illness, group by species/IFSAC code
boxplot/violin for top 10 food/state/location with hospitalization and illness
tables for IFSAC/ pathogens

# Data exploration through tables

Showing a bit of code to produce and save a summary table.

This uses a bit of code trickery to get things into the shape of a data frame (for easier saving/showing in manuscript)

```{r}
summary_df = data.frame(do.call(cbind, lapply(mydata, summary)))
print(summary_df)
# another way of looking at the data
dplyr::glimpse(mydata)
# save to file
summarytable_file = here("results", "summarytable.rds")
saveRDS(summary_df, file = summarytable_file)
```

We are saving the results to the `results` folder. Depending on how many tables/figures you have, it might make sense to have separate folders for each. And/or you could have separate folders for exploratory tables/figures and for final tables/figures. Just choose a setup that makes sense for your project and works for you, and provide enough documentation that someone can understand what you are doing.

# Data exploration through figures

```{r}
rawdata1%>%filter(grepl("Confirmed", Status))
food=rawdata1%>%
  filter(!is.na(Food))%>%
  group_by(Food)%>%
  top_n(n=10)
food
species=rawdata1%>%
  filter(!is.na(Species))%>%
  mutate(Species=fct_collapse(Species,Norovirus=c("Norovirus","Norovirus genogroup I","Norovirus genogroup II","Norovirus unknown")))%>%
  mutate(Species=fct_lump(Species,11,other_level = NA))%>%
  count(Species)%>%
  arrange(desc(n))%>%
  filter(!is.na(Species))
species
```

```{r}
p1 <- mydata %>% ggplot(aes(x=Height)) + geom_histogram() 
plot(p1)
figure_file = here("results","height_distribution.png")
ggsave(filename = figure_file, plot=p1) 
```

Now weights.

```{r}
p2 <- mydata %>% ggplot(aes(x=Weight)) + geom_histogram() 
plot(p2)
figure_file = here("results","weight_distribution.png")
ggsave(filename = figure_file, plot=p2) 
```

Now height as function of weight.

```{r}
p3 <- mydata %>% ggplot(aes(x=Height, y=Weight)) + geom_point() + geom_smooth(method='lm')
plot(p3)
figure_file = here("results","height_weight.png")
ggsave(filename = figure_file, plot=p3) 
```

Once more height as function of weight, stratified by sex.

```{r}
p4 <- mydata %>% ggplot(aes(x=Height, y=Weight, color = Sex)) + geom_point() + geom_smooth(method='lm')
plot(p4)
figure_file = here("results","height_weight_stratified.png")
ggsave(filename = figure_file, plot=p4) 
```

# Notes

For your own explorations, tables and figures can be "quick and dirty". As long as you can see what's going on, there is no need to polish them. That's in contrast to figures you'll produce for your final products (paper, report, presentation, website, etc.). Those should look as nice, polished and easy to understand as possible.

### 2023-02-24 code/processing_code/processingfile_v1.qmd
working on cleaning food variable+move EDA part to analysis code

@@
 
-## Checking Etiologies
+## Checking Etiologies Status
 ```{r}
@@
 ```
+#Data manipulation
+##checking and cleaning the `Species` column
+```{r}
+unique(training_data_fin$Species)
+#Separate the rows for coninfections
+training_data_fin=training_data_fin%>%
+  separate_rows(Species,sep=";")
+#Create lists for pathogen types
+Pathogen_Bacteria <- c("Bacillus", "Brucella", "Campylobacter", "Clostridium", "Staphylococcus", "Escherichia", "Listeria", "Salmonella", "Vibrio", "Cholera", "Enterococcus", "Bacterium", "Shigella", "Yersinia","Streptococcus")
+Pathogen_Parasite <- c("Cryptosporidium", "Cyclospora", "Entamoeba", "Giardia", "Parasite", "Toxoplasma", "Trichinella")
+Pathogen_Virus <- c("Astrovirus", "Norovirus", "Rotavirus","Hepatitis","Virus")
+Pathogen_Poison <- c("Poison", "Toxin", "Metal","Pesticide","Ciguatoxin","Scombroidtoxin")
+#simplify "Escherichia coli, Shiga toxin-producing" to "Escherichia coli"
+#Create two new columns: Pathogen_type & Simplified_Etiology
+training_data_fin=training_data_fin%>%
+  mutate(Species=str_replace(Species,"Escherichia coli, Shiga toxin-producing","Escherichia"))%>%
+  mutate(Pathogen_Type = case_when(
+    grepl(paste(Pathogen_Poison, 
+                collapse = "|"), Species,ignore.case = TRUE) ~ "Poison",
+    grepl(paste(Pathogen_Bacteria, 
+                collapse = "|"), Species,ignore.case = TRUE) ~ "Bacteria",
+    grepl(paste(Pathogen_Parasite, 
+                collapse = "|"), Species,ignore.case = TRUE) ~ "Parasite",
+    grepl(paste(Pathogen_Virus, 
+                collapse = "|"), Species,ignore.case = TRUE) ~ "Virus"
+    )) %>%
+  mutate(Simplified_Etiology = case_when(
+    Pathogen_Type == "Bacteria" ~ str_extract(Species, 
+      "Bacillus|Brucella|Campylobacter|Clostridium|Staphylococcus|Escherichia|Listeria|Salmonella|Vibrio|Cholera|Enterococcus|Bacterium|Shigella|Yersinia|Streptococcus"),
+    Pathogen_Type == "Parasite" ~ str_extract(Species, "Cryptosporidium|Cyclospora|Entamoeba|Giardia|Parasite|Toxoplasma|Trichinella"),
+    Pathogen_Type == "Virus" ~ str_extract(Species, "Astrovirus|Norovirus|Rotavirus|Hepatitis"),
+    Pathogen_Type == "Poison" ~ str_extract(Species, "Poison|Toxin|Heavy|pesticide|Ciguatoxin|Scombroid")
+    )) %>%
+      mutate(Simplified_Etiology = case_when(
+        Simplified_Etiology == "Escherichia" ~ "E. coli",
+        Simplified_Etiology == "Heavy" ~ "Heavy Metals",
+        TRUE ~ Simplified_Etiology))%>%
+  view()
+```
+# Column for `Location` 
+```{r}
+#checking the names of location and separate the collapsed names
+unique(training_data_fin$Location)
+training_data_fin=training_data_fin%>%
+ separate_rows(Location,sep=";")%>%
+  view()
+
+```
+# Location variable looks clean
+### Add Month as Numeric Column
+```{r}
+training_data_fin <- training_data_fin %>%
+  mutate(Numeric_Month = case_when(
+    Month == "January" ~ 1,
+    Month == "February" ~ 2,
+    Month == "March" ~ 3,
+    Month == "April" ~ 4,
+    Month == "May" ~ 5,
+    Month == "June" ~ 6,
+    Month == "July" ~ 7,
+    Month == "August" ~ 8,
+    Month == "September" ~ 9,
+    Month == "October" ~ 10,
+    Month == "November" ~ 11,
+    Month == "December" ~ 12)) %>%
+  relocate(Numeric_Month, .after = Month)
+```
+##Since there are over 3000+ different categories for food, we want to organize them by adding a IFSAC category column
+```{r}
+unique(analysis_data_fin$`IFSAC Category`)
+unique(training_data_fin$Food)
+Desserts=c("Cookie","Cream","Canoli","Cake","Pudding")
+Beef=c("Beef","Steak")
+Poultry=c("Chicken","Turkey")
+Pork=c("Pork","Sausage","Hot Dog","Deli","Spam","Strawberries","Blueberries","Pig","Ribs")
+Dairy=c("Milk","Cheese","Parsley")
+Crusteaceans=c("Lobster","Crab","Shrimp")
+Fish=c("Fish","Tuna","Salmon")
+Mollusks=c("Clams","Oysters")
+Fruits=c("Tomato","Banana","Melon","Grapes","Mango")
+`Grains-Beans`=c("Rice","Peas","Beans","wheat")
+`Vegetable-Row-Crops`=c("Lettuce","artichokes","broccoli","asparagus","celeries","spinach")
+`Root-underground-vegetables`=c("Roots","carrots","Beets","Onion","Tubers","yams","Garlic","Potatos","taro")
+Sprouts=c("alfalfa","mung")
+Salad=c("Salad")
+Other=c("Ice","Alcohol","Eggnog")
+Game=c("Bear")
+Herbs=c("basil","cilantro")
+Fungi=c("Mushroom","portabellas")
+`Nuts-Seeds`=c("peanuts","almonds","Seasames")
+Eggs=c("Egg")
+`Seeded-Vegetables`=c("Peppers","Squashes","Cucumbers","Okras","Corn")
+`other meat`=c("Lamb")
+
+training_data_fin <- training_data_fin %>%
+  mutate(`IFSAC Category` = case_when(
+    grepl("Salad",Food, ignore.case = TRUE) ~ "Salad",
+    grepl(paste(Desserts, 
+                collapse = "|"), Food,ignore.case = TRUE) ~ "Dessert",
+    grepl(paste(Poultry,
+                collapse = "|"), Food, ignore.case = TRUE) ~ "Poultry",
+    grepl(paste(Beef,
+                collapse = "|"), Food, ignore.case = TRUE) ~ "Beef",
+    grepl(paste(Pork,
+                collapse = "|"), Food, ignore.case = TRUE) ~ "Pork",
+    grepl(paste(Dairy,
+                collapse = "|"), Food, ignore.case = TRUE) ~ "Dairy",
+    grepl(paste(Crusteaceans,
+                collapse = "|"), Food, ignore.case = TRUE) ~ "Crusteaceans",
+    grepl(paste(Fish,
+                collapse = "|"), Food, ignore.case = TRUE) ~ "Fish",
+    grepl(paste(Mollusks,
+                collapse = "|"), Food, ignore.case = TRUE) ~ "Mollusks",
+    grepl(paste(Fruits,
+                collapse = "|"), Food, ignore.case = TRUE) ~ "Fruits",
+    grepl(paste(`Grains-Beans`,
+                collapse = "|"), Food, ignore.case = TRUE) ~ "Grains-Beans",
+    grepl(paste(`Vegetable-Row-Crops`,
+                collapse = "|"), Food, ignore.case = TRUE) ~ "Vegetable-Row-Crops",
+    grepl(paste(`Root-underground-vegetables`,
+                collapse = "|"), Food, ignore.case = TRUE) ~ "Root-underground-vegetables",
+    grepl(paste(Sprouts,
+                collapse = "|"), Food, ignore.case = TRUE) ~ "Sprouts",
+    grepl(paste(Other,
+                collapse = "|"), Food, ignore.case = TRUE) ~ "Other",
+    grepl(paste(Herbs,
+                collapse = "|"), Food, ignore.case = TRUE) ~ "Herbs",
+    grepl(paste(Fungi,
+                collapse = "|"), Food, ignore.case = TRUE) ~ "Fungi",
+    grepl(paste(`Nuts-Seeds`,
+                collapse = "|"), Food, ignore.case = TRUE) ~ "Nuts-Seeds",
+    grepl(paste(`Seeded-Vegetables`,
+                collapse = "|"), Food, ignore.case = TRUE) ~ "Seeded-Vegetables",
+    grepl("eggs",Food,ignore.case = TRUE)~"Eggs"
+
+   ))%>%
+  #mutate(`IFSAC Category`=case_when(Food==str_detect(Food, ";")~"Multiple"))%>%
+  view()
+```
 
@@
   filter(`Primary Mode` == "Food")
-```
-
-# checking the levels under each variable
-
-:D
-
-```{r}
-skim(d1$Food)
-#convert Food to factor
-d1$Food=as.factor(d1$Food)
-str(d1$Food)
-#Since there are over 3000 different levels, we need to combine some food into the same category
-d2=
-```
-
-checking the levels under `Species`
-```{r}
-
-```
-
-# Checking the levels under `Location`
-```{r}
-
 ```
@@
 ```{r}
-Pathogen_Bacteria <- c("Bacillus", "Brucella", "Campylobacter", "Clostridium", "Staphylococcus", "Escherichia", "Listeria", "Salmonella", "Vibrio", "Cholera", "Enterococcus", "Bacterium", "Shigella", "Yersinia")
-Pathogen_Parasite <- c("Cryptosporidium", "Cyclospora", "Entamoeba", "Giardia", "Parasite", "Toxoplasma", "Trichinella")
-Pathogen_Virus <- c("Astrovirus", "Norovirus", "Rotavirus")
-Pathogen_Poison <- c("Poison", "Toxin", "Metal")
 
@@
 ```
-
-### Add Month as Numeric Column
+#we want to drop the NA rows
 ```{r}
-training_data_fin <- training_data_fin %>%
-  mutate(Numeric_Month = case_when(
-    Month == "January" ~ 1,
-    Month == "February" ~ 2,
-    Month == "March" ~ 3,
-    Month == "April" ~ 4,
-    Month == "May" ~ 5,
-    Month == "June" ~ 6,
-    Month == "July" ~ 7,
-    Month == "August" ~ 8,
-    Month == "September" ~ 9,
-    Month == "October" ~ 10,
-    Month == "November" ~ 11,
-    Month == "December" ~ 12)) %>%
-  relocate(Numeric_Month, .after = Month)
+training_data_fin%>%
+  drop_na()
+analysis_data_fin%>%
+  drop_na()
 ```
@@
 ```{r}
+skim(training_data_fin)
 skim(analysis_data_fin)
-```
-
-```{r}
-rawdata1%>%filter(grepl("Confirmed", Status))
-food=rawdata1%>%
-  filter(!is.na(Food))%>%
-  group_by(Food)%>%
-  top_n(n=10)
-food
-species=rawdata1%>%
-  filter(!is.na(Species))%>%
-  mutate(Species=fct_collapse(Species,Norovirus=c("Norovirus","Norovirus genogroup I","Norovirus genogroup II","Norovirus unknown")))%>%
-  mutate(Species=fct_lump(Species,11,other_level = NA))%>%
-  count(Species)%>%
-  arrange(desc(n))%>%
-  filter(!is.na(Species))
-species
 ```
@@
 ```{r}
-processeddata <- d4
+processeddata1<- training_data_fin
+processeddata2<-analysis_data_fin
 ```
@@
 
-Finally, we save the clean data as RDS file.
+Finally, we save the clean data as RDS files.
 
 ```{r}
-save_data_location <- here::here("data","processed_data","processeddata1.rds")
-saveRDS(processeddata1, file = save_data_location)
+save_data_location1 <- here::here("data","processed_data","processeddata1.rds")
+saveRDS(processeddata1, file = save_data_location1)
+
+save_data_location2=here::here("data","processed_data","processeddata2.rds")
+saveRDS(processeddata2, file = save_data_location2)
+
 ```

### 2023-02-24 products/manuscript/Manuscript.qmd
working on cleaning food variable+move EDA part to analysis code

@@
 ```{r}
-#The detailed cleaning process can be found in the file located in ("code","processing","
+#The detailed cleaning process can be found in the file located in ("code","processing","processingfile_v1.qmd")
+
 ```
@@
 ## Statistical analysis
-
-*Explain anything related to your statistical analyses.*
 

### 2023-02-24 code/processing_code/processingfile_v1.qmd
processed code

--- 2023-02-24 code/processing_code/processingfile_v1.qmd
@@
 ```
+#Data manipulation
+##checking and cleaning the `Species` column
+```{r}
+unique(training_data_fin$Species)
+#Separate the rows for coninfections
+training_data_fin=training_data_fin%>%
+  separate_rows(Species,sep=";")
+#Create lists for pathogen types
+Pathogen_Bacteria <- c("Bacillus", "Brucella", "Campylobacter", "Clostridium", "Staphylococcus", "Escherichia", "Listeria", "Salmonella", "Vibrio", "Cholera", "Enterococcus", "Bacterium", "Shigella", "Yersinia","Streptococcus")
+Pathogen_Parasite <- c("Cryptosporidium", "Cyclospora", "Entamoeba", "Giardia", "Parasite", "Toxoplasma", "Trichinella")
+Pathogen_Virus <- c("Astrovirus", "Norovirus", "Rotavirus","Hepatitis","Virus")
+Pathogen_Poison <- c("Poison", "Toxin", "Metal","Pesticide","Ciguatoxin","Scombroidtoxin")
+#simplify "Escherichia coli, Shiga toxin-producing" to "Escherichia coli"
+#Create two new columns: Pathogen_type & Simplified_Etiology
+training_data_fin=training_data_fin%>%
+  mutate(Pathogen_Type = case_when(
+    grepl(paste(Pathogen_Virus, 
+                collapse = "|"), Species,ignore.case = TRUE) ~ "Virus",
+    grepl(paste(Pathogen_Bacteria, 
+                collapse = "|"), Species,ignore.case = TRUE) ~ "Bacteria",
+    grepl(paste(Pathogen_Parasite, 
+                collapse = "|"), Species,ignore.case = TRUE) ~ "Parasite",
+    grepl(paste(Pathogen_Poison, 
+                collapse = "|"), Species,ignore.case = TRUE) ~ "Poison"
+    )) %>%
+  mutate(Simplified_Etiology = case_when(
+    Pathogen_Type == "Bacteria" ~ str_extract(Species, 
+      "Bacillus|Brucella|Campylobacter|Clostridium|Staphylococcus|Escherichia|Listeria|Salmonella|Vibrio|Cholera|Enterococcus|Bacterium|Shigella|Yersinia|Streptococcus"),
+    Pathogen_Type == "Parasite" ~ str_extract(Species, "Cryptosporidium|Cyclospora|Entamoeba|Giardia|Parasite|Toxoplasma|Trichinella"),
+    Pathogen_Type == "Virus" ~ str_extract(Species, "Astrovirus|Norovirus|Rotavirus|Hepatitis"),
+    Pathogen_Type == "Poison" ~ str_extract(Species, "Poison|Toxin|Heavy|pesticide|Ciguatoxin|Scombroid")
+    )) %>%
+      mutate(Simplified_Etiology = case_when(
+        Simplified_Etiology == "Escherichia" ~ "E. coli",
+        Simplified_Etiology == "Heavy" ~ "Heavy Metals",
+        TRUE ~ Simplified_Etiology))%>%
+  view()
+```
+# Column for `Location` 
+```{r}
+#checking the names of location and separate the collapsed names
+unique(training_data_fin$Location)
+training_data_fin=training_data_fin%>%
+ separate_rows(Location,sep=";")%>%
+  view()
+
+```
+# Location variable looks clean
+### Add Month as Numeric Column
+```{r}
+training_data_fin <- training_data_fin %>%
+  mutate(Numeric_Month = case_when(
+    Month == "January" ~ 1,
+    Month == "February" ~ 2,
+    Month == "March" ~ 3,
+    Month == "April" ~ 4,
+    Month == "May" ~ 5,
+    Month == "June" ~ 6,
+    Month == "July" ~ 7,
+    Month == "August" ~ 8,
+    Month == "September" ~ 9,
+    Month == "October" ~ 10,
+    Month == "November" ~ 11,
+    Month == "December" ~ 12)) %>%
+  relocate(Numeric_Month, .after = Month)
+```
+##Since there are over 3000+ different categories for food, we want to organize them by adding a IFSAC category column
+#categorize everything else other than vectors into other
+```{r}
+#unique(analysis_data_fin$`IFSAC Category`)
+#unique(training_data_fin$Food)
+Desserts=c("Cookie","Cream","Canoli","Cake","Pudding")
+Beef=c("Beef","Steak")
+Poultry=c("Chicken","Turkey")
+Pork=c("Pork","Sausage","Hot Dog","Deli","Spam","Strawberries","Blueberries","Pig","Ribs")
+Dairy=c("Milk","Cheese","Parsley")
+Crusteaceans=c("Lobster","Crab","Shrimp")
+Fish=c("Fish","Tuna","Salmon")
+Mollusks=c("Clams","Oysters")
+Fruits=c("Tomato","Banana","Melon","Grapes","Mango")
+`Grains-Beans`=c("Rice","Peas","Beans","wheat")
+`Vegetable-Row-Crops`=c("Lettuce","artichokes","broccoli","asparagus","celeries","spinach")
+`Root-underground-vegetables`=c("Roots","carrots","Beets","Onion","Tubers","yams","Garlic","Potatos","taro")
+Sprouts=c("alfalfa","mung")
+Salad=c("Salad")
+#Other=c("Ice","Alcohol","Eggnog")
+Game=c("Bear")
+Herbs=c("basil","cilantro")
+Fungi=c("Mushroom","portabellas")
+`Nuts-Seeds`=c("peanuts","almonds","Seasames")
+Eggs=c("Egg")
+`Seeded-Vegetables`=c("Peppers","Squashes","Cucumbers","Okras","Corn")
+`other meat`=c("Lamb")
+
+training_data_fin <- training_data_fin %>%
+  mutate(`IFSAC Category` = case_when(
+    grepl("Salad",Food, ignore.case = TRUE) ~ "Salad",
+    grepl(paste(Desserts, 
+                collapse = "|"), Food,ignore.case = TRUE) ~ "Dessert",
+    grepl(paste(Poultry,
+                collapse = "|"), Food, ignore.case = TRUE) ~ "Poultry",
+    grepl(paste(Beef,
+                collapse = "|"), Food, ignore.case = TRUE) ~ "Beef",
+    grepl(paste(Pork,
+                collapse = "|"), Food, ignore.case = TRUE) ~ "Pork",
+    grepl(paste(Dairy,
+                collapse = "|"), Food, ignore.case = TRUE) ~ "Dairy",
+    grepl(paste(Crusteaceans,
+                collapse = "|"), Food, ignore.case = TRUE) ~ "Crusteaceans",
+    grepl(paste(Fish,
+                collapse = "|"), Food, ignore.case = TRUE) ~ "Fish",
+    grepl(paste(Mollusks,
+                collapse = "|"), Food, ignore.case = TRUE) ~ "Mollusks",
+    grepl(paste(Fruits,
+                collapse = "|"), Food, ignore.case = TRUE) ~ "Fruits",
+    grepl(paste(`Grains-Beans`,
+                collapse = "|"), Food, ignore.case = TRUE) ~ "Grains-Beans",
+    grepl(paste(`Vegetable-Row-Crops`,
+                collapse = "|"), Food, ignore.case = TRUE) ~ "Vegetable-Row-Crops",
+    grepl(paste(`Root-underground-vegetables`,
+                collapse = "|"), Food, ignore.case = TRUE) ~ "Root-underground-vegetables",
+    grepl(paste(Sprouts,
+                collapse = "|"), Food, ignore.case = TRUE) ~ "Sprouts",
+
+    grepl(paste(Herbs,
+                collapse = "|"), Food, ignore.case = TRUE) ~ "Herbs",
+    grepl(paste(Fungi,
+                collapse = "|"), Food, ignore.case = TRUE) ~ "Fungi",
+    grepl(paste(`Nuts-Seeds`,
+                collapse = "|"), Food, ignore.case = TRUE) ~ "Nuts-Seeds",
+    grepl(paste(`Seeded-Vegetables`,
+                collapse = "|"), Food, ignore.case = TRUE) ~ "Seeded-Vegetables",
+    grepl("eggs",Food,ignore.case = TRUE)~"Eggs",
+    grepl("Lamb",Food)~"Other Meat",
+    TRUE ~ if_else(is.na(Food),"","Other")
+   ))%>%
+  mutate(`IFSAC Category` = na_if(`IFSAC Category`, ""))%>%
+  relocate(`IFSAC Category`, .after = Food)
+training_data_fin=training_data_fin%>%
+  mutate(`IFSAC Category`=case_when(
+    grepl(";",Food)~"Multiple",TRUE~`IFSAC Category`))
+```
 
@@
 ```
-
-### Add Month as Numeric Column
+#we want to drop the NA rows
 ```{r}
-training_data_fin <- training_data_fin %>%
-  mutate(Numeric_Month = case_when(
-    Month == "January" ~ 1,
-    Month == "February" ~ 2,
-    Month == "March" ~ 3,
-    Month == "April" ~ 4,
-    Month == "May" ~ 5,
-    Month == "June" ~ 6,
-    Month == "July" ~ 7,
-    Month == "August" ~ 8,
-    Month == "September" ~ 9,
-    Month == "October" ~ 10,
-    Month == "November" ~ 11,
-    Month == "December" ~ 12)) %>%
-  relocate(Numeric_Month, .after = Month)
+training_data_final=training_data_fin%>%
+  drop_na()
+analysis_data_final=analysis_data_fin%>%
+  drop_na()
 ```
@@
 ```{r}
-skim(analysis_data_fin)
-```
-
-```{r}
-rawdata1%>%filter(grepl("Confirmed", Status))
-food=rawdata1%>%
-  filter(!is.na(Food))%>%
-  group_by(Food)%>%
-  top_n(n=10)
-food
-species=rawdata1%>%
-  filter(!is.na(Species))%>%
-  mutate(Species=fct_collapse(Species,Norovirus=c("Norovirus","Norovirus genogroup I","Norovirus genogroup II","Norovirus unknown")))%>%
-  mutate(Species=fct_lump(Species,11,other_level = NA))%>%
-  count(Species)%>%
-  arrange(desc(n))%>%
-  filter(!is.na(Species))
-species
+skim(training_data_final)
+skim(analysis_data_final)
 ```
@@
 ```{r}
-processeddata <- d4
+processeddata1<- training_data_final
+processeddata2<-analysis_data_final
 ```

### 2023-02-24 code/analysis_code/exploratory_analysis.qmd
part 2 additional work

@@
 
-# Setup
+# Load Libraries
 
 ```{r}
-#load needed packages. make sure they are installed.
-library(here) #for data loading/saving
+# General Data Handling
 library(tidyverse)
 
+# For Data Loading and Saving
+library(here)
 ```
 
-Load the data.
+# Load Data
+
+## Training Data
+```{r}
+# Path to Data
+data_location <- here(
+  "data","processed_data","processed_training_data.rds")
+# Import Data
+training_data_final <- readRDS(data_location)
+```
+
+## Analysis Data
+```{r}
+# Path to Data
+data_location <- here(
+  "data","processed_data","processed_analysis_data.rds")
+# Import Data
+analysis_data_final <- readRDS(data_location)
+```
+
+# Data Exploration
+## Training Data
+
+The outcome of interest we will explore for the training data is hospitalizations.
 
 ```{r}
-#Path to data. Note the use of the here() package and not absolute paths
-data_location <- here::here("data","processed_data","processeddata.rds")
-#load data
-mydata <- readRDS(data_location)
+# Time vs Hospitalizations
+ggplot(training_data_final %>% group_by(Year, Pathogen_Type, Hospitalizations) %>% count() %>% mutate(true_hospitalization = n * Hospitalizations), aes(x = Year, y = true_hospitalization, group = Pathogen_Type, color = Pathogen_Type)) + geom_point() + geom_jitter() + labs(x = "Year", y = "Hospitalizations", title = "Scatter Plot of Hospitalizations by Year and Pathogen Type")
+
+## Save Graph
+ggsave("time_v_hospitalization_pathogen_type.png", path = "results")
+
+# Pathogen Type + Etiology vs Hospitalization
+ggplot(training_data_final %>% group_by(Pathogen_Type, Simplified_Etiology, Hospitalizations) %>% count() %>% mutate(true_hospitalization = n * Hospitalizations), aes(x = Pathogen_Type, y = true_hospitalization, group = Simplified_Etiology, fill = Simplified_Etiology)) + geom_col() + labs(x = "Pathogen Type", y = "Hospitalizations", title = "Stacked Bar Chart of Hospitalizations by Pathogen Type and Etiology")
+
+## Save Graph
+ggsave("pathogen_type_etiology_v_hospitalizations.png", path = "results")
+
+## Closer Look at Bacterial Etiologies for Outbreaks
+ggplot(training_data_final %>% select(Pathogen_Type, Simplified_Etiology, Hospitalizations) %>% filter(Pathogen_Type == "Bacteria"), aes(x = Simplified_Etiology)) + geom_bar() + labs(x = "Bacteria", y = "Hospitalizations", title = "Bar Chart of Bacteria Frequency")
+
+ggsave("bacterial_etiology_v_hospitalization_frequency.png", path = "results")
+
+# States vs Hospitalization
+
+state_v_hospitalization_training <- training_data_final %>% group_by(State) %>% mutate(all_hospitalizations = sum(Hospitalizations)) %>% count(all_hospitalizations) %>% select(State, all_hospitalizations) %>% arrange(desc(all_hospitalizations))
+
+saveRDS(state_v_hospitalization_training, file = "results/state_v_hospitalization_training.rds")
+
+rmarkdown::paged_table(state_v_hospitalization_training)
+
+# Food Type and Hospitalizations
+ggplot(training_data_final %>% filter(`IFSAC Category` != "Multiple") %>% filter(`IFSAC Category` != "Other"), aes(x = `IFSAC Category`, y = Hospitalizations)) + geom_col() + labs(x = "Food Type", y = "Hospitalizations", title = "Hospitalizations by Food Type, Excluding 'Multiple' and 'Other'")
+
+ggsave("food_type_v_hospitalization.png", path = "results")
+
 ```
-#outcome: hospitalization/illness
-#predictors:year/month/Food/State/Location
-#logistic regression models
-#visualizaiton: 
-linegraphs for years/month and illness, group by species/IFSAC code
-boxplot/violin for top 10 food/state/location with hospitalization and illness
-tables for IFSAC/ pathogens
+All outbreaks with a high hospitalization count (> 200 cases) are caused by bacteria.
+It also appears that most outbreaks caused by bacteria are due to Salmonella.
+California and Minnesota are the two states with the highest total hospitalization count due to foodborne illness from 1998 to 2015.
 
-# Data exploration through tables
+## Analysis Data
+```{r}
+# Illnesses Caused by Pathogen Type Over Time
+ggplot(analysis_data_final %>% group_by(Year, Pathogen_Type, Illnesses) %>% count() %>% mutate(true_illness_count = n * Illnesses), aes(x = Year, y = true_illness_count, group = Pathogen_Type, color = Pathogen_Type)) + geom_point() + geom_jitter() + labs(x = "Year", y = "Cases", title = "Scatter Plot of Illness by Year and Pathogen Type")
 
-Showing a bit of code to produce and save a summary table.
+## Save Graph
+ggsave("time_v_illness_pathogen_type.png", path = "results")
 
-This uses a bit of code trickery to get things into the shape of a data frame (for easier saving/showing in manuscript)
+# Pathogen Type + Etiology vs Illnesses Caused
+ggplot(analysis_data_final %>% group_by(Pathogen_Type, Simplified_Etiology, Illnesses) %>% count() %>% mutate(true_illness_count = n * Illnesses), aes(x = Pathogen_Type, y = Illnesses, group = Simplified_Etiology, fill = Simplified_Etiology)) + geom_col() + labs(x = "Pathogen Type", y = "Cases", title = "Stacked Bar Chart of Illnesses by Pathogen Type and Etiology")
 
-```{r}
-summary_df = data.frame(do.call(cbind, lapply(mydata, summary)))
-print(summary_df)
-# another way of looking at the data
-dplyr::glimpse(mydata)
-# save to file
-summarytable_file = here("results", "summarytable.rds")
-saveRDS(summary_df, file = summarytable_file)
+## Save Graph
+ggsave("pathogen_type_etiology_v_illness.png", path = "results")
+
+## Closer Look at Bacterial Etiologies
+ggplot(analysis_data_final %>% group_by(Pathogen_Type, Simplified_Etiology, Illnesses) %>% filter(Pathogen_Type == "Bacteria"), aes(x = Simplified_Etiology)) + geom_bar() + labs(x = "Bacteria", y = "Cases", title = "Bar Chart of Bacteria Frequency")
+
+ggsave("bacterial_etiology_v_outbreak_frequency.png", path = "results")
+
+# States vs Illnesses
+
+state_v_illness_analysis <- analysis_data_final %>% group_by(State) %>% mutate(all_illnesses = sum(Illnesses)) %>% count(all_illnesses) %>% select(State, all_illnesses) %>% arrange(desc(all_illnesses))
+
+saveRDS(state_v_illness_analysis, file = "results/state_v_illness_analysis.rds")
+
+rmarkdown::paged_table(state_v_illness_analysis)
+
+# Food Type and Hospitalizations
+ggplot(analysis_data_final %>% filter(`IFSAC Category` != "Multiple") %>% filter(`IFSAC Category` != "Other"), aes(x = `IFSAC Category`, y = Illnesses)) + geom_col() + labs(x = "Food Type", y = "Illnesses", title = "Illnesses by Food Type, Excluding 'Multiple' and 'Other'")
+
+ggsave("food_type_v_illness_frequency.png", path = "results")
+
 ```
-
-We are saving the results to the `results` folder.
-Depending on how many tables/figures you have, it might make sense to have separate folders for each.
-And/or you could have separate folders for exploratory tables/figures and for final tables/figures.
-Just choose a setup that makes sense for your project and works for you, and provide enough documentation that someone can understand what you are doing.
-
-# Data exploration through figures
-
-```{r}
-rawdata1%>%filter(grepl("Confirmed", Status))
-food=rawdata1%>%
-  filter(!is.na(Food))%>%
-  group_by(Food)%>%
-  top_n(n=10)
-food
-species=rawdata1%>%
-  filter(!is.na(Species))%>%
-  mutate(Species=fct_collapse(Species,Norovirus=c("Norovirus","Norovirus genogroup I","Norovirus genogroup II","Norovirus unknown")))%>%
-  mutate(Species=fct_lump(Species,11,other_level = NA))%>%
-  count(Species)%>%
-  arrange(desc(n))%>%
-  filter(!is.na(Species))
-species
-```
-
-```{r}
-p1 <- mydata %>% ggplot(aes(x=Height)) + geom_histogram() 
-plot(p1)
-figure_file = here("results","height_distribution.png")
-ggsave(filename = figure_file, plot=p1) 
-```
-
-Now weights.
-
-```{r}
-p2 <- mydata %>% ggplot(aes(x=Weight)) + geom_histogram() 
-plot(p2)
-figure_file = here("results","weight_distribution.png")
-ggsave(filename = figure_file, plot=p2) 
-```
-
-Now height as function of weight.
-
-```{r}
-p3 <- mydata %>% ggplot(aes(x=Height, y=Weight)) + geom_point() + geom_smooth(method='lm')
-plot(p3)
-figure_file = here("results","height_weight.png")
-ggsave(filename = figure_file, plot=p3) 
-```
-
-Once more height as function of weight, stratified by sex.
-
-```{r}
-p4 <- mydata %>% ggplot(aes(x=Height, y=Weight, color = Sex)) + geom_point() + geom_smooth(method='lm')
-plot(p4)
-figure_file = here("results","height_weight_stratified.png")
-ggsave(filename = figure_file, plot=p4) 
-```
-
-# Notes
-
-For your own explorations, tables and figures can be "quick and dirty".
-As long as you can see what's going on, there is no need to polish them.
-That's in contrast to figures you'll produce for your final products (paper, report, presentation, website, etc.).
-Those should look as nice, polished and easy to understand as possible.
+???

### 2023-02-24 code/processing_code/processingfile_v1.qmd
part 2 additional work

--- 2023-02-24 code/processing_code/processingfile_v1.qmd
@@
 ```
+# Data Manipulation
+## Training Data
+### Cleaning the "Species" Column
+```{r}
+unique(training_data_fin$Species)
+# Separate the rows for coninfections
+training_data_fin=training_data_fin%>%
+  separate_rows(Species,sep=";")
+# Create lists for pathogen types
+Pathogen_Bacteria <- c("Bacillus", "Brucella", "Campylobacter", "Clostridium", "Staphylococcus", "Escherichia", "Listeria", "Salmonella", "Vibrio", "Cholera", "Enterococcus", "Bacterium", "Shigella", "Yersinia","Streptococcus")
+Pathogen_Parasite <- c("Cryptosporidium", "Cyclospora", "Entamoeba", "Giardia", "Parasite", "Toxoplasma", "Trichinella")
+Pathogen_Virus <- c("Astrovirus", "Norovirus", "Rotavirus","Hepatitis","Virus")
+Pathogen_Poison <- c("Poison", "Toxin", "Metal","Pesticide","Ciguatoxin","Scombroidtoxin")
+# simplify "Escherichia coli, Shiga toxin-producing" to "Escherichia coli"
+# Create two new columns: Pathogen_type & Simplified_Etiology
+training_data_fin=training_data_fin%>%
+  mutate(Pathogen_Type = case_when(
+    grepl(paste(Pathogen_Virus, 
+                collapse = "|"), Species,ignore.case = TRUE) ~ "Virus",
+    grepl(paste(Pathogen_Bacteria, 
+                collapse = "|"), Species,ignore.case = TRUE) ~ "Bacteria",
+    grepl(paste(Pathogen_Parasite, 
+                collapse = "|"), Species,ignore.case = TRUE) ~ "Parasite",
+    grepl(paste(Pathogen_Poison, 
+                collapse = "|"), Species,ignore.case = TRUE) ~ "Poison"
+    )) %>%
+  mutate(Simplified_Etiology = case_when(
+    Pathogen_Type == "Bacteria" ~ str_extract(Species, 
+      "Bacillus|Brucella|Campylobacter|Clostridium|Staphylococcus|Escherichia|Listeria|Salmonella|Vibrio|Cholera|Enterococcus|Bacterium|Shigella|Yersinia|Streptococcus"),
+    Pathogen_Type == "Parasite" ~ str_extract(Species, "Cryptosporidium|Cyclospora|Entamoeba|Giardia|Parasite|Toxoplasma|Trichinella"),
+    Pathogen_Type == "Virus" ~ str_extract(Species, "Astrovirus|Norovirus|Rotavirus|Hepatitis"),
+    Pathogen_Type == "Poison" ~ str_extract(Species, "Poison|Toxin|Heavy|pesticide|Ciguatoxin|Scombroid")
+    )) %>%
+      mutate(Simplified_Etiology = case_when(
+        Simplified_Etiology == "Escherichia" ~ "E. coli",
+        Simplified_Etiology == "Heavy" ~ "Heavy Metals",
+        TRUE ~ Simplified_Etiology))
+```
+
+### Column for Location
+```{r}
+# Checking Location Names and Separate Multiple Names
+unique(training_data_fin$Location)
+training_data_fin=training_data_fin %>%
+ separate_rows(Location,sep="; ")
+
+```
+
+The variable for location appears to be clean now.
+
+### Add Month as Numeric Column
+```{r}
+training_data_fin <- training_data_fin %>%
+  mutate(Numeric_Month = case_when(
+    Month == "January" ~ 1,
+    Month == "February" ~ 2,
+    Month == "March" ~ 3,
+    Month == "April" ~ 4,
+    Month == "May" ~ 5,
+    Month == "June" ~ 6,
+    Month == "July" ~ 7,
+    Month == "August" ~ 8,
+    Month == "September" ~ 9,
+    Month == "October" ~ 10,
+    Month == "November" ~ 11,
+    Month == "December" ~ 12)) %>%
+  relocate(Numeric_Month, .after = Month)
+```
+
+### Make IFSAC Category Column for Training Data
+
+Everything else not included in vectors will be categorized as "Other".
+
+```{r}
+# unique(analysis_data_fin$`IFSAC Category`)
+# unique(training_data_fin$Food)
+Desserts=c("Cookie","Cream","Canoli","Cake","Pudding")
+Beef=c("Beef","Steak")
+Poultry=c("Chicken","Turkey")
+Pork=c("Pork","Sausage","Hot Dog","Deli","Spam","Strawberries","Blueberries","Pig","Ribs")
+Dairy=c("Milk","Cheese","Parsley")
+Crusteaceans=c("Lobster","Crab","Shrimp")
+Fish=c("Fish","Tuna","Salmon")
+Mollusks=c("Clams","Oysters")
+Fruits=c("Tomato","Banana","Melon","Grapes","Mango")
+`Grains-Beans`=c("Rice","Peas","Beans","wheat")
+`Vegetable-Row-Crops`=c("Lettuce","artichokes","broccoli","asparagus","celeries","spinach")
+`Root-underground-vegetables`=c("Roots","carrots","Beets","Onion","Tubers","yams","Garlic","Potatos","taro")
+Sprouts=c("alfalfa","mung")
+Salad=c("Salad")
+# Other=c("Ice","Alcohol","Eggnog")
+Game=c("Bear")
+Herbs=c("basil","cilantro")
+Fungi=c("Mushroom","portabellas")
+`Nuts-Seeds`=c("peanuts","almonds","Seasames")
+Eggs=c("Egg")
+`Seeded-Vegetables`=c("Peppers","Squashes","Cucumbers","Okras","Corn")
+`other meat`=c("Lamb")
+
+training_data_fin <- training_data_fin %>%
+  mutate(`IFSAC Category` = case_when(
+    grepl("Salad",Food, ignore.case = TRUE) ~ "Salad",
+    grepl(paste(Desserts, 
+                collapse = "|"), Food,ignore.case = TRUE) ~ "Dessert",
+    grepl(paste(Poultry,
+                collapse = "|"), Food, ignore.case = TRUE) ~ "Poultry",
+    grepl(paste(Beef,
+                collapse = "|"), Food, ignore.case = TRUE) ~ "Beef",
+    grepl(paste(Pork,
+                collapse = "|"), Food, ignore.case = TRUE) ~ "Pork",
+    grepl(paste(Dairy,
+                collapse = "|"), Food, ignore.case = TRUE) ~ "Dairy",
+    grepl(paste(Crusteaceans,
+                collapse = "|"), Food, ignore.case = TRUE) ~ "Crusteaceans",
+    grepl(paste(Fish,
+                collapse = "|"), Food, ignore.case = TRUE) ~ "Fish",
+    grepl(paste(Mollusks,
+                collapse = "|"), Food, ignore.case = TRUE) ~ "Mollusks",
+    grepl(paste(Fruits,
+                collapse = "|"), Food, ignore.case = TRUE) ~ "Fruits",
+    grepl(paste(`Grains-Beans`,
+                collapse = "|"), Food, ignore.case = TRUE) ~ "Grains-Beans",
+    grepl(paste(`Vegetable-Row-Crops`,
+                collapse = "|"), Food, ignore.case = TRUE) ~ "Vegetable-Row-Crops",
+    grepl(paste(`Root-underground-vegetables`,
+                collapse = "|"), Food, ignore.case = TRUE) ~ "Root-underground-vegetables",
+    grepl(paste(Sprouts,
+                collapse = "|"), Food, ignore.case = TRUE) ~ "Sprouts",
+
+    grepl(paste(Herbs,
+                collapse = "|"), Food, ignore.case = TRUE) ~ "Herbs",
+    grepl(paste(Fungi,
+                collapse = "|"), Food, ignore.case = TRUE) ~ "Fungi",
+    grepl(paste(`Nuts-Seeds`,
+                collapse = "|"), Food, ignore.case = TRUE) ~ "Nuts-Seeds",
+    grepl(paste(`Seeded-Vegetables`,
+                collapse = "|"), Food, ignore.case = TRUE) ~ "Seeded-Vegetables",
+    grepl("eggs",Food,ignore.case = TRUE)~"Eggs",
+    grepl("Lamb",Food)~"Other Meat",
+    TRUE ~ if_else(is.na(Food),"","Other")
+   ))%>%
+  mutate(`IFSAC Category` = na_if(`IFSAC Category`, ""))%>%
+  relocate(`IFSAC Category`, .after = Food)
+training_data_fin=training_data_fin%>%
+  mutate(`IFSAC Category`=case_when(
+    grepl(";",Food)~"Multiple",TRUE~`IFSAC Category`))
+```
 
@@
 
-# checking the levels under each variable
-
-:D
-
-```{r}
-skim(d1$Food)
-#convert Food to factor
-d1$Food=as.factor(d1$Food)
-str(d1$Food)
-#Since there are over 3000 different levels, we need to combine some food into the same category
-d2=
-```
-
-checking the levels under `Species`
-```{r}
-
-```
-
-# Checking the levels under `Location`
-```{r}
-
-```
-
-## Data Manipulation
 ### Separate Rows to Address Coinfections
@@
 
-### Add Month as Numeric Column
+# Drop Rows Containing NA
 ```{r}
-training_data_fin <- training_data_fin %>%
-  mutate(Numeric_Month = case_when(
-    Month == "January" ~ 1,
-    Month == "February" ~ 2,
-    Month == "March" ~ 3,
-    Month == "April" ~ 4,
-    Month == "May" ~ 5,
-    Month == "June" ~ 6,
-    Month == "July" ~ 7,
-    Month == "August" ~ 8,
-    Month == "September" ~ 9,
-    Month == "October" ~ 10,
-    Month == "November" ~ 11,
-    Month == "December" ~ 12)) %>%
-  relocate(Numeric_Month, .after = Month)
+training_data_final=training_data_fin%>%
+  drop_na()
+analysis_data_final=analysis_data_fin%>%
+  drop_na()
+
 ```
@@
 ```{r}
-skim(analysis_data_fin)
+skim(training_data_final)
+skim(analysis_data_final)
 ```
 
-```{r}
-rawdata1%>%filter(grepl("Confirmed", Status))
-food=rawdata1%>%
-  filter(!is.na(Food))%>%
-  group_by(Food)%>%
-  top_n(n=10)
-food
-species=rawdata1%>%
-  filter(!is.na(Species))%>%
-  mutate(Species=fct_collapse(Species,Norovirus=c("Norovirus","Norovirus genogroup I","Norovirus genogroup II","Norovirus unknown")))%>%
-  mutate(Species=fct_lump(Species,11,other_level = NA))%>%
-  count(Species)%>%
-  arrange(desc(n))%>%
-  filter(!is.na(Species))
-species
-```
+# Save Data
 
-All done, data is clean now.
-
-Let's assign at the end to some final variable, this makes it easier to add further cleaning steps above.
+The cleaned datasets will be saved as RDS files.
 
 ```{r}
-processeddata <- d4
+# Label Final Datasets as Cleaned/Processed
+processed_training_data <- training_data_final
+processed_analysis_data <- analysis_data_final
+
+# Save Data
+save_data_location1 <- here::here("data","processed_data","processed_training_data.rds")
+saveRDS(processed_training_data, file = save_data_location1)
+
+save_data_location2=here::here("data","processed_data","processed_analysis_data.rds")
+saveRDS(processed_analysis_data, file = save_data_location2)
+
 ```
-
-# Save data
-
-Finally, we save the clean data as RDS file.
-
-```{r}
-save_data_location <- here::here("data","processed_data","processeddata1.rds")
-saveRDS(processeddata1, file = save_data_location)
-```
-
-# Notes
-
-Removing anyone who had "faulty" or missing data is one approach.
-It's often not the best.
-based on your question and your analysis approach, you might want to do cleaning differently (e.g.
-keep individuals with some missing information).

### 2023-02-24 products/manuscript/Manuscript.qmd

@@
 
-*Write a summary of your project.*
-
 {{< pagebreak >}}
@@
 
-## Description of data and data source
+## Data Description and Sources
 
@@
 There are 12 variables and 16,950+ observations.
+
 The second dataset was collected from National Outbreak Reporting System (NORS), Centers for Disease Control and Prevention (CDC) website.
 This dataset covers the reports of foodborne and waterborne disease outbreaks and enteric (intestinal) disease outbreaks spread by contact with environmental sources, infected people or animals, and other means from 2009 to 2020.
-The raw datasets may be viewed within the repository of this project.
+The raw datasets may be viewed within the repository of this project (data/raw_data/).
 
-## Questions/Hypotheses to be addressed
+## Questions and Hypotheses to be Addressed
 
-*State the research questions you plan to answer with this analysis.*
-
-We have several questions we would like this data to answer: (1) do foodborne disease outbreaks occur more often in restaurants or in private residences, (2) what pathogen species are most frequently associated with foodborne outbreaks, and (3) what type of food is most commonly involved in foodborne outbreaks.
+We have several questions we would like our datasets to answer: (1) do foodborne disease outbreaks occur more often in restaurants or in private residences, (2) what pathogen species are most frequently associated with foodborne outbreaks, and (3) what type of food is most commonly involved in foodborne outbreaks.
 The outcomes of interest would therefore be location (for whether more outbreaks occur in restaurants/homes), species of pathogen, and food type most often associated with outbreaks.
@@
 
-### Reminder
-
-To cite other work (important everywhere, but likely happens first in introduction), make sure your references are in the bibtex file specified in the YAML header above (here `dataanalysis_template_references.bib`) and have the right bibtex key.
-
 # Methods
@@
 *Describe your methods. That should describe the data, the cleaning processes, and the analysis approaches. You might want to provide a shorter description here and all the details in the supplement.*
+
+Using the datasets previously identified above, we analyzed the structure of each dataset and how many NA values were present for each column in the datasets.
+Afterwards, we selected the variables containing less than 80% NA values for each dataset, which included our predictors and outcomes of interest.
+We then filtered the dataset designated for analysis to include only foodborne disease outbreaks.
+For both datasets, we separated rows to address coinfections individually, and either created or modified the IFSAC Category variable to categorize salads, desserts, and poultry.
+We also created a variable for pathogen type (virus, bacteria, parasite, or toxin) and species classification.
+
+Post-preprocessing, we performed an exploratory data analysis to view general relationships between the predictors: time (year), pathogen type and etiology, state, food type, and health outcomes of interest: hospitalizations and illnesses.
+
+We will be building regression models using these predictors and health outcomes as well; we
 
@@
 
-## Data acquisition
+## Data Acquisition
 
@@
 
+For more information, in-depth preprocessing may be viewed in "code/processing_code/processingfile_v1.qmd".
+
 *Write code that reads in the file and cleans it so it's ready for analysis.
@@
 All materials, including files that contain code, should be commented well so everyone can follow along.*
+
 ```{r}

### 2023-02-24 products/manuscript/Manuscript.qmd
part 2 update

@@
 
-*Describe your methods. That should describe the data, the cleaning processes, and the analysis approaches. You might want to provide a shorter description here and all the details in the supplement.*
+## Data Acquisition
 
-Using the datasets previously identified above, we analyzed the structure of each dataset and how many NA values were present for each column in the datasets.
+As previously mentioned, the data was acquired from [Kaggle](https://www.kaggle.com/datasets/cdc/foodborne-diseases) and [CDC] (https://wwwn.cdc.gov/norsdashboard/).
+
+## Data Importation and Cleaning
+
+Data was imported into RStudio using the tidyverse and readxl packages.
+Using the datasets identified above, we analyzed the structure of each dataset and how many NA values were present for each column in the datasets.
 Afterwards, we selected the variables containing less than 80% NA values for each dataset, which included our predictors and outcomes of interest.
@@
 We also created a variable for pathogen type (virus, bacteria, parasite, or toxin) and species classification.
-
-Post-preprocessing, we performed an exploratory data analysis to view general relationships between the predictors: time (year), pathogen type and etiology, state, food type, and health outcomes of interest: hospitalizations and illnesses.
-
-We will be building regression models using these predictors and health outcomes as well; we
-
-1.  Data wrangling: handling the missing values and cleaning inconsistent data.
-2.  Selecting the interested predictors and outcomes to plot their relationships.
-3.  If the graphs indicate any significant or interesting trend, we will conduct a general linear model and omnibus test.
-4.  Post-hoc test to identify the significant difference between levels.
-5.  Mapping cases with different predictors across the states.
-6.  Applying Machine learning approach to test the model accuracy.
-
-## Data Acquisition
-
-*As applicable, explain where and how you got the data. If you directly import the data from an online source, you can combine this section with the next.*
-
-As previously mentioned, the data was acquired from [Kaggle](https://www.kaggle.com/datasets/cdc/foodborne-diseases) and [CDC] (https://wwwn.cdc.gov/norsdashboard/).
-
-## Data import and cleaning
+We also renamed setting/location (private residence, restaurant, etc.) of outbreaks for consistency.
 
@@
 
-*Write code that reads in the file and cleans it so it's ready for analysis.
-Since this will be fairly long code for most datasets, it might be a good idea to have it in one or several R scripts.
-If that is the case, explain here briefly what kind of cleaning/processing you do, and provide more details and well documented code somewhere (e.g.
-as supplement in a paper).
-All materials, including files that contain code, should be commented well so everyone can follow along.*
+## Statistical Analysis
+Post-preprocessing, we performed an exploratory data analysis to view general relationships between the predictors: time (year), pathogen type and etiology, state, food type, and health outcomes of interest: hospitalizations and illnesses.
 
-```{r}
-#The detailed cleaning process can be found in the file located in ("code","processing","processingfile_v1.qmd")
-
-```
-
-## Statistical analysis
+We will be building regression models using these predictors and health outcomes as well; we intend to implement machine learning to train a model to predict health outcomes using the Kaggle dataset for the CDC NORS data.
 
@@
 
-## Exploratory/Descriptive analysis
+## Exploratory/Descriptive Analysis
 

### 2023-02-24 code/analysis_code/exploratory_analysis.qmd
part 2 work

@@
 ## Save Graph
-ggsave("time_v_hospitalization_pathogen_type.png", path = "results")
+image_path <- here("results")
+ggsave("time_v_hospitalization_pathogen_type.png", path = image_path)
 
@@
 ## Save Graph
-ggsave("pathogen_type_etiology_v_hospitalizations.png", path = "results")
+ggsave("pathogen_type_etiology_v_hospitalizations.png", path = image_path)
 
@@
 
-ggsave("bacterial_etiology_v_hospitalization_frequency.png", path = "results")
+ggsave("bacterial_etiology_v_hospitalization_frequency.png", path = image_path)
 
@@
 
-saveRDS(state_v_hospitalization_training, file = "results/state_v_hospitalization_training.rds")
+## Save Table
+table_path <- here("results", "state_v_hospitalization_training.rds")
+saveRDS(state_v_hospitalization_training, file = table_path)
 
@@
 
-ggsave("food_type_v_hospitalization.png", path = "results")
+ggsave("food_type_v_hospitalization.png", path = image_path)
 
@@
 ## Save Graph
-ggsave("time_v_illness_pathogen_type.png", path = "results")
+ggsave("time_v_illness_pathogen_type.png", path = image_path)
 
@@
 ## Save Graph
-ggsave("pathogen_type_etiology_v_illness.png", path = "results")
+ggsave("pathogen_type_etiology_v_illness.png", path = image_path)
 
@@
 
-ggsave("bacterial_etiology_v_outbreak_frequency.png", path = "results")
+ggsave("bacterial_etiology_v_outbreak_frequency.png", path = image_path)
 
@@
 
-saveRDS(state_v_illness_analysis, file = "results/state_v_illness_analysis.rds")
+saveRDS(state_v_illness_analysis, file = table_path)
 
@@
 
-ggsave("food_type_v_illness_frequency.png", path = "results")
+ggsave("food_type_v_illness_frequency.png", path = image_path)
 

### 2023-02-24 products/manuscript/Manuscript.qmd
part 2 work

@@
 
-set up
 ```{r, echo=FALSE, message=FALSE}
@@
 
-As previously mentioned, the data was acquired from [Kaggle](https://www.kaggle.com/datasets/cdc/foodborne-diseases) and [CDC] (https://wwwn.cdc.gov/norsdashboard/).
+As previously mentioned, the data was acquired from [Kaggle](https://www.kaggle.com/datasets/cdc/foodborne-diseases) and the [CDC](https://wwwn.cdc.gov/norsdashboard/) [@cdc2017; @cdc2020a].
 
@@
 ## Statistical Analysis
+
 Post-preprocessing, we performed an exploratory data analysis to view general relationships between the predictors: time (year), pathogen type and etiology, state, food type, and health outcomes of interest: hospitalizations and illnesses.

### 2023-02-24 code/analysis_code/exploratory_analysis.qmd
exploratory_analysis+processing code final1

--- 2023-02-24 code/analysis_code/exploratory_analysis.qmd
@@
 
-This Quarto file loads the cleaned data and does some exploring.
-
-I'm only showing it the way where the code is included in the file. 
-As described in the `processing_code` materials, I currently prefer the approach of having R code in a separate file and pulling it in.
-
-But I already had this written and haven't yet re-done it that way. Feel free to redo and send a pull request on GitHub :)
-
-Again, it is largely a matter of preference and what makes the most sense to decide if one wants to have code inside Quarto files, or as separate R files.
-And sometimes, an R script with enough comments is good enough and one doesn't need a Quarto file.
-
-Also note that while here I split cleaning and exploring, this is iterative.
-You saw that as part of the processing, we already had to explore the data somewhat to understand how to clean it.
-In general, as you explore, you'll find things that need cleaning.
-As you clean, you can explore more.
-Therefore, at times it might make more sense to combine the cleaning and exploring code parts into a single R or Quarto file.
-Or split things in any other logical way.
-
-As part of the exploratory analysis, you should produce plots or tables or other summary quantities for the most interesting/important quantities in your data.
-Depending on the total number of variables in your dataset, explore all or some of the others.
-Figures produced here might be histograms or density plots, correlation plots, etc.
-Tables might summarize your data.
-
-Start by exploring one variable at a time.
-Then continue by creating plots or tables of the outcome(s) of interest and the predictor/exposure/input variables you are most interested in.
-If your dataset is small, you can do that for all variables.
-
-Plots produced here can be scatterplots, boxplots, violinplots, etc. Tables can be simple 2x2 tables or larger ones.
-
 # Load Libraries
-
 ```{r}
@@
 # Time vs Hospitalizations
-ggplot(training_data_final %>% group_by(Year, Pathogen_Type, Hospitalizations) %>% count() %>% mutate(true_hospitalization = n * Hospitalizations), aes(x = Year, y = true_hospitalization, group = Pathogen_Type, color = Pathogen_Type)) + geom_point() + geom_jitter() + labs(x = "Year", y = "Hospitalizations", title = "Scatter Plot of Hospitalizations by Year and Pathogen Type")
+ggplot(training_data_final %>% 
+         group_by(Year, Pathogen_Type, Hospitalizations) %>% 
+         count() %>% 
+         mutate(true_hospitalization = n * Hospitalizations), aes(x = Year, y = true_hospitalization, group = Pathogen_Type, color = Pathogen_Type)) + geom_point() + 
+  geom_jitter() + 
+  labs(x = "Year", y = "Hospitalizations", title = "Scatter Plot of Hospitalizations by Year and Pathogen Type")
 
@@
 # Pathogen Type + Etiology vs Hospitalization
-ggplot(training_data_final %>% group_by(Pathogen_Type, Simplified_Etiology, Hospitalizations) %>% count() %>% mutate(true_hospitalization = n * Hospitalizations), aes(x = Pathogen_Type, y = true_hospitalization, group = Simplified_Etiology, fill = Simplified_Etiology)) + geom_col() + labs(x = "Pathogen Type", y = "Hospitalizations", title = "Stacked Bar Chart of Hospitalizations by Pathogen Type and Etiology")
+ggplot(training_data_final %>% 
+         group_by(Pathogen_Type, Simplified_Etiology, Hospitalizations) %>% count() %>% 
+         mutate(true_hospitalization = n * Hospitalizations), aes(x = Pathogen_Type, y = true_hospitalization, group = Simplified_Etiology, fill = Simplified_Etiology)) +
+  geom_col() + 
+  labs(x = "Pathogen Type", y = "Hospitalizations", title = "Stacked Bar Chart of Hospitalizations by Pathogen Type and Etiology")
 
@@
 ## Closer Look at Bacterial Etiologies for Outbreaks
-ggplot(training_data_final %>% select(Pathogen_Type, Simplified_Etiology, Hospitalizations) %>% filter(Pathogen_Type == "Bacteria"), aes(x = Simplified_Etiology)) + geom_bar() + labs(x = "Bacteria", y = "Hospitalizations", title = "Bar Chart of Bacteria Frequency")
+ggplot(training_data_final %>% 
+         select(Pathogen_Type, Simplified_Etiology, Hospitalizations) %>% filter(Pathogen_Type == "Bacteria"), aes(x = Simplified_Etiology)) + geom_bar() + 
+  labs(x = "Bacteria", y = "Hospitalizations", title = "Bar Chart of Bacteria Frequency")
 
@@
 
-state_v_hospitalization_training <- training_data_final %>% group_by(State) %>% mutate(all_hospitalizations = sum(Hospitalizations)) %>% count(all_hospitalizations) %>% select(State, all_hospitalizations) %>% arrange(desc(all_hospitalizations))
+state_v_hospitalization_training <- training_data_final %>% 
+  group_by(State) %>% 
+  mutate(all_hospitalizations = sum(Hospitalizations)) %>%     count(all_hospitalizations) %>% 
+  select(State, all_hospitalizations) %>% arrange(desc(all_hospitalizations))
 
@@
 # Food Type and Hospitalizations
-ggplot(training_data_final %>% filter(`IFSAC Category` != "Multiple") %>% filter(`IFSAC Category` != "Other"), aes(x = `IFSAC Category`, y = Hospitalizations)) + geom_col() + labs(x = "Food Type", y = "Hospitalizations", title = "Hospitalizations by Food Type, Excluding 'Multiple' and 'Other'")
+ggplot(training_data_final %>% 
+         filter(`IFSAC Category` != "Multiple") %>% 
+         filter(`IFSAC Category` != "Other"), 
+       aes(x = `IFSAC Category`, y = Hospitalizations)) + 
+  geom_col() + 
+  labs(x = "Food Type", y = "Hospitalizations", title = "Hospitalizations by Food Type, Excluding 'Multiple' and 'Other'")
+ggsave("Food_type_v_hospitalization.png", path = "results")
 
-ggsave("food_type_v_hospitalization.png", path = "results")
+# Locations and Hospitalizations
+Location_v_hospitalization=ggplot(training_data_final, aes(x = Location, y = Hospitalizations)) +
+  geom_col() + 
+  labs(x = "Locations", y = "Hospitalizations", title = "Hospitalizations by Food Type, Excluding 'Multiple' and 'Other'")+
+  scale_x_discrete(guide=guide_axis(n.dodge=3))
+Location_v_hospitalization
+
+ggsave("Location_v_hospitalization.png", path = "results")
 
@@
 California and Minnesota are the two states with the highest total hospitalization count due to foodborne illness from 1998 to 2015.
+Resturants and Private home are the most frequent locations with the highest total hospitalization count.
 
@@
 # Illnesses Caused by Pathogen Type Over Time
-ggplot(analysis_data_final %>% group_by(Year, Pathogen_Type, Illnesses) %>% count() %>% mutate(true_illness_count = n * Illnesses), aes(x = Year, y = true_illness_count, group = Pathogen_Type, color = Pathogen_Type)) + geom_point() + geom_jitter() + labs(x = "Year", y = "Cases", title = "Scatter Plot of Illness by Year and Pathogen Type")
+ggplot(analysis_data_final %>% 
+         group_by(Year, Pathogen_Type, Illnesses) %>% 
+         count() %>% 
+         mutate(true_illness_count = n * Illnesses), 
+       aes(x = Year, y = true_illness_count, group = Pathogen_Type, color = Pathogen_Type)) + 
+  geom_point() + 
+  geom_jitter() + 
+  labs(x = "Year", y = "Cases", title = "Scatter Plot of Illness by Year and Pathogen Type")
 
@@
 # Pathogen Type + Etiology vs Illnesses Caused
-ggplot(analysis_data_final %>% group_by(Pathogen_Type, Simplified_Etiology, Illnesses) %>% count() %>% mutate(true_illness_count = n * Illnesses), aes(x = Pathogen_Type, y = Illnesses, group = Simplified_Etiology, fill = Simplified_Etiology)) + geom_col() + labs(x = "Pathogen Type", y = "Cases", title = "Stacked Bar Chart of Illnesses by Pathogen Type and Etiology")
+ggplot(analysis_data_final %>% 
+         group_by(Pathogen_Type, Simplified_Etiology, Illnesses) %>% 
+         count() %>% 
+         mutate(true_illness_count = n * Illnesses), 
+       aes(x = Pathogen_Type, y = Illnesses, group = Simplified_Etiology, fill = Simplified_Etiology)) + 
+  geom_col() + 
+  labs(x = "Pathogen Type", y = "Cases", title = "Stacked Bar Chart of Illnesses by Pathogen Type and Etiology")
 
@@
 
+#save graph
 ggsave("bacterial_etiology_v_outbreak_frequency.png", path = "results")
@@
 
-# Food Type and Hospitalizations
-ggplot(analysis_data_final %>% filter(`IFSAC Category` != "Multiple") %>% filter(`IFSAC Category` != "Other"), aes(x = `IFSAC Category`, y = Illnesses)) + geom_col() + labs(x = "Food Type", y = "Illnesses", title = "Illnesses by Food Type, Excluding 'Multiple' and 'Other'")
+# Food Type and Illnesses
+ggplot(analysis_data_final %>% filter(`IFSAC Category` != "Multiple") %>% filter(`IFSAC Category` != "Other"), aes(x = `IFSAC Category`, y = Illnesses)) + geom_col() + labs(x = "Food Type", y = "Illnesses", title = "Illnesses by Food Type, Excluding 'Multiple' and 'Other'")+
+  scale_x_discrete(guide = guide_axis(n.dodge=3))
 
+#save graph
 ggsave("food_type_v_illness_frequency.png", path = "results")
 
+# Locations and Illnesses
+ggplot(analysis_data_final, aes(x = Location, y = Illnesses)) +
+  geom_col() + 
+  labs(x = "Locations", y = "Illnesses", title = "Illnesses by Locations")+
+  scale_x_discrete(guide=guide_axis(n.dodge=3))
+
+#save graph
+ggsave("Location_v_Illness.png", path = "results")
+
 ```
-???
+The outbreaks with a high number of cases (> 500) are caused by bacteria.
+The stacked barplot indicates that the top two etiologic agents that contribute to fooborne-illnesses are Salmonella and Norovirus, which belong to the pathogen type of Bacteria and Virus respectively.
+In bacterial pathogen, Salmonella genius is the most dominant factor that is associated with foodborne outbreaks, followed by E.
+coli species and Vibrio genius.
+Geographically speaking, Arkansas and Minnesota are the two states with the highest accumulated foodborne illnesses from 1998 to 2015.
+Salad, Poultry, and Fruits are the three most common types of food vectors that are involved in transmitting foodborne pathogens.
+Resturants and Private home are the most frequent locations that foodborne outbreaks occur.
+Overall, the relationships between predictors and outcomes of training dataset share similar patterns with those of analysis dataset.

### 2023-02-24 code/processing_code/processingfile_v1.qmd
exploratory_analysis+processing code final1

@@
 ```
+###Column for Location
+```{r}
+analysis_data_fin_re <- analysis_data_fin %>%
+  rename(Location=Setting)%>%
+  mutate(Location=case_when(
+    grepl("Restaurant",Location)~"Restaurant",TRUE~Location))%>%
+    separate_rows(Location,sep="; ")
+analysis_data_fin_re
+```
 
@@
   drop_na()
-analysis_data_final=analysis_data_fin%>%
+analysis_data_final=analysis_data_fin_re%>%
   drop_na()

### 2023-02-24 products/manuscript/Manuscript.qmd
...

--- 2023-02-24 products/manuscript/Manuscript.qmd
@@
 
-We have several questions we would like our datasets to answer: (1) do foodborne disease outbreaks occur more often in restaurants or in private residences, (2) what pathogen species are most frequently associated with foodborne outbreaks, and (3) what type of food is most commonly involved in foodborne outbreaks.
+We have several questions we would like our datasets to answer: (1) what are the most frequent locations that foodborne disease outbreaks occur (2) what pathogen species are most frequently associated with foodborne outbreaks, and (3) what type of food is most commonly involved in foodborne outbreaks.
 The outcomes of interest would therefore be location (for whether more outbreaks occur in restaurants/homes), species of pathogen, and food type most often associated with outbreaks.

### 2023-02-24 code/analysis_code/exploratory_analysis.qmd
part 2 continued work

@@
   labs(x = "Food Type", y = "Hospitalizations", title = "Hospitalizations by Food Type, Excluding 'Multiple' and 'Other'")
-ggsave("Food_type_v_hospitalization.png", path = "results")
+ggsave("Food_type_v_hospitalization.png", path = image_path)
 
 # Locations and Hospitalizations
-Location_v_hospitalization=ggplot(training_data_final, aes(x = Location, y = Hospitalizations)) +
+Location_v_Hospitalization <- ggplot(training_data_final, aes(x = Location, y = Hospitalizations)) +
   geom_col() + 
@@
   scale_x_discrete(guide=guide_axis(n.dodge=3))
-Location_v_hospitalization
 
-ggsave("Location_v_hospitalization.png", path = "results")
+Location_v_Hospitalization
+
+ggsave("Location_v_hospitalization.png", path = image_path)
 
@@
 
-#save graph
-ggsave("bacterial_etiology_v_outbreak_frequency.png", path = "results")
+# Save Graph
+ggsave("bacterial_etiology_v_outbreak_frequency.png", path = image_path)
 
@@
 
-#save graph
-ggsave("food_type_v_illness_frequency.png", path = "results")
+# Save Graph
+ggsave("food_type_v_illness_frequency.png", path = image_path)
 
@@
 
-#save graph
-ggsave("Location_v_Illness.png", path = "results")
+# Save Graph
+ggsave("Location_v_Illness.png", path = image_path)
 

### 2023-02-24 products/manuscript/Manuscript.qmd
part 2 continued work

@@
 We adopted two datasets for this study.
-The first dataset was uploaded onto Kaggle in 2017 and contains data from the Centers for Disease Control and Prevention (CDC) on foodborne disease outbreaks from 1998 to 2015 @centersfordiseasecontrolandprevention2017.
+The first dataset [@cdc2020]was uploaded onto Kaggle in 2017 and contains data from the Centers for Disease Control and Prevention (CDC) on foodborne disease outbreaks from 1998 to 2015 [@cdc2017].
 There are 12 variables and 16,950+ observations.
 
-The second dataset was collected from National Outbreak Reporting System (NORS), Centers for Disease Control and Prevention (CDC) website.
+The second dataset was collected from the National Outbreak Reporting System (NORS), Centers for Disease Control and Prevention (CDC) website .
 This dataset covers the reports of foodborne and waterborne disease outbreaks and enteric (intestinal) disease outbreaks spread by contact with environmental sources, infected people or animals, and other means from 2009 to 2020.
@@
 
-We have several questions we would like our datasets to answer: (1) what are the most frequent locations that foodborne disease outbreaks occur (2) what pathogen species are most frequently associated with foodborne outbreaks, and (3) what type of food is most commonly involved in foodborne outbreaks.
+We have several questions we would like our datasets to answer: (1) what are the most frequent locations that foodborne disease outbreaks occur, (2) what pathogen species are most frequently associated with foodborne outbreaks, and (3) what type of food is most commonly involved in foodborne outbreaks.
 The outcomes of interest would therefore be location (for whether more outbreaks occur in restaurants/homes), species of pathogen, and food type most often associated with outbreaks.

### 2023-02-24 code/analysis_code/exploratory_analysis.qmd
work for part 2

@@
 image_path <- here("results")
-ggsave("time_v_hospitalization_pathogen_type.png", path = image_path)
+ggsave("time_v_hospitalization_pathogen_type.png", path = image_path, width = 3000, height = 2000, units = "px")
 
@@
 ## Save Graph
-ggsave("pathogen_type_etiology_v_hospitalizations.png", path = image_path)
+ggsave("pathogen_type_etiology_v_hospitalizations.png", path = image_path, width = 3000, height = 2000, units = "px")
 
@@
 
-ggsave("bacterial_etiology_v_hospitalization_frequency.png", path = image_path)
+ggsave("bacterial_etiology_v_hospitalization_frequency.png", path = image_path, width = 3000, height = 2000, units = "px")
 
@@
   labs(x = "Food Type", y = "Hospitalizations", title = "Hospitalizations by Food Type, Excluding 'Multiple' and 'Other'")
-ggsave("Food_type_v_hospitalization.png", path = image_path)
+ggsave("Food_type_v_hospitalization.png", path = image_path, width = 3000, height = 2000, units = "px")
 
@@
 
-ggsave("Location_v_hospitalization.png", path = image_path)
+ggsave("Location_v_hospitalization.png", path = image_path, width = 3000, height = 2000, units = "px")
 
@@
 ## Save Graph
-ggsave("time_v_illness_pathogen_type.png", path = image_path)
+ggsave("time_v_illness_pathogen_type.png", path = image_path, width = 3000, height = 2000, units = "px")
 
@@
 ## Save Graph
-ggsave("pathogen_type_etiology_v_illness.png", path = image_path)
+ggsave("pathogen_type_etiology_v_illness.png", path = image_path, width = 3000, height = 2000, units = "px")
 
@@
 # Save Graph
-ggsave("bacterial_etiology_v_outbreak_frequency.png", path = image_path)
+ggsave("bacterial_etiology_v_outbreak_frequency.png", path = image_path, width = 3000, height = 2000, units = "px")
 
@@
 # Save Graph
-ggsave("food_type_v_illness_frequency.png", path = image_path)
+ggsave("food_type_v_illness_frequency.png", path = image_path, width = 3000, height = 2000, units = "px")
 
@@
 # Save Graph
-ggsave("Location_v_Illness.png", path = image_path)
+ggsave("Location_v_Illness.png", path = image_path, width = 3000, height = 2000, units = "px")
 

### 2023-02-24 products/manuscript/Manuscript.qmd
work for part 2

@@
 
-*Use a combination of text/tables/figures to explore and describe your data.
-Show the most important descriptive results here.
-Additional ones should go in the supplement.
-Even more can be in the R and Quarto files that are part of your project.*
+### Training Data
 
-@tbl-summarytable shows a summary of the data.
-
-Note the loading of the data providing a **relative** path using the `../../` notation.
-(Two dots means a folder up).
-You never want to specify an **absolute** path like `C:\ahandel\myproject\results\` because if you share this with someone, it won't work for them since they don't have that path.
-You can also use the `here` R package to create paths.
-See examples of that below.
+![Figure 2: Hospitalizations by Food Type.](images/Food_type_v_hospitalization-01.png)
 
 ```{r}
-#| label: tbl-summarytable
-#| tbl-cap: "Data summary table."
-#| echo: FALSE
-resulttable=readRDS("../../results/summarytable.rds")
-knitr::kable(resulttable)
+rmarkdown::paged_table(readRDS("../../results/state_v_hospitalization_training.rds"))
 ```
 
-## Basic statistical analysis
+### Analysis Data
 
-*To get some further insight into your data, if reasonable you could compute simple statistics (e.g.
-simple models with 1 predictor) to look for associations between your outcome(s) and each individual predictor variable.
-Though note that unless you pre-specified the outcome and main exposure, any "p\<0.05 means statistical significance" interpretation is not valid.*
+![Figure 1. Year vs Illnesses Caused by Pathogen Type](images/time_v_illness_pathogen_type.png)
 
-@fig-result shows a scatterplot figure produced by one of the R scripts.
+![Figure 3. Illnesses by outbreak location.](images/Location_v_Illness.png)
 
-```{r}
-#| label: fig-result
-#| fig-cap: "Height and weight stratified by sex."
-#| echo: FALSE
-knitr::include_graphics(here("results","height_weight_stratified.png"))
-```
+![Figure 4. Bacterial etiology by outbreak frequency.](images/bacterial_etiology_v_outbreak_frequency.png)
 
-## Full analysis
+## Basic Statistical Analysis
+
+## Full Analysis
 

### 2023-02-24 products/manuscript/Manuscript.qmd
Update Manuscript.qmd

@@
 
-## Full Analysis
-
-*Use one or several suitable statistical/machine learning methods to analyze your data and to produce meaningful figures, tables, etc.
-This might again be code that is best placed in one or several separate R scripts that need to be well documented.
-You want the code to produce figures and data ready for display as tables, and save those.
-Then you load them here.*
-
-Example @tbl-resulttable2 shows a summary of a linear model fit.
-
-```{r}
-#| label: tbl-resulttable2
-#| tbl-cap: "Linear model fit table."
-#| echo: FALSE
-resulttable2 = readRDS(here("results","resulttable2.rds"))
-knitr::kable(resulttable2)
-```
+## Full Analysis (Machine Learning)
 
@@
 
-*Summarize what you did, what you found and what it means.*
-
 ## Strengths and Limitations
 
-*Discuss what you perceive as strengths and limitations of your analysis.*
-
 ## Conclusions
-
-*What are the main take-home messages?*
-
-*Include citations in your Rmd file using bibtex, the list of references will automatically be placed at the end*
-
-This paper [@leek2015] discusses types of analyses.
-
-These papers [@mckay2020; @mckay2020a] are good examples of papers published using a fully reproducible setup similar to the one shown in this template.
-
-Note that this cited reference will show up at the end of the document, the reference formatting is determined by the CSL file specified in the YAML header.
-Many more style files for almost any journal [are available](https://www.zotero.org/styles).
-You also specify the location of your bibtex reference file in the YAML.
-You can call your reference file anything you like, I just used the generic word `references.bib` but giving it a more descriptive name is probably better.
 

### 2023-02-24 products/manuscript/Manuscript.qmd
work for part 2

@@
 ```{r, echo=FALSE, message=FALSE}
-# load a few R packages
+# Load Libraries
 library(here)
@@
 We adopted two datasets for this study.
-The first dataset [@cdc2020]was uploaded onto Kaggle in 2017 and contains data from the Centers for Disease Control and Prevention (CDC) on foodborne disease outbreaks from 1998 to 2015 [@cdc2017].
+The first dataset was uploaded onto Kaggle in 2017 and contains data from the Centers for Disease Control and Prevention (CDC) on foodborne disease outbreaks from 1998 to 2015 [@cdc2017].
 There are 12 variables and 16,950+ observations.
 
-The second dataset was collected from the National Outbreak Reporting System (NORS), Centers for Disease Control and Prevention (CDC) website .
+The second dataset was collected from the National Outbreak Reporting System (NORS), Centers for Disease Control and Prevention (CDC) website [@cdc2020].
 This dataset covers the reports of foodborne and waterborne disease outbreaks and enteric (intestinal) disease outbreaks spread by contact with environmental sources, infected people or animals, and other means from 2009 to 2020.

### 2023-02-26 code/analysis_code/exploratory_analysis.qmd
add map

@@
 
+## Project total hospitalizations in states map
+training_hosp=training_data_final%>%
+  mutate(state=tolower(State))%>%
+  group_by(state)%>%
+  mutate(all_hosp = sum(Hospitalizations)) %>%  count(all_hosp) %>%
+  select(state, all_hosp)
+training_hosp
+
+state_map=training_hosp%>%
+  distinct(state)%>%
+  inner_join(.,map_data("state"),by=c("state"="region"))
+state_map
+
+map_final=right_join(training_hosp,state_map,by=c("state"="state"))
+
+map_final%>%
+  ggplot(aes(x=long,y=lat,fill=all_hosp,group=group))+
+  geom_polygon(color="black",size=0.1)+
+  geom_polygon(data=state_map,aes(x=long,y=lat,group=group),fill=NA,color="black",size=0.3)+
+  scale_fill_gradient(name="Number of \n Hospitalizations",
+                      low="#FFFFFF",high="#FF0000",
+                       limits=c(0,NA))+
+  labs(title="Cases of Hospitalizations Caused by Foodborne Illnesses\n in Each State During 1998-2015")+
+  theme_void()+
+  coord_quickmap()
+
+ggsave("State_hospitalization_map.png",path=image_path,width=3000,height=2000,units="px")
+
 # Food Type and Hospitalizations
@@
   labs(x = "Food Type", y = "Hospitalizations", title = "Hospitalizations by Food Type, Excluding 'Multiple' and 'Other'")
+
 ggsave("Food_type_v_hospitalization.png", path = image_path, width = 3000, height = 2000, units = "px")
@@
 The stacked barplot indicates that the top two etiologic agents that contribute to fooborne-illnesses are Salmonella and Norovirus, which belong to the pathogen type of Bacteria and Virus respectively.
-In bacterial pathogen, Salmonella genius is the most dominant factor that is associated with foodborne outbreaks, followed by E.
-coli species and Vibrio genius.
+In bacterial pathogen, Salmonella genus is the most dominant factor that is associated with foodborne outbreaks, followed by E.
+coli species and Vibrio genus.
 Geographically speaking, Arkansas and Minnesota are the two states with the highest accumulated foodborne illnesses from 1998 to 2015.

### 2023-02-26 code/analysis_code/statistical_analysis.R

@@ -40,22 +40,7 @@ saveRDS(lmtable1, file = table_file1)
 ############################
 #### Second model fit
 # fit linear model using height as outcome, weight and sex as predictor
-# Mapping the illness data through the states
 
-us=map_data("usa")
-map_States=training_data_final%>%
-  group_by(State)%>%
-  mutate(all_illnesses = sum(Illnesses)) %>%  count(all_illnesses) %>% select(State, all_illnesses)
-
-map1=map_States%>%
-  rename("region"="State")%>%
-  mutate(region=tolower(region))
-map1
-ggplot(data=map1)+
-  geom_map(aes(map_id=region,fill=as.factor(map1$all_illnesses)),map=us) + 
-  expand_limits(x=us$long,y=us$lat)
-map1
-lmfit2 <- lm(Height ~ Weight + Sex, mydata)  
 
 # place results from fit into a data frame with the tidy function
 lmtable2 <- broom::tidy(lmfit2)

### 2023-03-14 code/analysis_code/exploratory_analysis.qmd
update processing and exploratory analysis files

@@
 data_location <- here(
-  "data","processed_data","processed_analysis_data.rds")
+  "data","processed_data","processed_testing_data.rds")
 # Import Data
@@
 ggplot(training_data_final %>% 
-         select(Pathogen_Type, Simplified_Etiology, Hospitalizations) %>% filter(Pathogen_Type == "Bacteria"), aes(x = Simplified_Etiology)) + geom_bar() + 
-  labs(x = "Bacteria", y = "Hospitalizations", title = "Bar Chart of Bacteria Frequency")
+         select(Pathogen_Type, Simplified_Etiology, Hospitalizations, `IFSAC Category`) %>% filter(Pathogen_Type == "Bacteria"), aes(x = Simplified_Etiology,y=`Hospitalizations`))+ 
+  geom_col() + 
+  labs(x = "Bacteria", y = "Hospitalizations", title = "Bar Chart of Bacteria Frequency")+
+  coord_flip()
 
@@
   geom_col() + 
-  labs(x = "Food Type", y = "Hospitalizations", title = "Hospitalizations by Food Type, Excluding 'Multiple' and 'Other'")
+  labs(x = "Food Type", y = "Hospitalizations", title = "Hospitalizations by Food Type, Excluding 'Multiple' and 'Other'")+
+  coord_flip()
 
@@
   labs(x = "Locations", y = "Hospitalizations", title = "Hospitalizations by Food Type, Excluding 'Multiple' and 'Other'")+
-  scale_x_discrete(guide=guide_axis(n.dodge=3))
+  coord_flip()
 
@@
 
-## Analysis Data
+## Testing Data
 ```{r}
 # Illnesses Caused by Pathogen Type Over Time
-ggplot(analysis_data_final %>% 
+ggplot(testing_data_final %>% 
          group_by(Year, Pathogen_Type, Illnesses) %>% 
@@
 # Pathogen Type + Etiology vs Illnesses Caused
-ggplot(analysis_data_final %>% 
+ggplot(testing_data_final %>% 
          group_by(Pathogen_Type, Simplified_Etiology, Illnesses) %>% 
@@
 ## Closer Look at Bacterial Etiologies
-ggplot(analysis_data_final %>% group_by(Pathogen_Type, Simplified_Etiology, Illnesses) %>% filter(Pathogen_Type == "Bacteria"), aes(x = Simplified_Etiology)) + geom_bar() + labs(x = "Bacteria", y = "Cases", title = "Bar Chart of Bacteria Frequency")
+ggplot(testing_data_final %>% group_by(Pathogen_Type, Simplified_Etiology, Illnesses) %>% filter(Pathogen_Type == "Bacteria"), aes(x = Simplified_Etiology)) + geom_bar() + labs(x = "Bacteria", y = "Cases", title = "Bar Chart of Bacteria Frequency")+
+  coord_flip()
 
@@
 
-state_v_illness_analysis <- analysis_data_final %>% group_by(State) %>% mutate(all_illnesses = sum(Illnesses)) %>% count(all_illnesses) %>% select(State, all_illnesses) %>% arrange(desc(all_illnesses))
+state_v_illness_testing <- testing_data_final %>% group_by(State) %>% mutate(all_illnesses = sum(Illnesses)) %>% count(all_illnesses) %>% select(State, all_illnesses) %>% arrange(desc(all_illnesses))
 
-saveRDS(state_v_illness_analysis, file = table_path)
+saveRDS(state_v_illness_testing, file = table_path)
 
@@
 # Food Type and Illnesses
-ggplot(analysis_data_final %>% filter(`IFSAC Category` != "Multiple") %>% filter(`IFSAC Category` != "Other"), aes(x = `IFSAC Category`, y = Illnesses)) + geom_col() + labs(x = "Food Type", y = "Illnesses", title = "Illnesses by Food Type, Excluding 'Multiple' and 'Other'")+
-  scale_x_discrete(guide = guide_axis(n.dodge=3))
+ggplot(testing_data_final %>% filter(`IFSAC Category` != "Multiple") %>% filter(`IFSAC Category` != "Other"), aes(x = `IFSAC Category`, y = Illnesses)) + geom_col() + labs(x = "Food Type", y = "Illnesses", title = "Illnesses by Food Type, Excluding 'Multiple' and 'Other'")+
+  coord_flip()
 
@@
 # Locations and Illnesses
-ggplot(analysis_data_final, aes(x = Location, y = Illnesses)) +
+ggplot(testing_data_final, aes(x = Location, y = Illnesses)) +
   geom_col() + 
   labs(x = "Locations", y = "Illnesses", title = "Illnesses by Locations")+
-  scale_x_discrete(guide=guide_axis(n.dodge=3))
+  coord_flip()
 

### 2023-03-14 code/processing_code/processingfile_v1.qmd
update processing and exploratory analysis files

@@
   "data","raw_data","NationalOutbreakPublicDataTool.xlsx")
-analysis_data <- readxl::read_excel(data_location)
+testing_data <- readxl::read_excel(data_location)
 ```
@@
 
-## Analysis Data
+## Testing Data
 ```{r}
@@
     grepl(paste(Pathogen_Virus,
-                collapse = "|"), Species,ignore.case = TRUE) ~ "Virus",
+                collapse="|"), Species,ignore.case = TRUE) ~ "Virus",
     grepl(paste(Pathogen_Bacteria, 
@@
     )) %>%
-  mutate(Simplified_Etiology = case_when(
-    Pathogen_Type == "Bacteria" ~ str_extract(Species, 
-      "Bacillus|Brucella|Campylobacter|Clostridium|Staphylococcus|Escherichia|Listeria|Salmonella|Vibrio|Cholera|Enterococcus|Bacterium|Shigella|Yersinia|Streptococcus"),
-    Pathogen_Type == "Parasite" ~ str_extract(Species, "Cryptosporidium|Cyclospora|Entamoeba|Giardia|Parasite|Toxoplasma|Trichinella"),
-    Pathogen_Type == "Virus" ~ str_extract(Species, "Astrovirus|Norovirus|Rotavirus|Hepatitis"),
-    Pathogen_Type == "Poison" ~ str_extract(Species, "Poison|Toxin|Heavy|pesticide|Ciguatoxin|Scombroid")
-    )) %>%
+  mutate(Simplified_Etiology = word(Species,1))%>%
       mutate(Simplified_Etiology = case_when(
@@
  separate_rows(Location,sep="; ")
-
 ```
@@
     )) %>%
-  mutate(Simplified_Etiology = case_when(
-    Pathogen_Type == "Bacteria" ~ str_extract(Etiology, 
-      "Bacillus|Brucella|Campylobacter|Clostridium|Staphylococcus|Escherichia|Listeria|Salmonella|Vibrio|Cholera|Enterococcus|Bacterium|Shigella|Yersinia"),
-    Pathogen_Type == "Parasite" ~ str_extract(Etiology, "Cryptosporidium|Cyclospora|Entamoeba|Giardia|Parasite|Toxoplasma|Trichinella"),
-    Pathogen_Type == "Virus" ~ str_extract(Etiology, "Astrovirus|Norovirus|Rotavirus"),
-    Pathogen_Type == "Poison" ~ str_extract(Etiology, "Poison|Toxin|Heavy")
-    )) %>%
+  mutate(Simplified_Etiology = word(Etiology,1)) %>%
       mutate(Simplified_Etiology = case_when(

### 2023-03-14 code/analysis_code/exploratory_analysis.qmd

@@
   geom_col() + 
-  labs(x = "Pathogen Type", y = "Hospitalizations", title = "Stacked Bar Chart of Hospitalizations by Pathogen Type and Etiology")
+  labs(x = "Pathogen Type", y = "Hospitalizations", title = "Stacked Bar Chart of Hospitalizations by \nPathogen Type and Etiology")
 
@@
   geom_col() + 
-  labs(x = "Locations", y = "Hospitalizations", title = "Hospitalizations by Food Type, Excluding 'Multiple' and 'Other'")+
+  labs(x = "Locations", y = "Hospitalizations", title = "Hospitalizations by Food Type, Excluding 'Multiple' and \n'Other'")+
   coord_flip()

### 2023-03-16 code/analysis_code/exploratory_analysis.qmd
statistical_analysis_code

--- 2023-02-26 code/analysis_code/exploratory_analysis.qmd
@@
 
-```{r}
+```{r fig.height=8,fig.width=8}
 # Time vs Hospitalizations
@@
 ggsave("bacterial_etiology_v_hospitalization_frequency.png", path = image_path, width = 3000, height = 2000, units = "px")
+
+####
+
+Norovirus_Hosp=training_data_final%>%
+  select(Simplified_Etiology, Hospitalizations, `IFSAC Category`)%>%
+  group_by(Simplified_Etiology, Hospitalizations, `IFSAC Category`)%>%
+  filter(Simplified_Etiology=="Salmonella")%>%
+  filter(`IFSAC Category`!= %n% C("Multiple","Other"))%>%
+  count()%>%
+  mutate(total_hospitalizations=n*Hospitalizations)
+
+ggplot(Norovirus_Hosp,aes(x=`IFSAC Category`,y=total_hospitalizations))+
+  geom_boxplot()+
+  scale_x_discrete(guide=guide_axis(n.dodge=3))
 

### 2023-03-16 code/analysis_code/statistical_analysis.qmd
statistical_analysis_code

# Load libraries
```{r}
library(here)
library(tidyverse)
library(tidymodels)
library(poissonreg)
```
# Load data
```{r}
# Path to Data
data_location <- here(
  "data","processed_data","processed_training_data.rds")
# Import Data
training_data_final <- readRDS(data_location)
glimpse(training_data_final)
```
## First health outcome of interest: Illnesses
### Predictors: time (year), pathogen type, etiology, location, state, and food type (`IFSAC Category`)
```{r}
# Checking the distribution of Illnesses
training_data_final%>%
  ggplot(aes(Illnesses))+
  geom_histogram()+
  scale_x_continuous(trans=scales::pseudo_log_trans(base=10))
image_path <- here("results")
ggsave("histogram_Illnesses.png", path = image_path, width = 3000, height = 2000, units = "px")
# Creating a tibble only includes our predictors of interest
illness_data=training_data_final%>%
  select(c(1,4,5,7,10,13,14))%>%
  mutate_if(is.character, as.factor)
glimpse(illness_data)
```
### Modeling `Illness` using linear regression and Poisson regression
#### Since the outcome is counts and the event ocurrs independently, it follows a Poisson distribution
```{r}
# Model specification and fitting for linear regression model
lm_spec=linear_reg()%>%
  set_engine(engine="lm")
# Modeling with main predictor `Simplified_Etiology`
lm_fit1=lm_spec%>%
  fit(Illnesses~Simplified_Etiology, data=illness_data)
# Modeling with all predictors
lm_fit2=lm_spec%>%
  fit(Illnesses~., data=illness_data)
# Model specification and fitting for Poisson regression model
# create a workflow for poisson regression model, we will use the default glm engine
# Modeling with main predictor `Simplified_Etiology`
poisson_wf1=workflow(Illnesses~Simplified_Etiology, poisson_reg())
# Modeling with all predictors
poisson_wf2=workflow(Illnesses~., poisson_reg())
# Fitting models
poisson_fit1=fit(poisson_wf1,data=illness_data)
poisson_fit2=fit(poisson_wf2,data=illness_data)
```

### Model Evaluation
```{r}
# Checking parameters for linear regression model
tidy(lm_fit1) %>%
  arrange(-estimate)
tidy(lm_fit2)%>%
  arrange(-estimate)
# Checking parameters for Poisson regression model
tidy(poisson_fit1)%>%
 arrange(-estimate)
tidy(poisson_fit2)%>%
 arrange(-estimate)
# Model evaluation
Result=lm_fit1%>%
 predict(new_data=illness_data)%>%
 mutate (.pred=round(.pred),true_value=illness_data$Illnesses, model= "lm_single")%>%
bind_rows(lm_fit1%>%
            predict(new_data=illness_data)%>%
            mutate (.pred=round(.pred),true_value=illness_data$Illnesses, model= "lm_multiple")%>%
            bind_rows(poisson_fit1%>%
                        predict(new_data= illness_data)%>% 
                        mutate (true_value=illness_data$Illnesses, model= "glm_single")%>%
                        bind_rows(poisson_fit2%>%
                                    predict(new_data= illness_data)%>% 
                                    mutate (true_value=illness_data$Illnesses, model= "glm_multiple")
                        )
            )
)
Result%>%
  group_by(model)%>%
  rmse(truth=true_value,estimate=.pred)

```
### According to the result, the model built using Poisson regression with multiple predictors seems to have the smallest root mean squared error.
## Second health outcome of interest: Hospitilization
### Predictors: time (year), pathogen type, etiology, location, state, and food type (`IFSAC Category`)
```{r}
# Checking the distribution of Hospitalization
training_data_final%>%
  ggplot(aes(Hospitalizations))+
  geom_histogram()+
  scale_x_continuous(trans=scales::pseudo_log_trans(base=10))
# Creating a tibble only includes our predictors of interest
hosp_data=training_data_final%>%
  select(c(1,4,5,7,11,13,14))%>%
  mutate_if(is.character, as.factor)
glimpse(hosp_data)
ggsave("histogram_Hospitalizations.png",path=image_path)
```
### Modeling `Hospitalizations` using linear regression and Poisson regression
```{r}
# Model specification and fitting for linear regression model
lm_spec=linear_reg()%>%
  set_engine(engine="lm")
# Modeling with main predictor `Simplified_Etiology`
lm_fit3=lm_spec%>%
  fit(Hospitalizations~Simplified_Etiology, data=hosp_data)
# Modeling with all predictors
lm_fit4=lm_spec%>%
  fit(Hospitalizations~., data=hosp_data)
# Model specification and fitting for Poisson regression model
# create a workflow for poisson regression model, we will use the default glm engine
# Modeling with main predictor `Simplified_Etiology`
poisson_wf3=workflow(Hospitalizations~Simplified_Etiology, poisson_reg())
# Modeling with all predictors
poisson_wf4=workflow(Hospitalizations~., poisson_reg())
# Fitting models
poisson_fit3=fit(poisson_wf3,data=hosp_data)
poisson_fit4=fit(poisson_wf4,data=hosp_data)
```

### Model evaluation
```{r}
Result=lm_fit3%>%
 predict(new_data=hosp_data)%>%
 mutate (.pred=round(.pred),true_value=hosp_data$Hospitalizations, model= "lm_single")%>%
bind_rows(lm_fit4%>%
            predict(new_data=hosp_data)%>%
            mutate (.pred=round(.pred),true_value=hosp_data$Hospitalizations, model= "lm_multiple")%>%
            bind_rows(poisson_fit1%>%
                        predict(new_data= hosp_data)%>% 
                        mutate (true_value=hosp_data$Hospitalizations, model= "glm_single")%>%
                        bind_rows(poisson_fit2%>%
                                    predict(new_data= hosp_data)%>% 
                                    mutate (true_value=hosp_data$Hospitalizations, model= "glm_multiple")
                        )
            )
)
Result%>%
  group_by(model)%>%
  rmse(truth=true_value,estimate=.pred)

```

### 2023-03-16 code/processing_code/processingfile_v1.qmd
statistical_analysis_code

@@
   "data","raw_data","NationalOutbreakPublicDataTool.xlsx")
-testing_data <- readxl::read_excel(data_location)
+analysis_data <- readxl::read_excel(data_location)
 ```
@@
 
-## Testing Data
+## Analysis Data
 ```{r}

### 2023-03-16 code/analysis_code/exploratory_analysis.qmd
update

@@
   filter(Simplified_Etiology=="Salmonella")%>%
-  filter(`IFSAC Category`!= %n% C("Multiple","Other"))%>%
+  filter(`IFSAC Category` != "Multiple")%>%
+  filter(`IFSAC Category` != "Other") %>%
   count()%>%

### 2023-03-16 code/analysis_code/statistical_analysis.qmd
update

@@
 ```{r}
+# Pathing
 library(here)
+
+# Data Handling and Modeling
 library(tidyverse)
@@
 ```
-# Load data
+# Load Data
 ```{r}
@@
 ```
-## First health outcome of interest: Illnesses
-### Predictors: time (year), pathogen type, etiology, location, state, and food type (`IFSAC Category`)
+# First Health Outcome of Interest: Illnesses
+
+Predictors
+-   Time (Year)
+-   Pathogen Type
+-   Etiology
+-   Location
+-   State
+-   Food Type (IFSAC Category)
+
 ```{r}
-# Checking the distribution of Illnesses
-training_data_final%>%
+# Checking Illness Distribution
+training_data_final %>%
   ggplot(aes(Illnesses))+
@@
 ggsave("histogram_Illnesses.png", path = image_path, width = 3000, height = 2000, units = "px")
-# Creating a tibble only includes our predictors of interest
-illness_data=training_data_final%>%
-  select(c(1,4,5,7,10,13,14))%>%
+
+# Tibble with Predictors of Interest
+illness_data=training_data_final %>%
+  select(c(1, 4, 5, 7, 10, 13, 14)) %>%
   mutate_if(is.character, as.factor)
@@
 ```
-### Modeling `Illness` using linear regression and Poisson regression
-#### Since the outcome is counts and the event ocurrs independently, it follows a Poisson distribution
+
+## Modeling Illness with Linear Regression and Poisson Regression
+
+Since the outcome is counts/frequency and the event occurs independently, it follows a Poisson distribution. To create a workflow for poisson regression model, the default *glm* engine will be used.
+
 ```{r}
-# Model specification and fitting for linear regression model
-lm_spec=linear_reg()%>%
-  set_engine(engine="lm")
-# Modeling with main predictor `Simplified_Etiology`
-lm_fit1=lm_spec%>%
-  fit(Illnesses~Simplified_Etiology, data=illness_data)
-# Modeling with all predictors
-lm_fit2=lm_spec%>%
-  fit(Illnesses~., data=illness_data)
-# Model specification and fitting for Poisson regression model
-# create a workflow for poisson regression model, we will use the default glm engine
-# Modeling with main predictor `Simplified_Etiology`
-poisson_wf1=workflow(Illnesses~Simplified_Etiology, poisson_reg())
-# Modeling with all predictors
-poisson_wf2=workflow(Illnesses~., poisson_reg())
-# Fitting models
-poisson_fit1=fit(poisson_wf1,data=illness_data)
-poisson_fit2=fit(poisson_wf2,data=illness_data)
+# Linear Regression
+lm_spec = linear_reg() %>%
+  set_engine(engine = "lm")
+## Modeling with Main Predictor `Simplified_Etiology`
+lm_fit1 = lm_spec %>%
+  fit(Illnesses ~ Simplified_Etiology, data = illness_data)
+## Modeling with All Predictors
+lm_fit2 = lm_spec %>%
+  fit(Illnesses ~ ., data = illness_data)
+
+# Poisson Regression
+## Modeling with Main Predictor `Simplified_Etiology`
+poisson_wf1 = workflow(Illnesses ~ Simplified_Etiology, poisson_reg())
+## Modeling with All Predictors
+poisson_wf2 = workflow(Illnesses ~ ., poisson_reg())
+## Model Fitting
+poisson_fit1 = fit(poisson_wf1, data=illness_data)
+poisson_fit2 = fit(poisson_wf2, data=illness_data)
 ```
@@
 ```{r}
-# Checking parameters for linear regression model
+# Checking Parameters for Linear Regression Model
 tidy(lm_fit1) %>%
   arrange(-estimate)
-tidy(lm_fit2)%>%
+tidy(lm_fit2) %>%
   arrange(-estimate)
-# Checking parameters for Poisson regression model
-tidy(poisson_fit1)%>%
+
+# Checking Parameters for Poisson Regression Model
+tidy(poisson_fit1) %>%
  arrange(-estimate)
-tidy(poisson_fit2)%>%
+tidy(poisson_fit2) %>%
  arrange(-estimate)
-# Model evaluation
-Result=lm_fit1%>%
- predict(new_data=illness_data)%>%
- mutate (.pred=round(.pred),true_value=illness_data$Illnesses, model= "lm_single")%>%
-bind_rows(lm_fit1%>%
-            predict(new_data=illness_data)%>%
-            mutate (.pred=round(.pred),true_value=illness_data$Illnesses, model= "lm_multiple")%>%
-            bind_rows(poisson_fit1%>%
+
+# Model Evaluation
+Result = lm_fit1 %>% 
+  predict(new_data = illness_data) %>% 
+  mutate(.pred = round(.pred), 
+        true_value = illness_data$Illnesses, model= "lm_single") %>% 
+  bind_rows(lm_fit1 %>% 
+              predict(new_data=illness_data) %>% 
+              mutate(.pred=round(.pred), 
+                     true_value = illness_data$Illnesses, model= "lm_multiple") %>% 
+              bind_rows(poisson_fit1 %>% predict(new_data= illness_data) %>%
+                          mutate(true_value=illness_data$Illnesses, 
+                                 model= "glm_single") %>% 
+                          bind_rows(poisson_fit2 %>% 
                                       predict(new_data= illness_data)%>% 
-                        mutate (true_value=illness_data$Illnesses, model= "glm_single")%>%
-                        bind_rows(poisson_fit2%>%
-                                    predict(new_data= illness_data)%>% 
-                                    mutate (true_value=illness_data$Illnesses, model= "glm_multiple")
+                                      mutate(true_value=illness_data$Illnesses, 
+                                             model= "glm_multiple")))
 )
-            )
-)
-Result%>%
-  group_by(model)%>%
-  rmse(truth=true_value,estimate=.pred)
+
+Result <- Result %>%
+  group_by(model) %>%
+  rmse(truth=true_value, estimate = .pred) %>%
+  rename(illness_model = model)
+
+saveRDS(Result, file = here("results", "illness_model_evaluation.rds"))
 
 ```
-### According to the result, the model built using Poisson regression with multiple predictors seems to have the smallest root mean squared error.
-## Second health outcome of interest: Hospitilization
-### Predictors: time (year), pathogen type, etiology, location, state, and food type (`IFSAC Category`)
+According to the result, the model built using Poisson regression with multiple predictors seems to have the smallest root mean squared error.
+
+# Second Health Outcome of Interest: Hospitilization
+
+Predictors
+-   Time (Year)
+-   Pathogen Type
+-   Etiology
+-   Location
+-   State
+-   Food Type (IFSAC Category)
+
 ```{r}
-# Checking the distribution of Hospitalization
-training_data_final%>%
-  ggplot(aes(Hospitalizations))+
-  geom_histogram()+
-  scale_x_continuous(trans=scales::pseudo_log_trans(base=10))
-# Creating a tibble only includes our predictors of interest
-hosp_data=training_data_final%>%
-  select(c(1,4,5,7,11,13,14))%>%
+# Checking the Distribution of Hospitalization
+training_data_final %>%
+  ggplot(aes(Hospitalizations)) +
+  geom_histogram() +
+  scale_x_continuous(trans=scales::pseudo_log_trans(base = 10))
+# Tibble for Variables of Interest
+hosp_data = training_data_final %>%
+  select(c(1, 4, 5, 7, 11, 13, 14))%>%
   mutate_if(is.character, as.factor)
 glimpse(hosp_data)
-ggsave("histogram_Hospitalizations.png",path=image_path)
-```
-### Modeling `Hospitalizations` using linear regression and Poisson regression
-```{r}
-# Model specification and fitting for linear regression model
-lm_spec=linear_reg()%>%
-  set_engine(engine="lm")
-# Modeling with main predictor `Simplified_Etiology`
-lm_fit3=lm_spec%>%
-  fit(Hospitalizations~Simplified_Etiology, data=hosp_data)
-# Modeling with all predictors
-lm_fit4=lm_spec%>%
-  fit(Hospitalizations~., data=hosp_data)
-# Model specification and fitting for Poisson regression model
-# create a workflow for poisson regression model, we will use the default glm engine
-# Modeling with main predictor `Simplified_Etiology`
-poisson_wf3=workflow(Hospitalizations~Simplified_Etiology, poisson_reg())
-# Modeling with all predictors
-poisson_wf4=workflow(Hospitalizations~., poisson_reg())
-# Fitting models
-poisson_fit3=fit(poisson_wf3,data=hosp_data)
-poisson_fit4=fit(poisson_wf4,data=hosp_data)
+ggsave("histogram_Hospitalizations.png", path = image_path)
 ```
 
-### Model evaluation
+## Modeling `Hospitalizations` using Linear Regression and Poisson Regression
+
+To create a workflow for the Poisson regression model, we will use the default *glm* engine again.
+
 ```{r}
-Result=lm_fit3%>%
- predict(new_data=hosp_data)%>%
- mutate (.pred=round(.pred),true_value=hosp_data$Hospitalizations, model= "lm_single")%>%
-bind_rows(lm_fit4%>%
-            predict(new_data=hosp_data)%>%
-            mutate (.pred=round(.pred),true_value=hosp_data$Hospitalizations, model= "lm_multiple")%>%
-            bind_rows(poisson_fit1%>%
-                        predict(new_data= hosp_data)%>% 
-                        mutate (true_value=hosp_data$Hospitalizations, model= "glm_single")%>%
-                        bind_rows(poisson_fit2%>%
-                                    predict(new_data= hosp_data)%>% 
-                                    mutate (true_value=hosp_data$Hospitalizations, model= "glm_multiple")
+# Model Specification and Fitting for Linear Regression Model
+lm_spec = linear_reg() %>%
+  set_engine(engine = "lm")
+## Modeling with Main Predictor `Simplified_Etiology`
+lm_fit3 = lm_spec %>%
+  fit(Hospitalizations ~ Simplified_Etiology, data = hosp_data)
+## Modeling with All Predictors
+lm_fit4 = lm_spec %>%
+  fit(Hospitalizations ~ ., data = hosp_data)
+
+# Model Specification and Fitting for Poisson Regression Model
+## Modeling with Main Predictor `Simplified_Etiology`
+poisson_wf3 = workflow(Hospitalizations ~ Simplified_Etiology, poisson_reg())
+## Modeling with all predictors
+poisson_wf4=workflow(Hospitalizations~., poisson_reg())
+## Fitting models
+poisson_fit3 = fit(poisson_wf3,data = hosp_data)
+poisson_fit4 = fit(poisson_wf4,data = hosp_data)
+```
+
+### Model Evaluation
+```{r}
+Result = lm_fit3 %>%
+ predict(new_data = hosp_data) %>%
+ mutate(.pred = round(.pred), true_value = hosp_data$Hospitalizations,
+        model= "lm_single") %>% 
+  bind_rows(lm_fit4 %>% 
+              predict(new_data=hosp_data) %>% 
+              mutate (.pred=round(.pred), 
+                      true_value=hosp_data$Hospitalizations, 
+                      model= "lm_multiple") %>% 
+              bind_rows(poisson_fit1 %>% 
+                        predict(new_data = hosp_data) %>%
+                        mutate(true_value = hosp_data$Hospitalizations, 
+                               model= "glm_single") %>% 
+                          bind_rows(poisson_fit2 %>% 
+                                      predict(new_data= hosp_data) %>% 
+                                      mutate(true_value = hosp_data$Hospitalizations, 
+                                             model= "glm_multiple")))
 )
-            )
-)
-Result%>%
+
+Result <- Result %>%
   group_by(model)%>%
-  rmse(truth=true_value,estimate=.pred)
+  rmse(truth = true_value, estimate = .pred) %>%
+  rename(hospital_model = model)
+
+saveRDS(Result, file = here("results", "hospital_model_evaluation.rds"))
 
 ```
+
+According to the result, the model built using linear regression with multiple predictors seems to have the smallest root mean squared error.

### 2023-03-16 products/manuscript/Manuscript.qmd
update

@@
 
-We will be building regression models using these predictors and health outcomes as well; we intend to implement machine learning to train a model to predict health outcomes using the Kaggle dataset for the CDC NORS data.
+Afterwards, we built regression models using these predictors and health outcomes; we intend to implement machine learning to train a model to predict health outcomes using the Kaggle dataset for the CDC NORS data.
 
@@
 ## Basic Statistical Analysis
+Multivariate Poisson regression performed for illness as the health outcome proved to be the best model (RMSE = 55.6) among simple linear, multiple linear, simple Poisson, and multivariate Poisson regressions.
+
+```{r echo=FALSE}
+illness_model_evaluated <- readRDS(here("results", "illness_model_evaluation.rds"))
+rmarkdown::paged_table(illness_model_evaluated)
+```
+
+For hospitalization as the health outcome, multivariate linear regression was the best model (RMSE = 7.09) among those previously listed.
+
+```{r echo=FALSE}
+illness_model_evaluated <- readRDS(here("results", "hospital_model_evaluation.rds"))
+rmarkdown::paged_table(illness_model_evaluated)
+```
 

### 2023-03-17 code/processing_code/processingfile_v1.qmd
update2

--- 2023-03-14 code/processing_code/processingfile_v1.qmd
@@
 ```{r}
-str(analysis_data)
-summary(analysis_data)
+str(testing_data)
+summary(testing_data)
 
-# Closer Look at Training Data
-head(analysis_data)
-skim(analysis_data)
+# Closer Look at Testing Data
+head(testing_data)
+skim(testing_data)
 
 # Plotting NAs of Training Data
-plot_na_pareto(analysis_data)
+plot_na_pareto(testing_data)
 ```
@@
 
-Because the analysis dataset contains more than just foodborne outbreaks, we want to specify it to just foodborne illnesses.
+Because the testing dataset contains more than just foodborne outbreaks, we want to specify it to just foodborne illnesses.
 
@@
 # Plot of NAs After Limiting to Analysis Data to Food
-plot_na_pareto(analysis_data %>% filter(`Primary Mode` == "Food"))
+plot_na_pareto(testing_data %>% filter(`Primary Mode` == "Food"))
 ```
@@
 # Analysis Data
-unique(analysis_data$`Etiology Status`)
+unique(testing_data$`Etiology Status`)
 ```
@@
 ```{r}
-analysis_data_fin <- analysis_data %>%
+testing_data_fin <- testing_data %>%
   select(!c(`Animal Type`, `Animal Type Specify`, 
@@
 ```{r}
-analysis_data_fin <- analysis_data_fin %>%
+testing_data_fin <- testing_data_fin %>%
   separate_rows(Etiology, sep = "; ")
@@
 
-analysis_data_fin <- analysis_data_fin %>%
+testing_data_fin <- testing_data_fin %>%
   mutate(`IFSAC Category` = case_when(
@@
 
-analysis_data_fin <- analysis_data_fin %>%
+testing_data_fin <- testing_data_fin %>%
   mutate(Pathogen_Type = case_when(
@@
 ```{r}
-analysis_data_fin_re <- analysis_data_fin %>%
+testing_data_fin <- analysis_data_fin %>%
   rename(Location=Setting)%>%
@@
     separate_rows(Location,sep="; ")
-analysis_data_fin_re
+testing_data_fin
 ```
@@
   drop_na()
-analysis_data_final=analysis_data_fin_re%>%
+testing_data_final=testing_data_fin%>%
   drop_na()
@@
 skim(training_data_final)
-skim(analysis_data_final)
+skim(testing_data_final)
 ```
@@
 processed_training_data <- training_data_final
-processed_analysis_data <- analysis_data_final
+processed_testing_data <- testing_data_final
 
@@
 
-save_data_location2=here::here("data","processed_data","processed_analysis_data.rds")
-saveRDS(processed_analysis_data, file = save_data_location2)
+save_data_location2=here::here("data","processed_data","processed_testing_data.rds")
+saveRDS(processed_testing_data, file = save_data_location2)
 

### 2023-03-17 products/manuscript/Manuscript.qmd
update2

@@
 There are 12 variables and 16,950+ observations.
+This data set is intended to be used for model building.
 
@@
 The raw datasets may be viewed within the repository of this project (data/raw_data/).
+This data set is collected with the purpose of evaluating the model performance in the end.
 
@@
 
-Afterwards, we built regression models using these predictors and health outcomes; we intend to implement machine learning to train a model to predict health outcomes using the Kaggle dataset for the CDC NORS data.
+Afterwards, we built regression models including linear and Poisson regression using these predictors and health outcomes from training data set [@cdc2017]; data was re-sampled using bootstrapping approach for 1000 times for a better estimate of the accuracy and variability of the model coefficients; we intend to implement random forest to train a model to predict health outcomes using the Kaggle dataset for the CDC NORS data.
+The performance of models was compared across different metrics.
+The best model was ultimately employed to predict the outcome based on predictors in the testing data set [@cdc2020a].
 

### 2023-03-30 code/analysis_code/exploratory_analysis.qmd
update according to comments

@@
 ---
-title: "An example exploratory analysis script"
-author: "Andreas Handel"
+title: "Exploratory analysis script"
+author: "Weifan Wu and Kailin Chen"
 date: "10/28/2021"
@@
          select(Pathogen_Type, Simplified_Etiology, Hospitalizations) %>% filter(Pathogen_Type == "Bacteria"), aes(x = Simplified_Etiology)) + geom_bar() + 
-  labs(x = "Bacteria", y = "Hospitalizations", title = "Bar Chart of Bacteria Frequency")
+  labs(x = "Bacteria", y = "Hospitalizations", title = "Bar Chart of Bacteria Frequency")+
+  coord_flip()
 
@@
   geom_boxplot()+
-  scale_x_discrete(guide=guide_axis(n.dodge=3))
+  coord_flip()
 
@@
 California and Minnesota are the two states with the highest total hospitalization count due to foodborne illness from 1998 to 2015.
-Resturants and Private home are the most frequent locations with the highest total hospitalization count.
+Restaurants and Private home are the most frequent locations with the highest total hospitalization count.
 
-## Analysis Data
+## Testing Data
 ```{r}
@@
 ## Closer Look at Bacterial Etiologies
-ggplot(analysis_data_final %>% group_by(Pathogen_Type, Simplified_Etiology, Illnesses) %>% filter(Pathogen_Type == "Bacteria"), aes(x = Simplified_Etiology)) + geom_bar() + labs(x = "Bacteria", y = "Cases", title = "Bar Chart of Bacteria Frequency")
+ggplot(analysis_data_final %>% 
+         group_by(Pathogen_Type, Simplified_Etiology, Illnesses) %>% 
+         filter(Pathogen_Type == "Bacteria"), aes(x = Simplified_Etiology)) + 
+  geom_bar() + 
+  labs(x = "Bacteria", y = "Cases", title = "Bar Chart of Bacteria Frequency")+
+  coord_flip()
 
@@
 
-state_v_illness_analysis <- analysis_data_final %>% group_by(State) %>% mutate(all_illnesses = sum(Illnesses)) %>% count(all_illnesses) %>% select(State, all_illnesses) %>% arrange(desc(all_illnesses))
+state_v_illness_analysis <- analysis_data_final %>% 
+  group_by(State) %>%
+  mutate(all_illnesses = sum(Illnesses)) %>%
+  count(all_illnesses) %>% 
+  select(State, all_illnesses) %>% 
+  arrange(desc(all_illnesses))
 
@@
 # Food Type and Illnesses
-ggplot(analysis_data_final %>% filter(`IFSAC Category` != "Multiple") %>% filter(`IFSAC Category` != "Other"), aes(x = `IFSAC Category`, y = Illnesses)) + geom_col() + labs(x = "Food Type", y = "Illnesses", title = "Illnesses by Food Type, Excluding 'Multiple' and 'Other'")+
-  scale_x_discrete(guide = guide_axis(n.dodge=3))
+ggplot(analysis_data_final %>% filter(`IFSAC Category` != "Multiple") %>% filter(`IFSAC Category` != "Other"), aes(x = `IFSAC Category`, y = Illnesses)) + 
+  geom_col() + 
+  labs(x = "Food Type", y = "Illnesses", title = "Illnesses by Food Type, Excluding 'Multiple' and 'Other'")+
+  coord_flip()
 
@@
   labs(x = "Locations", y = "Illnesses", title = "Illnesses by Locations")+
-  scale_x_discrete(guide=guide_axis(n.dodge=3))
+  coord_flip()
 

### 2023-03-30 code/processing_code/processingfile_v1.qmd
update according to comments

@@
 title: "Data Pre-Processing/Cleaning Script"
-author: "Andreas Handel"
+author: "Weifan Wu and Kailin Chen"
 date: "2023-01-03"
@@
 ```{r}
-testing_data_fin <- analysis_data_fin %>%
+testing_data_fin <- testing_data_fin %>%
   rename(Location=Setting)%>%

### 2023-03-30 products/manuscript/Manuscript.qmd
update according to comments

@@
 subtitle: ""
-author: Andreas Handel
+author: Weifan Wu and Kailin Chen
 date: "`r Sys.Date()`"

### 2023-04-05 code/analysis_code/exploratory_analysis.qmd
update linear&Poisson regression model and add random forest

@@
 ## Save Graph
-image_path <- here("results")
+image_path <- here("results","images_EDA")
 ggsave("time_v_hospitalization_pathogen_type.png", path = image_path, width = 3000, height = 2000, units = "px")

### 2023-04-05 code/analysis_code/statistical_analysis.qmd
update according to comments
update linear&Poisson regression model and add random forest

@@
-# Load libraries
+---
+title: "Statistical analysis script"
+author: "Weifan Wu and Kailin Chen"
+date: "03/16/2023"
+output: html_document
+---
+
+## Load libraries
 ```{r}
@@
 library(poissonreg)
+library(ranger)
+library(knitr)
+library(skimr)
 ```
-# Load Data
+## Load Data
 ```{r}
@@
 training_data_final <- readRDS(data_location)
-glimpse(training_data_final)
+skim(training_data_final)
 ```
-# First Health Outcome of Interest: Illnesses
+## First Health Outcome of Interest: Illnesses
 
@@
   scale_x_continuous(trans=scales::pseudo_log_trans(base=10))
-image_path <- here("results")
-ggsave("histogram_Illnesses.png", path = image_path, width = 3000, height = 2000, units = "px")
+images_stats_path=here("results","images_statistic_analysis")
+ggsave("histogram_Illnesses.png", path = images_stats_path, width = 3000, height = 2000, units = "px")
 
-# Tibble with Predictors of Interest
+# Create a tibble with predictors of interest
 illness_data=training_data_final %>%
-  select(c(1, 4, 5, 7, 10, 13, 14)) %>%
+  select(c(1, 4, 5, 7, 10, 14)) %>%
+  filter(!`IFSAC Category`%in%c("Multiple","Other"),Location!="Unknown")%>%
   mutate_if(is.character, as.factor)
-glimpse(illness_data)
+skim(illness_data)
 ```
@@
 ```{r}
-# Linear Regression
+set.seed(123)
+# Set up cross validation
+illness_fold=vfold_cv(illness_data,repeats = 2,strata = Illnesses)
+
+# Data pre-processing
+## Recipe for main predictor
+illness_main_rec=recipe(Illnesses~Simplified_Etiology,data=illness_data)%>%
+  step_dummy(Simplified_Etiology)
+## Recipe for all predictors
+### Lump small proportion into "other" using step_other
+### Create dummy code for nominal variables
+illness_all_rec=recipe(Illnesses~.,data=illness_data)%>%
+  step_other(State,Location,`IFSAC Category`,Simplified_Etiology,threshold = 0.01)%>%
+  step_dummy(all_nominal())
+### Checking processed data set
+illness_all_rec%>%
+  prep()%>%
+  bake(new_data=NULL)
+
+# Model specification
+## Linear Regression
 lm_spec = linear_reg() %>%
   set_engine(engine = "lm")
+
+# Creating workflow
+## Main Predictor `Simplified_Etiology`
+lm_wf1 = workflow() %>%
+ add_recipe(illness_main_rec)%>%
+  add_model(lm_spec)
+## Modeling with All Predictors
+lm_wf2 = workflow() %>%
+  add_recipe(illness_all_rec)%>%
+  add_model(lm_spec)
+## Poisson Regression
 ## Modeling with Main Predictor `Simplified_Etiology`
-lm_fit1 = lm_spec %>%
-  fit(Illnesses ~ Simplified_Etiology, data = illness_data)
+poisson_wf1 = workflow()%>%
+  add_recipe(illness_main_rec)%>%
+  add_model(poisson_reg())
 ## Modeling with All Predictors
-lm_fit2 = lm_spec %>%
-  fit(Illnesses ~ ., data = illness_data)
+poisson_wf2 = workflow()%>%
+  add_recipe(illness_all_rec)%>%
+  add_model(poisson_reg())
+```
 
-# Poisson Regression
-## Modeling with Main Predictor `Simplified_Etiology`
-poisson_wf1 = workflow(Illnesses ~ Simplified_Etiology, poisson_reg())
-## Modeling with All Predictors
-poisson_wf2 = workflow(Illnesses ~ ., poisson_reg())
-## Model Fitting
-poisson_fit1 = fit(poisson_wf1, data=illness_data)
-poisson_fit2 = fit(poisson_wf2, data=illness_data)
+## Model fitting
+```{r}
+## Linear regression
+### Main predictor
+set.seed(234)
+lm_rs1=lm_wf1%>%
+fit_resamples(resamples=illness_fold,
+    control=control_resamples(save_pred=TRUE))
+
+### All predictors
+lm_rs2=lm_wf2%>%
+  fit_resamples(resamples=illness_fold,
+  control=control_resamples(save_pred=TRUE))
+
+## Poisson regression
+### Main predictor
+poisson_rs1 = poisson_wf1%>%
+  fit_resamples(resamples=illness_fold,
+    control=control_resamples(save_pred=TRUE))
+
+### All predictors
+poisson_rs2 = poisson_wf2%>%
+  fit_resamples(resamples=illness_fold,
+    control=control_resamples(save_pred=TRUE))
 ```
@@
 ```{r}
+# Checking metrics for Linear Regression Model
+lm_rs1%>%
+  collect_metrics()
+lm_rs2%>%
+  collect_metrics()
+# Checking metrics for poisson regression model
+poisson_rs1%>%
+  collect_metrics()
+poisson_rs2%>%
+  collect_metrics()
 # Checking Parameters for Linear Regression Model
-tidy(lm_fit1) %>%
+lm_fit1=lm_wf1%>%
+  fit(illness_data)%>%
+  tidy()%>%
   arrange(-estimate)
-tidy(lm_fit2) %>%
+
+lm_fit2=lm_wf2%>%
+  fit(illness_data)%>%
+  tidy()%>%
   arrange(-estimate)
@@
 # Checking Parameters for Poisson Regression Model
-tidy(poisson_fit1) %>%
- arrange(-estimate)
-tidy(poisson_fit2) %>%
+poisson_fit1=poisson_wf1%>%
+  fit(illness_data)%>%
+  tidy()%>%
   arrange(-estimate)
 
-# Model Evaluation
-Result = lm_fit1 %>% 
-  predict(new_data = illness_data) %>% 
-  mutate(.pred = round(.pred), 
-        true_value = illness_data$Illnesses, model= "lm_single") %>% 
-  bind_rows(lm_fit1 %>% 
-              predict(new_data=illness_data) %>% 
-              mutate(.pred=round(.pred), 
-                     true_value = illness_data$Illnesses, model= "lm_multiple") %>% 
-              bind_rows(poisson_fit1 %>% predict(new_data= illness_data) %>%
-                          mutate(true_value=illness_data$Illnesses, 
-                                 model= "glm_single") %>% 
-                          bind_rows(poisson_fit2 %>% 
-                                      predict(new_data= illness_data)%>% 
-                                      mutate(true_value=illness_data$Illnesses, 
-                                             model= "glm_multiple")))
-)
+poisson_fit2=poisson_wf2%>%
+  fit(illness_data)%>%
+  tidy()%>%
+  arrange(-estimate)
 
-Result <- Result %>%
-  group_by(model) %>%
-  rmse(truth=true_value, estimate = .pred) %>%
-  rename(illness_model = model)
-
-saveRDS(Result, file = here("results", "illness_model_evaluation.rds"))
+# Plotting predictions vs Illnesses for Linear model
+lm_rs1%>%
+  collect_predictions()%>%
+  filter(Illnesses<300)%>%
+  ggplot(aes(Illnesses,.pred,color=id2,group=id2))+
+  geom_point()+
+  geom_smooth(se=FALSE)+
+  labs(y="Predicted Illnesses",title="Predicted Illnesses VS Illnesses by linear regression with main predictor")
+ggsave("Predicted Illnesses VS Illnesses by linear regression with main predictor.png",path=images_stats_path,width=3000,height=2000,units="px")
+lm_rs2%>%
+  collect_predictions()%>%
+  #filter(Illnesses<300)%>%
+  ggplot(aes(Illnesses,.pred,color=id2,group=id2))+
+  geom_point()+
+  geom_smooth(se=FALSE)+
+  labs(y="Predicted Illnesses",title="Predicted Illnesses VS Illnesses by linear regression with all predictors")
+ggsave("Predicted Illnesses VS Illnesses by linear regression with all predictors.png",path=images_stats_path,width=3000,height=2000,units="px")
+# Plotting predictions vs Illnesses for Poisson model
+poisson_rs1%>%
+  collect_predictions()%>%
+  #filter(Illnesses<300)%>%
+  ggplot(aes(Illnesses,.pred,color=id2,group=id2))+
+  geom_point()+
+  geom_smooth(se=FALSE)+
+  labs(y="Predicted Illnesses",title="Predicted Illnesses VS Illnesses by linear regression with main predictor")
+ggsave("Predicted Illnesses VS Illnesses by poisson regression with main predictor.png",path=images_stats_path,width=3000,height=2000,units="px")
+poisson_rs2%>%
+  collect_predictions()%>%
+  #filter(Illnesses<500)%>%
+  ggplot(aes(Illnesses,.pred,color=id2,group=id2))+
+  geom_point()+
+  geom_smooth(se=FALSE)+
+  labs(y="Predicted Illnesses",title="Predicted Illnesses VS Illnesses by linear regression with all predictors")
+ggsave("Predicted Illnesses VS Illnesses by poisson regression with all predictors.png",path=images_stats_path,width=3000,height=2000,units="px")
 
 ```
-According to the result, the model built using Poisson regression with multiple predictors seems to have the smallest root mean squared error.
+According to the result, the model built using Poisson regression with multiple predictors seems to have the smallest root mean squared error and highest R^2.
+However, none of those models can catch the outliers (Illness>1000)
 
@@
 hosp_data = training_data_final %>%
-  select(c(1, 4, 5, 7, 11, 13, 14))%>%
+  select(c(1, 4, 5, 7, 11, 14))%>%
+  filter(!`IFSAC Category`%in%c("Multiple","other"),!Location%in%c("Unknown"))%>%
   mutate_if(is.character, as.factor)
-glimpse(hosp_data)
-ggsave("histogram_Hospitalizations.png", path = image_path)
+skim(hosp_data)
+ggsave("histogram_Hospitalizations.png", path = images_stats_path)
 ```
@@
 ```{r}
-# Model Specification and Fitting for Linear Regression Model
+set.seed(1234)
+# Set up cross validation
+hosp_fold=vfold_cv(hosp_data,repeats = 2,strata = Hospitalizations)
+
+# Data pre-processing
+## Recipe for main predictor
+hosp_main_rec=recipe(Hospitalizations ~ Simplified_Etiology,data=hosp_data)%>%
+  step_dummy(Simplified_Etiology)
+## Recipe for all predictors
+### Lump small proportion into "other" using step_other
+### Create dummy code for nominal variables
+hosp_all_rec=recipe(Hospitalizations ~.,data=hosp_data)%>%
+  step_other(State,Location,`IFSAC Category`, Simplified_Etiology,threshold = 0.01)%>%
+  step_dummy(all_nominal_predictors())
+### Checking processed data set
+hosp_all_rec%>%
+  prep()%>%
+  bake(new_data=NULL)
+
+# Model specification
+## Linear Regression
 lm_spec = linear_reg() %>%
   set_engine(engine = "lm")
+
+# Creating workflow
+## Main Predictor `Simplified_Etiology`
+lm_wf3 = workflow() %>%
+ add_recipe(hosp_main_rec)%>%
+  add_model(lm_spec)
+## Modeling with All Predictors
+lm_wf4 = workflow() %>%
+  add_recipe(hosp_all_rec)%>%
+  add_model(lm_spec)
+## Poisson Regression
 ## Modeling with Main Predictor `Simplified_Etiology`
-lm_fit3 = lm_spec %>%
-  fit(Hospitalizations ~ Simplified_Etiology, data = hosp_data)
+poisson_wf3 = workflow()%>%
+  add_recipe(hosp_main_rec)%>%
+  add_model(poisson_reg())
 ## Modeling with All Predictors
-lm_fit4 = lm_spec %>%
-  fit(Hospitalizations ~ ., data = hosp_data)
+poisson_wf4 = workflow()%>%
+  add_recipe(hosp_all_rec)%>%
+  add_model(poisson_reg())
+```
+## Model fitting
+```{r}
+## Linear regression
+### Main predictor
+set.seed(2345)
+lm_rs3=lm_wf3%>%
+fit_resamples(resamples=hosp_fold,
+    control=control_resamples(save_pred=TRUE))
 
-# Model Specification and Fitting for Poisson Regression Model
-## Modeling with Main Predictor `Simplified_Etiology`
-poisson_wf3 = workflow(Hospitalizations ~ Simplified_Etiology, poisson_reg())
-## Modeling with all predictors
-poisson_wf4=workflow(Hospitalizations~., poisson_reg())
-## Fitting models
-poisson_fit3 = fit(poisson_wf3,data = hosp_data)
-poisson_fit4 = fit(poisson_wf4,data = hosp_data)
+### All predictors
+lm_rs4=lm_wf4%>%
+  fit_resamples(resamples=hosp_fold,
+  control=control_resamples(save_pred=TRUE))
+
+## Poisson regression
+### Main predictor
+poisson_rs3 = poisson_wf3%>%
+  fit_resamples(resamples=hosp_fold,
+    control=control_resamples(save_pred=TRUE))
+
+### All predictors
+poisson_rs4 = poisson_wf4%>%
+  fit_resamples(resamples=hosp_fold,
+    control=control_resamples(save_pred=TRUE))
 ```
-
 ### Model Evaluation
 ```{r}
-Result = lm_fit3 %>%
- predict(new_data = hosp_data) %>%
- mutate(.pred = round(.pred), true_value = hosp_data$Hospitalizations,
-        model= "lm_single") %>% 
-  bind_rows(lm_fit4 %>% 
-              predict(new_data=hosp_data) %>% 
-              mutate (.pred=round(.pred), 
-                      true_value=hosp_data$Hospitalizations, 
-                      model= "lm_multiple") %>% 
-              bind_rows(poisson_fit1 %>% 
-                        predict(new_data = hosp_data) %>%
-                        mutate(true_value = hosp_data$Hospitalizations, 
-                               model= "glm_single") %>% 
-                          bind_rows(poisson_fit2 %>% 
-                                      predict(new_data= hosp_data) %>% 
-                                      mutate(true_value = hosp_data$Hospitalizations, 
-                                             model= "glm_multiple")))
-)
+# Checking metrics for Linear Regression Model
+lm_rs3%>%
+  collect_metrics()
+lm_rs4%>%
+  collect_metrics()
+# Checking metrics for poisson regression model
+poisson_rs3%>%
+  collect_metrics()
+poisson_rs4%>%
+  collect_metrics()
+# Checking Parameters for Linear Regression Model
+lm_fit3=lm_wf3%>%
+  fit(hosp_data)%>%
+  tidy()%>%
+  arrange(-estimate)
 
-Result <- Result %>%
-  group_by(model)%>%
-  rmse(truth = true_value, estimate = .pred) %>%
-  rename(hospital_model = model)
+lm_fit4=lm_wf4%>%
+  fit(hosp_data)%>%
+  tidy()%>%
+  arrange(-estimate)
 
-saveRDS(Result, file = here("results", "hospital_model_evaluation.rds"))
+# Checking Parameters for Poisson Regression Model
+poisson_fit3=poisson_wf3%>%
+  fit(hosp_data)%>%
+  tidy()%>%
+  arrange(-estimate)
+
+poisson_fit4=poisson_wf4%>%
+  fit(hosp_data)%>%
+  tidy()%>%
+  arrange(-estimate)
+
+# Plotting predictions vs Hospitalizations for Linear model
+lm_rs3%>%
+  collect_predictions()%>%
+  #filter(Hospitalizations<50)%>%
+  ggplot(aes(Hospitalizations,.pred,color=id2))+
+  geom_point()+
+  geom_smooth(se=FALSE)+
+  labs(y="Predicted Hospitalizations",title="Predicted Hospitalizations VS Hospitalizations \nby linear regression with main predictor",color="Fold")
+ggsave("Predicted Hospitalizations VS Hospitalizations by linear regression with main predictor.png",path=images_stats_path,width=3000,height=2000,units="px")
+lm_rs4%>%
+  collect_predictions()%>%
+  #filter(Hospitalizations<50)%>%
+  ggplot(aes(Hospitalizations,.pred,color=id2))+
+  geom_point()+
+  geom_smooth(se=FALSE)+
+  labs(y="Predicted Hospitalizations",title="Predicted Hospitalizations VS Hospitalizations \nby linear regression with all predictors",color="Fold")
+ggsave("Predicted Hospitalizations VS Hospitalizations by linear regression with all predictors.png",path=images_stats_path,width=3000,height=2000,units="px")
+# Plotting predictions vs Hospitalizations for Poisson model
+poisson_rs3%>%
+  collect_predictions()%>%
+  #filter(Hospitalizations<50)%>%
+  ggplot(aes(Hospitalizations,.pred,color=id2))+
+  geom_point()+
+  geom_smooth(se=FALSE)+
+  labs(y="Predicted Hospitalizations",title="Predicted Hospitalizations VS Hospitalizations \nby linear regression with main predictor",color="Fold")
+ggsave("Predicted Hospitalizations VS Hospitalizations by poisson regression with main predictor.png",path=images_stats_path,width=3000,height=2000,units="px")
+poisson_rs4%>%
+  collect_predictions()%>%
+  #filter(Hospitalizations<50)%>%
+  ggplot(aes(Hospitalizations,.pred,color=id2))+
+  geom_point()+
+  geom_smooth(se=FALSE)+
+  labs(y="Predicted Hospitalizations",title="Predicted Hospitalizations VS Hospitalizations \nby linear regression with all predictors",color="Fold")
+ggsave("Predicted Hospitalizations VS Hospitalizations by poisson regression with all predictors.png",path=images_stats_path,width=3000,height=2000,units="px")
 
@@
 
-According to the result, the model built using linear regression with multiple predictors seems to have the smallest root mean squared error.
+According to the result, the model built using Poisson regression with multiple predictors seems to have the smallest root mean squared error and highest R^2 (34%).
+But again, outliers (Hospitalizations > 100) can not be predicted well by none of those models
+
+## Machine learning approach
+## Random Forest
+### Predicting First Outcome `Illnesses` using all predictors
+```{r}
+# Model Specification
+rf_spec=rand_forest(mtry=tune(),min_n=tune(),trees=1000)%>%
+  set_engine("ranger")%>%
+  set_mode("regression")
+illness_all_rec=recipe(Illnesses~.,data=illness_data)%>%
+  step_other(State,Location,`IFSAC Category`,Simplified_Etiology,threshold = 0.01)%>%
+  step_dummy(all_nominal())
+# Creating workflow
+rf_wf1=workflow()%>%
+  add_recipe(illness_all_rec)%>%
+  add_model(rf_spec)
+# Training and tuning model using space filling design
+set.seed(345)
+doParallel::registerDoParallel()
+rf_tune1=rf_wf1%>%
+  tune_grid(
+    resamples=illness_fold,
+    grid=15,
+    control=control_grid(save_pred=TRUE),
+    metrics = metric_set(rmse)
+  )
+```
+### Model evaluation
+```{r}
+rf_tune1%>%
+  autoplot()
+rf_best1=rf_tune1%>%
+  select_best()
+rf_final1=rf_wf1%>%
+  finalize_workflow(rf_best1)
+rf_fit1=rf_final1%>%
+  fit(illness_data)
+rf_predict1=rf_fit1%>%
+  predict(illness_data)
+illness_data%>%
+  select(Illnesses)%>%
+  bind_cols(rf_predict1)%>%
+  ggplot(aes(Illnesses,.pred))+
+  geom_point()+
+  geom_abline(intercept = 0,slope=1)+
+  geom_smooth(se=FALSE)+
+  labs(y="Predicted Illnesses",title="Predicted illnesses VS illnesses by random forest model")
+ggsave("Predicted illnesses VS illnesses by random forest model.png",path=images_stats_path,width=3000,height=2000,units="px")
+```
+
+### Predicting second outcome `Hospitalizations` using all predictors
+```{r}
+
+# Creating workflow
+rf_wf2=workflow()%>%
+  add_recipe(hosp_all_rec)%>%
+  add_model(rf_spec)
+# Training and tuning model
+doParallel::registerDoParallel()
+set.seed(456)
+rf_tune2=rf_wf2%>%
+  tune_grid(
+    resamples=hosp_fold,
+    grid=15,
+    control=control_grid(save_pred=TRUE),
+    metrics = metric_set(rmse)
+  )
+```
+
+### Model evaluation
+```{r}
+rf_tune2%>%
+  autoplot()
+rf_tune2%>%
+  show_best()
+rf_best2=rf_tune2%>%
+  select_best()
+rf_final2=rf_wf2%>%
+  finalize_workflow(rf_best2)
+rf_fit2=rf_final2%>%
+  fit(hosp_data)
+rf_predict2=rf_fit2%>%
+  predict(hosp_data)
+hosp_data%>%
+  select(Hospitalizations)%>%
+  bind_cols(rf_predict2)%>%
+  ggplot(aes(Hospitalizations,.pred))+
+  geom_point()+
+  geom_abline(intercept = 0,slope=1)+
+  geom_smooth(se=FALSE)+
+  labs(y="Predicted Hospitalizations",title="Predicted Hospitalizations VS Hospitalizations by random forest model")
+ggsave("Predicted Hospitalizations VS Hospitalizations by random forest model.png",path=images_stats_path,width=3000,height=2000,units="px")
+```
+## Overall, random forest model has the lowesr rmse among those selected models when training data is fitted; Also, the random forest model outperforms other models in predicting outliers.
+
+## Fast and Frugal Trees
+## Organize four models performance and choose the best one
+## fit testing data

### 2023-04-05 code/analysis_code/statistical_analysis.qmd
add vip

@@
 library(skimr)
+library(vip)
+library(caret)
 ```
@@
 
+# Checking variable importance for Linear and Poisson regression models
+lm_wf1%>%
+  fit(illness_data)%>%
+  extract_fit_engine()%>%
+  vip()
+lm_wf2%>%
+  fit(illness_data)%>%
+  extract_fit_engine()%>%
+  vip()
+poisson_wf1%>%
+  fit(illness_data)%>%
+  extract_fit_engine()%>%
+  vip()
+poisson_wf2%>%
+  fit(illness_data)%>%
+  extract_fit_engine()%>%
+  vip()
 # Plotting predictions vs Illnesses for Linear model
@@
   collect_predictions()%>%
-  filter(Illnesses<300)%>%
+  #filter(Illnesses<300)%>%
   ggplot(aes(Illnesses,.pred,color=id2,group=id2))+
@@
   geom_smooth(se=FALSE)+
+  geom_abline(intercept=0,slope=1)+
   labs(y="Predicted Illnesses",title="Predicted Illnesses VS Illnesses by linear regression with main predictor")
@@
   geom_smooth(se=FALSE)+
+  geom_abline(intercept=0,slope=1)+
   labs(y="Predicted Illnesses",title="Predicted Illnesses VS Illnesses by linear regression with all predictors")
@@
 # Checking Parameters for Linear Regression Model
-lm_fit3=lm_wf3%>%
+lm_wf3%>%
   fit(hosp_data)%>%
@@
 
-lm_fit4=lm_wf4%>%
+lm_wf4%>%
   fit(hosp_data)%>%
@@
 # Checking Parameters for Poisson Regression Model
-poisson_fit3=poisson_wf3%>%
+poisson_wf3%>%
   fit(hosp_data)%>%
@@
 
-poisson_fit4=poisson_wf4%>%
+poisson_wf4%>%
   fit(hosp_data)%>%
@@
   arrange(-estimate)
+
+# Checking variable importance for linear and Poisson regression models
+lm_wf3%>%
+  fit(hosp_data)%>%
+  extract_fit_engine()%>%
+  vip()
+lm_wf4%>%
+  fit(hosp_data)%>%
+  extract_fit_engine()%>%
+  vip()
+poisson_wf3%>%
+  fit(hosp_data)%>%
+  extract_fit_engine()%>%
+  vip()
+poisson_wf4%>%
+  fit(hosp_data)%>%
+  extract_fit_engine()%>%
+  vip()
 
@@
   geom_smooth(se=FALSE)+
+  geom_abline(intercept=0,slope=1)+
   labs(y="Predicted Hospitalizations",title="Predicted Hospitalizations VS Hospitalizations \nby linear regression with main predictor",color="Fold")
@@
   geom_smooth(se=FALSE)+
+  geom_abline(intercept=0,slope=1)+
   labs(y="Predicted Hospitalizations",title="Predicted Hospitalizations VS Hospitalizations \nby linear regression with all predictors",color="Fold")
@@
 rf_spec=rand_forest(mtry=tune(),min_n=tune(),trees=1000)%>%
-  set_engine("ranger")%>%
+  set_engine("ranger",importance = 'impurity')%>%
   set_mode("regression")
@@
   fit(illness_data)
+# Plotting predicted illness VS Illnesses
 rf_predict1=rf_fit1%>%

### 2023-04-07 code/analysis_code/exploratory_analysis.qmd
update

@@
 
-## Analysis Data
+## Testing/Analysis Data
 ```{r}
@@
          count() %>% 
-         mutate(true_hospitalization = n * Hospitalizations), aes(x = Year, y = true_hospitalization, group = Pathogen_Type, color = Pathogen_Type)) + geom_point() + 
-  geom_jitter() + 
-  labs(x = "Year", y = "Hospitalizations", title = "Scatter Plot of Hospitalizations by Year and Pathogen Type")
+         mutate(true_hospitalization = n * Hospitalizations), 
+       aes(x = Year, y = true_hospitalization, 
+           group = Pathogen_Type, color = Pathogen_Type)) + 
+  geom_point() + geom_jitter() + 
+  labs(x = "Year", y = "Hospitalizations", 
+       title = "Scatter Plot of Hospitalizations by Year and Pathogen Type")
 
@@
          group_by(Pathogen_Type, Simplified_Etiology, Hospitalizations) %>% count() %>% 
-         mutate(true_hospitalization = n * Hospitalizations), aes(x = Pathogen_Type, y = true_hospitalization, group = Simplified_Etiology, fill = Simplified_Etiology)) +
+         mutate(true_hospitalization = n * Hospitalizations), 
+       aes(x = Pathogen_Type, y = true_hospitalization, 
+           group = Simplified_Etiology, fill = Simplified_Etiology)) +
   geom_col() + 
-  labs(x = "Pathogen Type", y = "Hospitalizations", title = "Stacked Bar Chart of Hospitalizations by Pathogen Type and Etiology")
+  labs(x = "Pathogen Type", y = "Hospitalizations", 
+       title = "Stacked Bar Chart of Hospitalizations by Pathogen Type and Etiology")
 
@@
          select(Pathogen_Type, Simplified_Etiology, Hospitalizations) %>% filter(Pathogen_Type == "Bacteria"), aes(x = Simplified_Etiology)) + geom_bar() + 
-  labs(x = "Bacteria", y = "Hospitalizations", title = "Bar Chart of Bacteria Frequency")+
+  labs(x = "Bacteria", y = "Hospitalizations", 
+       title = "Bar Chart of Bacteria Frequency") +
   coord_flip()
@@
 
-####
+### Even Closer Look at Salmonella Cases
 
-Norovirus_Hosp=training_data_final%>%
-  select(Simplified_Etiology, Hospitalizations, `IFSAC Category`)%>%
-  group_by(Simplified_Etiology, Hospitalizations, `IFSAC Category`)%>%
-  filter(Simplified_Etiology=="Salmonella")%>%
-  filter(`IFSAC Category` != "Multiple")%>%
+Salmonella_Hosp <- training_data_final %>%
+  select(Simplified_Etiology, Hospitalizations, `IFSAC Category`) %>%
+  group_by(Simplified_Etiology, Hospitalizations, `IFSAC Category`) %>%
+  filter(Simplified_Etiology == "Salmonella") %>%
+  filter(`IFSAC Category` != "Multiple") %>%
   filter(`IFSAC Category` != "Other") %>%
-  count()%>%
-  mutate(total_hospitalizations=n*Hospitalizations)
+  count() %>%
+  mutate(total_hospitalizations = n * Hospitalizations) %>%
+  arrange(desc(total_hospitalizations))
 
-ggplot(Norovirus_Hosp,aes(x=`IFSAC Category`,y=total_hospitalizations))+
-  geom_boxplot()+
+ggplot(Salmonella_Hosp, aes(x=`IFSAC Category`, y = total_hospitalizations))+
+  geom_boxplot() +
   coord_flip()
@@
   group_by(State) %>% 
-  mutate(all_hospitalizations = sum(Hospitalizations)) %>%     count(all_hospitalizations) %>% 
+  mutate(all_hospitalizations = sum(Hospitalizations)) %>% 
+  count(all_hospitalizations) %>% 
   select(State, all_hospitalizations) %>% arrange(desc(all_hospitalizations))
@@
 
-## Project total hospitalizations in states map
-training_hosp=training_data_final%>%
-  mutate(state=tolower(State))%>%
+## Project Total Hospitalizations onto State Map
+training_hosp <- training_data_final%>%
+  mutate(state = tolower(State))%>%
   group_by(state)%>%
@@
 
-state_map=training_hosp%>%
+state_map <- training_hosp %>%
   distinct(state)%>%
-  inner_join(.,map_data("state"),by=c("state"="region"))
+  inner_join(., map_data("state"), by = c("state"="region"))
+
 state_map
 
-map_final=right_join(training_hosp,state_map,by=c("state"="state"))
+map_final <- right_join(training_hosp,state_map,by=c("state"="state"))
 
-map_final%>%
-  ggplot(aes(x=long,y=lat,fill=all_hosp,group=group))+
-  geom_polygon(color="black",size=0.1)+
-  geom_polygon(data=state_map,aes(x=long,y=lat,group=group),fill=NA,color="black",size=0.3)+
-  scale_fill_gradient(name="Number of \n Hospitalizations",
-                      low="#FFFFFF",high="#FF0000",
-                       limits=c(0,NA))+
-  labs(title="Cases of Hospitalizations Caused by Foodborne Illnesses\n in Each State During 1998-2015")+
-  theme_void()+
+map_final %>%
+  ggplot(aes(x=long, y=lat, 
+             fill = all_hosp, group = group)) +
+  geom_polygon(color="black", size = 0.1) +
+  geom_polygon(data = state_map, aes(x= long, y= lat, group = group), 
+               fill = NA, color = "black", size= 0.3) +
+  scale_fill_gradient(name = "Number of \n Hospitalizations",
+                      low = "#FFFFFF", high="#FF0000",
+                       limits = c(0,NA)) +
+  labs(title = "Cases of Hospitalizations Caused by Foodborne Illnesses\n in Each State During 1998-2015") +
+  theme_void() +
   coord_quickmap()
 
-ggsave("State_hospitalization_map.png",path=image_path,width=3000,height=2000,units="px")
+ggsave("State_hospitalization_map.png", path = image_path, width=3000, height=2000, units="px")
 
@@
 ```
+
 All outbreaks with a high hospitalization count (> 200 cases) are caused by bacteria.

### 2023-04-07 code/analysis_code/statistical_analysis.qmd
update

--- 2023-03-16 code/analysis_code/statistical_analysis.qmd
@@
-# Load libraries
+---
+title: "Statistical analysis script"
+author: "Weifan Wu and Kailin Chen"
+date: "03/16/2023"
+output: html_document
+---
+
+# Load Libraries
 ```{r}
@@
 library(poissonreg)
+library(knitr)
+library(skimr)
+library(vip)
+## For Decision Trees
+library(ranger)
+library(caret)
+library(rpart)
+library(ipred)
+library(e1071)
 ```
@@
 ```{r}
-# Path to Data
+# Import Training Data
 data_location <- here(
   "data","processed_data","processed_training_data.rds")
-# Import Data
 training_data_final <- readRDS(data_location)
-glimpse(training_data_final)
+skim(training_data_final)
+
+# Import Test Data
+data_location <- here(
+  "data","processed_data","processed_testing_data.rds")
+test_data_final <- readRDS(data_location)
+skim(test_data_final)
+
 ```
-# First Health Outcome of Interest: Illnesses
+
+```{r}
+# Set Image Path
+images_stats_path <- here("results","images_statistic_analysis")
+
+# Checking General Illness Distribution
+training_data_final %>%
+  ggplot(aes(Illnesses)) +
+  geom_histogram() +
+  scale_x_continuous(trans = scales::pseudo_log_trans(base = 10))
+## Save Histogram
+ggsave("histogram_Illnesses.png", path = images_stats_path, width = 3000, height = 2000, units = "px")
+
+# Checking General Hospitalization Distribution
+training_data_final %>%
+  ggplot(aes(Hospitalizations)) +
+  geom_histogram() +
+  scale_x_continuous(trans=scales::pseudo_log_trans(base = 10))
+## Save Generated Histogram
+ggsave("histogram_Hospitalizations.png", path = images_stats_path)
+```
+
+# Create Tibbles for Illness & Hospitalizations
+```{r}
+process_datasets_to_tibbles <- function(dataset) {
+  dataset %>%
+    filter(!`IFSAC Category` %in% c("Multiple", "Other"), 
+           Location!= "Unknown") %>%
+    mutate_if(is.character, as.factor) %>%
+    select(!Pathogen_Type)
+}
+
+illness_data <- process_datasets_to_tibbles(training_data_final) %>%
+  select(!Month) %>%
+  rename(Month = Numeric_Month) %>%
+  select(!Food) %>% select(!Hospitalizations)
+
+hosp_data <- process_datasets_to_tibbles(training_data_final) %>%
+  select(!Month) %>%
+  rename(Month = Numeric_Month) %>%
+  select(!Food) %>% select(!Illnesses)
+
+skim(illness_data)
+skim(hosp_data)
+
+test_data_final <- process_datasets_to_tibbles(test_data_final) %>%
+  select(!Food)
+
+skim(test_data_final)
+```
+
+The two health outcomes of interest are illnesses and hospitalizations.
 
 Predictors
--   Time (Year)
+-   Time (Month, Year)
 -   Pathogen Type
@@
 
-```{r}
-# Checking Illness Distribution
-training_data_final %>%
-  ggplot(aes(Illnesses))+
-  geom_histogram()+
-  scale_x_continuous(trans=scales::pseudo_log_trans(base=10))
-image_path <- here("results")
-ggsave("histogram_Illnesses.png", path = image_path, width = 3000, height = 2000, units = "px")
-
-# Tibble with Predictors of Interest
-illness_data=training_data_final %>%
-  select(c(1, 4, 5, 7, 10, 13, 14)) %>%
-  mutate_if(is.character, as.factor)
-glimpse(illness_data)
-```
-
-## Modeling Illness with Linear Regression and Poisson Regression
+# Modeling Illness
 
@@
 ```{r}
-# Linear Regression
-lm_spec = linear_reg() %>%
+# Set Seed for Reproducibility
+set.seed(123)
+
+# Cross-Validation Setup
+illness_fold <- vfold_cv(illness_data, repeats = 2, strata = Illnesses)
+
+# Additional Preprocessing to Deal with Variables with > 50 Categories
+preprocessing_recipe <- function(anterior) {
+  anterior %>%
+    step_dummy(Simplified_Etiology) %>%
+    step_dummy(State) %>%
+    step_dummy(`IFSAC Category`)
+}
+
+## Simple Recipe for Illnesses
+illness_simple_recipe <- recipe(Illnesses ~ Simplified_Etiology, 
+                                data = illness_data) %>%
+  step_dummy(Simplified_Etiology)
+
+## Recipe with All Predictors for Illnesses
+illness_complex_recipe <- recipe(Illnesses ~ ., data = illness_data) %>%
+  preprocessing_recipe()
+
+### Lump small proportion into "other" using step_other
+### Create dummy code for nominal variables
+#illness_all_rec=recipe(Illnesses~.,data=illness_data)%>%
+#  step_other(State,Location,`IFSAC Category`,Simplified_Etiology,threshold = 0.01)%>%
+#  step_dummy(all_nominal())
+
+### Check Processed Dataset
+illness_complex_recipe %>% prep() %>% bake(new_data = NULL)
+
+# Model Specification
+## Linear Regression
+linear_model <- linear_reg() %>%
   set_engine(engine = "lm")
+
+# Create Workflow
+## Main Predictor `Simplified_Etiology`
+linear_workflow_simple_illness <- workflow() %>%
+  add_recipe(illness_simple_recipe) %>%
+  add_model(linear_model)
+
+## All Predictors
+linear_workflow_complex_illness <- workflow() %>%
+  add_recipe(illness_complex_recipe)%>%
+  add_model(linear_model)
+
+## Poisson Regression
 ## Modeling with Main Predictor `Simplified_Etiology`
-lm_fit1 = lm_spec %>%
-  fit(Illnesses ~ Simplified_Etiology, data = illness_data)
+poisson_workflow_simple_illness <- workflow() %>%
+  add_recipe(illness_simple_recipe) %>%
+  add_model(poisson_reg())
+
 ## Modeling with All Predictors
-lm_fit2 = lm_spec %>%
-  fit(Illnesses ~ ., data = illness_data)
+poisson_workflow_complex_illness <- workflow() %>%
+  add_recipe(illness_complex_recipe) %>%
+  add_model(poisson_reg())
+```
 
-# Poisson Regression
-## Modeling with Main Predictor `Simplified_Etiology`
-poisson_wf1 = workflow(Illnesses ~ Simplified_Etiology, poisson_reg())
-## Modeling with All Predictors
-poisson_wf2 = workflow(Illnesses ~ ., poisson_reg())
-## Model Fitting
-poisson_fit1 = fit(poisson_wf1, data=illness_data)
-poisson_fit2 = fit(poisson_wf2, data=illness_data)
+## Linear Regression
+### Model Fitting
+```{r}
+set.seed(234)
+
+## Linear Regression
+### Simple Model
+lm_simple_resampled_illness <- linear_workflow_simple_illness %>%
+  fit_resamples(resamples = illness_fold, control = control_resamples(save_pred = TRUE))
+
+### All Predictors
+lm_complex_resampled_illness <- linear_workflow_complex_illness %>%
+  fit_resamples(resamples = illness_fold, control = control_resamples(save_pred = TRUE))
 ```
@@
 ```{r}
-# Checking Parameters for Linear Regression Model
-tidy(lm_fit1) %>%
+# Check Metrics
+## Simple Regression
+lm_simple_resampled_illness %>%
+  collect_metrics()
+## Multivariate Regression
+lm_complex_resampled_illness %>%
+  collect_metrics()
+
+# Check Parameters for Linear Regression Model
+## Simple Regression
+linear_model_simple_fit_illness <- linear_workflow_simple_illness %>%
+  fit(illness_data) %>%
+  tidy() %>%
   arrange(-estimate)
-tidy(lm_fit2) %>%
+## Multivariate Regression
+linear_model_complex_fit_illness <- linear_workflow_complex_illness %>%
+  fit(illness_data) %>%
+  tidy() %>%
   arrange(-estimate)
 
-# Checking Parameters for Poisson Regression Model
-tidy(poisson_fit1) %>%
+# Check Variable Importance
+## Simple Regression
+linear_workflow_simple_illness %>%
+  fit(illness_data) %>%
+  extract_fit_engine() %>%
+  vip()
+## Multivariate Regression
+linear_workflow_complex_illness %>%
+  fit(illness_data) %>%
+  extract_fit_engine() %>%
+  vip()
+
+# Plotting Predicted Values vs True Values (Illnesses)
+lm_simple_resampled_illness %>%
+  collect_predictions() %>%
+  # filter(Illnesses < 300) %>%
+  ggplot(aes(Illnesses, .pred, color = id2, group = id2)) +
+  geom_point() +
+  geom_smooth(se = FALSE) +
+  geom_abline(intercept = 0,slope = 1) +
+  labs(y = "Predicted Illnesses", 
+       title = "Predicted Illnesses VS Illnesses via Simple Linear Regression")
+
+ggsave("Predicted Illnesses VS Illnesses by linear regression with main predictor.png", path = images_stats_path, width = 3000, height = 2000, units = "px")
+
+lm_complex_resampled_illness %>%
+  collect_predictions() %>%
+  # filter(Illnesses < 300) %>%
+  ggplot(aes(Illnesses, .pred, color = id2, group = id2)) +
+  geom_point() +
+  geom_smooth(se = FALSE) +
+  geom_abline(intercept = 0, slope = 1) +
+  labs(y = "Predicted Illnesses", title = "Predicted Illnesses VS Illnesses via Multivariate Linear Regression")
+
+ggsave("Predicted Illnesses VS Illnesses by linear regression with all predictors.png", path=images_stats_path, width = 3000, height = 2000, units = "px")
+
+```
+
+## Poisson Regression
+### Model Fitting
+```{r}
+## Poisson Regression
+### Simple
+poisson_simple_resampled_illness <- poisson_workflow_simple_illness %>%
+  fit_resamples(resamples = illness_fold,
+    control = control_resamples(save_pred = TRUE))
+
+### Multivariate
+poisson_complex_resampled_illness <- poisson_workflow_complex_illness %>%
+  fit_resamples(resamples = illness_fold,
+    control = control_resamples(save_pred = TRUE))
+```
+
+### Model Evaluation
+```{r}
+# Metrics
+## Simple
+poisson_simple_resampled_illness %>%
+  collect_metrics()
+## Complex
+poisson_complex_resampled_illness %>%
+  collect_metrics()
+
+# Check Parameters
+## Simple
+poisson_simple_fit_illness <- poisson_workflow_simple_illness %>%
+  fit(illness_data) %>%
+  tidy() %>%
   arrange(-estimate)
-tidy(poisson_fit2) %>%
+## Complex
+poisson_complex_fit_illness <- poisson_workflow_complex_illness %>%
+  fit(illness_data) %>%
+  tidy() %>%
   arrange(-estimate)
 
-# Model Evaluation
-Result = lm_fit1 %>% 
-  predict(new_data = illness_data) %>% 
-  mutate(.pred = round(.pred), 
-        true_value = illness_data$Illnesses, model= "lm_single") %>% 
-  bind_rows(lm_fit1 %>% 
-              predict(new_data=illness_data) %>% 
-              mutate(.pred=round(.pred), 
-                     true_value = illness_data$Illnesses, model= "lm_multiple") %>% 
-              bind_rows(poisson_fit1 %>% predict(new_data= illness_data) %>%
-                          mutate(true_value=illness_data$Illnesses, 
-                                 model= "glm_single") %>% 
-                          bind_rows(poisson_fit2 %>% 
-                                      predict(new_data= illness_data)%>% 
-                                      mutate(true_value=illness_data$Illnesses, 
-                                             model= "glm_multiple")))
-)
+# Check Variable Importance
+## Simple
+poisson_workflow_simple_illness %>%
+  fit(illness_data) %>%
+  extract_fit_engine() %>%
+  vip()
+## Complex
+poisson_workflow_complex_illness %>%
+  fit(illness_data) %>%
+  extract_fit_engine() %>%
+  vip()
 
-Result <- Result %>%
-  group_by(model) %>%
-  rmse(truth=true_value, estimate = .pred) %>%
-  rename(illness_model = model)
+# Plotting Predictions vs Illnesses
+## Simple
+poisson_simple_resampled_illness %>%
+  collect_predictions() %>%
+  # filter(Illnesses < 300) %>%
+  ggplot(aes(Illnesses, .pred, color = id2, group = id2)) +
+  geom_point() +
+  geom_smooth(se = FALSE) +
+  geom_abline(intercept = 0, slope = 1) +
+  labs(y = "Predicted Illnesses", title = "Predicted Illnesses VS Illnesses by Simple Poisson Regression")
 
-saveRDS(Result, file = here("results", "illness_model_evaluation.rds"))
+ggsave("Predicted Illnesses VS Illnesses by poisson regression with main predictor.png", path = images_stats_path, width = 3000, height = 2000, units = "px")
+
+## Complex
+poisson_complex_resampled_illness %>%
+  collect_predictions() %>%
+  # filter(Illnesses < 500) %>%
+  ggplot(aes(Illnesses, .pred, color = id2, group = id2)) +
+  geom_point() +
+  geom_smooth(se = FALSE) +
+  geom_abline(intercept = 0,slope = 1) +
+  labs(y = "Predicted Illnesses", title = "Predicted Illnesses VS Illnesses by linear regression with all predictors")
+
+ggsave("Predicted Illnesses VS Illnesses by poisson regression with all predictors.png", path = images_stats_path, width = 3000, height = 2000, units = "px")
 
 ```
-According to the result, the model built using Poisson regression with multiple predictors seems to have the smallest root mean squared error.
+According to the result, the model built using multivariate linear regression seems to have the smallest root mean squared error and highest R^2.
+However, none of those models can catch outliers (Illness > 1000).
 
-# Second Health Outcome of Interest: Hospitilization
-
-Predictors
--   Time (Year)
--   Pathogen Type
--   Etiology
--   Location
--   State
--   Food Type (IFSAC Category)
-
-```{r}
-# Checking the Distribution of Hospitalization
-training_data_final %>%
-  ggplot(aes(Hospitalizations)) +
-  geom_histogram() +
-  scale_x_continuous(trans=scales::pseudo_log_trans(base = 10))
-# Tibble for Variables of Interest
-hosp_data = training_data_final %>%
-  select(c(1, 4, 5, 7, 11, 13, 14))%>%
-  mutate_if(is.character, as.factor)
-glimpse(hosp_data)
-ggsave("histogram_Hospitalizations.png", path = image_path)
-```
-
-## Modeling `Hospitalizations` using Linear Regression and Poisson Regression
+# Modeling Hospitilizations
 
@@
 ```{r}
-# Model Specification and Fitting for Linear Regression Model
-lm_spec = linear_reg() %>%
-  set_engine(engine = "lm")
-## Modeling with Main Predictor `Simplified_Etiology`
-lm_fit3 = lm_spec %>%
-  fit(Hospitalizations ~ Simplified_Etiology, data = hosp_data)
-## Modeling with All Predictors
-lm_fit4 = lm_spec %>%
-  fit(Hospitalizations ~ ., data = hosp_data)
+# Cross-Validation
+hosp_fold <- vfold_cv(hosp_data, repeats = 2, strata = Hospitalizations)
 
-# Model Specification and Fitting for Poisson Regression Model
-## Modeling with Main Predictor `Simplified_Etiology`
-poisson_wf3 = workflow(Hospitalizations ~ Simplified_Etiology, poisson_reg())
-## Modeling with all predictors
-poisson_wf4=workflow(Hospitalizations~., poisson_reg())
-## Fitting models
-poisson_fit3 = fit(poisson_wf3,data = hosp_data)
-poisson_fit4 = fit(poisson_wf4,data = hosp_data)
+# Data Preprocessing
+## Simple Regression Recipe
+hosp_simple_recipe <- recipe(Hospitalizations ~ Simplified_Etiology, data = hosp_data) %>%
+  step_dummy(Simplified_Etiology)
+
+## Multivariate Regression Recipe
+hosp_complex_recipe <- recipe(Hospitalizations ~ ., data = hosp_data) %>%
+  preprocessing_recipe()
+
+## Recipe for all predictors
+### Lump small proportion into "other" using step_other
+### Create dummy code for nominal variables
+# hosp_all_rec=recipe(Hospitalizations ~.,data=hosp_data)%>%
+#  step_other(State,Location,`IFSAC Category`, Simplified_Etiology,threshold = 0.01)%>%
+#  step_dummy(all_nominal_predictors())
+
+# Check Preprocessed Dataset
+hosp_complex_recipe %>%
+  prep() %>%
+  bake(new_data = NULL)
+
+# Create Workflows
+## Simple
+linear_workflow_simple_hosp <- workflow() %>%
+  add_recipe(hosp_simple_recipe) %>%
+  add_model(linear_model)
+poisson_workflow_simple_hosp <- workflow() %>%
+  add_recipe(hosp_simple_recipe) %>%
+  add_model(poisson_reg())
+## Multivariate
+linear_workflow_complex_hosp <- workflow() %>%
+  add_recipe(hosp_complex_recipe) %>%
+  add_model(linear_model)
+poisson_workflow_complex_hosp <- workflow() %>%
+  add_recipe(hosp_complex_recipe) %>%
+  add_model(poisson_reg())
+```
+
+## Linear Regression
+```{r}
+set.seed(234)
+```
+
+### Model Fitting
+```{r}
+# Simple
+lm_simple_resampled_hosp <- linear_workflow_simple_hosp %>%
+  fit_resamples(resamples = hosp_fold,
+    control = control_resamples(save_pred = TRUE))
+
+# Complex
+lm_complex_resampled_hosp <- linear_workflow_complex_hosp %>%
+  fit_resamples(resamples = hosp_fold,
+    control = control_resamples(save_pred = TRUE))
 ```
@@
 ```{r}
-Result = lm_fit3 %>%
- predict(new_data = hosp_data) %>%
- mutate(.pred = round(.pred), true_value = hosp_data$Hospitalizations,
-        model= "lm_single") %>% 
-  bind_rows(lm_fit4 %>% 
-              predict(new_data=hosp_data) %>% 
-              mutate (.pred=round(.pred), 
-                      true_value=hosp_data$Hospitalizations, 
-                      model= "lm_multiple") %>% 
-              bind_rows(poisson_fit1 %>% 
-                        predict(new_data = hosp_data) %>%
-                        mutate(true_value = hosp_data$Hospitalizations, 
-                               model= "glm_single") %>% 
-                          bind_rows(poisson_fit2 %>% 
-                                      predict(new_data= hosp_data) %>% 
-                                      mutate(true_value = hosp_data$Hospitalizations, 
-                                             model= "glm_multiple")))
+# Metrics
+lm_simple_resampled_hosp %>%
+  collect_metrics()
+lm_complex_resampled_hosp %>%
+  collect_metrics()
+
+# Check Parameters
+linear_workflow_simple_hosp %>%
+  fit(hosp_data) %>%
+  tidy() %>%
+  arrange(-estimate)
+linear_workflow_complex_hosp %>%
+  fit(hosp_data) %>%
+  tidy() %>%
+  arrange(-estimate)
+
+## Visualizing Variable Importance
+linear_workflow_simple_hosp %>%
+  fit(hosp_data) %>%
+  extract_fit_engine() %>%
+  vip()
+linear_workflow_complex_hosp %>%
+  fit(hosp_data) %>%
+  extract_fit_engine() %>%
+  vip()
+
+# Plotting predictions vs Hospitalizations
+lm_simple_resampled_hosp %>%
+  collect_predictions() %>%
+  # filter(Hospitalizations < 50) %>%
+  ggplot(aes(Hospitalizations, .pred, color = id2))+
+  geom_point() +
+  geom_smooth(se = FALSE) +
+  geom_abline(intercept = 0, slope = 1) +
+  labs(y = "Predicted Hospitalizations", 
+       title = "Predicted Hospitalizations VS Hospitalizations \by Simple Linear Regression", color = "Fold")
+
+ggsave("Predicted Hospitalizations VS Hospitalizations by linear regression with main predictor.png", path = images_stats_path, width = 3000, height = 2000, units = "px")
+
+lm_complex_resampled_hosp %>%
+  collect_predictions() %>%
+  #filter(Hospitalizations < 50) %>%
+  ggplot(aes(Hospitalizations, .pred, color = id2))+
+  geom_point() +
+  geom_smooth(se = FALSE) +
+  geom_abline(intercept = 0, slope = 1) +
+  labs(y = "Predicted Hospitalizations", 
+       title="Predicted Hospitalizations VS Hospitalizations \by Multivariate Linear Regression", color = "Fold")
+
+ggsave("Predicted Hospitalizations VS Hospitalizations by linear regression with all predictors.png", path = images_stats_path, width = 3000, height = 2000, units = "px")
+```
+## Poisson Regression
+### Model Fitting
+```{r}
+## Poisson Regression
+### Simple
+poisson_simple_resampled_hosp <- poisson_workflow_simple_hosp %>%
+  fit_resamples(resamples = hosp_fold,
+    control = control_resamples(save_pred = TRUE))
+
+### Multivariate
+poisson_complex_resampled_hosp <- poisson_workflow_complex_hosp %>%
+  fit_resamples(resamples = hosp_fold,
+    control = control_resamples(save_pred = TRUE))
+```
+
+### Model Evaluation
+```{r}
+# Metrics
+poisson_simple_resampled_hosp %>%
+  collect_metrics()
+poisson_complex_resampled_hosp %>%
+  collect_metrics()
+
+# Check Parameters
+poisson_workflow_simple_hosp %>%
+  fit(hosp_data) %>%
+  tidy() %>%
+  arrange(-estimate)
+
+poisson_workflow_complex_hosp %>%
+  fit(hosp_data) %>%
+  tidy() %>%
+  arrange(-estimate)
+
+# Variable Importance
+poisson_workflow_simple_hosp %>%
+  fit(hosp_data) %>%
+  extract_fit_engine() %>%
+  vip()
+poisson_workflow_complex_hosp %>%
+  fit(hosp_data) %>%
+  extract_fit_engine() %>%
+  vip()
+
+# Predictions vs Hospitalizations
+poisson_simple_resampled_hosp %>%
+  collect_predictions() %>%
+  # filter(Hospitalizations < 50) %>%
+  ggplot(aes(Hospitalizations, .pred, color = id2)) +
+  geom_point() +
+  geom_smooth(se = FALSE) +
+  geom_abline(intercept = 0,slope = 1) +
+  labs(y = "Predicted Hospitalizations", 
+       title = "Predicted Hospitalizations VS Hospitalizations \by Simple Linear Regression", color = "Fold")
+
+ggsave("Predicted Hospitalizations VS Hospitalizations by poisson regression with main predictor.png", path = images_stats_path, width = 3000, height = 2000, units = "px")
+
+poisson_complex_resampled_hosp %>%
+  collect_predictions() %>%
+  # filter(Hospitalizations < 50) %>%
+  ggplot(aes(Hospitalizations, .pred, color = id2)) +
+  geom_point() +
+  geom_smooth(se = FALSE) +
+  geom_abline(intercept = 0, slope = 1) +
+  labs(y = "Predicted Hospitalizations", title = "Predicted Hospitalizations VS Hospitalizations \by Multivariate Linear Regression", color = "Fold")
+
+ggsave("Predicted Hospitalizations VS Hospitalizations by poisson regression with all predictors.png", path = images_stats_path, width = 3000, height = 2000, units = "px")
+
+```
+According to the results, the model built using multivariate linear regression seems to have the smallest root mean squared error and highest R^2 (14%).
+But again, outliers (Hospitalizations > 100) cannot be predicted well by any of those models.
+
+# Machine Learning Approach
+```{r}
+set.seed(345)
+```
+
+## Random Forest
+### Predicting First Outcome `Illnesses` Using All Predictors
+```{r}
+# Model Specification
+rf_spec <- rand_forest(mtry = tune(), min_n = tune(), trees = 1000) %>%
+  set_engine("ranger", importance = 'impurity') %>%
+  set_mode("regression")
+
+# illness_all_rec <- recipe(Illnesses~.,data=illness_data) %>%
+#  step_other(State,Location,`IFSAC Category`,Simplified_Etiology,threshold = 0.01)%>%
+#  step_dummy(all_nominal())
+
+# Workflow
+random_forest_illness_workflow <- workflow() %>%
+  add_recipe(illness_complex_recipe) %>%
+  add_model(rf_spec)
+
+# Training and Tuning Model
+doParallel::registerDoParallel()
+random_forest_illness_tuned <- random_forest_illness_workflow %>%
+  tune_grid(
+    resamples = illness_fold,
+    grid = 15,
+    control = control_grid(save_pred = TRUE),
+    metrics = metric_set(rmse)
   )
+```
 
-Result <- Result %>%
-  group_by(model)%>%
-  rmse(truth = true_value, estimate = .pred) %>%
-  rename(hospital_model = model)
+### Model Evaluation
+```{r}
+# Plot
+random_forest_illness_tuned %>%
+  autoplot()
 
-saveRDS(Result, file = here("results", "hospital_model_evaluation.rds"))
+# Final Specifications
+random_forest_illness_tuned_best <- random_forest_illness_tuned %>%
+  select_best()
+random_forest_illness_final <- random_forest_illness_workflow %>%
+  finalize_workflow(random_forest_illness_tuned_best)
+random_forest_illness_fit <- random_forest_illness_final %>%
+  fit(illness_data)
+
+# Plotting Predicted Illness VS Illnesses
+random_forest_illness_predict <- random_forest_illness_fit %>%
+  predict(illness_data)
+
+illness_data %>%
+  select(Illnesses) %>%
+  bind_cols(random_forest_illness_predict) %>%
+  ggplot(aes(Illnesses, .pred)) +
+  geom_point() +
+  geom_abline(intercept = 0, slope = 1)+
+  geom_smooth(se = FALSE) +
+  labs(y = "Predicted Illnesses", 
+       title = "Predicted Illnesses VS Illnesses by Random Forest Model")
+ggsave("Predicted illnesses VS illnesses by random forest model.png", path = images_stats_path, width = 3000, height = 2000, units = "px")
+```
+
+### Predicting Second Outcome `Hospitalizations` Using All Predictors
+```{r}
+set.seed(456)
+
+# Creating workflow
+random_forest_hosp_workflow <- workflow() %>%
+  add_recipe(hosp_complex_recipe) %>%
+  add_model(rf_spec)
+# Training and tuning model
+doParallel::registerDoParallel()
+random_forest_hosp_tuned <- random_forest_hosp_workflow %>%
+  tune_grid(
+    resamples = hosp_fold,
+    grid = 15,
+    control = control_grid(save_pred = TRUE),
+    metrics = metric_set(rmse)
+  )
+```
+
+### Model Evaluation
+```{r}
+# Plot
+random_forest_hosp_tuned %>%
+  autoplot()
+
+# Final Specifications
+random_forest_hosp_tuned %>%
+  show_best()
+random_forest_hosp_best <- random_forest_hosp_tuned %>%
+  select_best()
+
+random_forest_hosp_final <- random_forest_hosp_workflow %>%
+  finalize_workflow(random_forest_hosp_best)
+
+random_forest_hosp_fit <- random_forest_hosp_final %>%
+  fit(hosp_data)
+
+random_forest_hosp_predict <- random_forest_hosp_fit %>%
+  predict(hosp_data)
+
+hosp_data %>%
+  select(Hospitalizations) %>%
+  bind_cols(random_forest_hosp_predict) %>%
+  ggplot(aes(Hospitalizations, .pred)) +
+  geom_point() +
+  geom_abline(intercept = 0, slope = 1) +
+  geom_smooth(se = FALSE) +
+  labs(y = "Predicted Hospitalizations",
+       title = "Predicted Hospitalizations VS Hospitalizations by Random Forest")
+
+ggsave("Predicted Hospitalizations VS Hospitalizations by random forest model.png", path = images_stats_path, width = 3000, height = 2000, units = "px")
+```
+
+Overall, random forest model has the lowest rmse among those selected models when training data is fitted; Also, the random forest model outperforms other models in predicting outliers.
+
+## Bagging
+### Illnesses
+```{r}
+bag_training_illness <- bagging(formula = Illnesses ~ .,
+                        data = illness_data,
+                        method = "treebag",
+                        trControl = trainControl(method = "cv", number = 30),
+                        nbagg = 500,
+                        coob = TRUE,
+                        control = rpart.control(minsplit = 2, cp = 0))
+
+bag_training_illness
+```
+
+### Hospitalizations
+```{r}
+bag_training_hosp <- bagging(formula = Hospitalizations ~ .,
+                        data = hosp_data,
+                        method = "treebag",
+                        trControl = trainControl(method = "cv", number = 30),
+                        nbagg = 500,
+                        coob = TRUE,
+                        control = rpart.control(minsplit = 2, cp = 0))
+
+bag_training_hosp
+```
+
+## Organize Model Performance and Choose the Best One
+### Illnesses
+```{r}
+# Linear Regression
+lm_complex_resampled_illness %>%
+  collect_metrics()
+
+# Poisson Regression
+poisson_complex_resampled_illness %>%
+  collect_metrics()
+
+# Random Forest
+random_forest_illness_predict %>%
+  collect_metrics()
+
+# Make Dataframe of Model RMSEs
 
@@
 
-According to the result, the model built using linear regression with multiple predictors seems to have the smallest root mean squared error.
+### Hospitalizations
+```{r}
+# Linear Regression
+# Poisson Regression
+# Random Forest
+
+```
+
+## Fit Testing Data

### 2023-04-07 code/processing_code/processingfile_v1.qmd
update

@@
 # Separate the rows for coninfections
-training_data_fin=training_data_fin%>%
-  separate_rows(Species,sep=";")
+training_data_fin <- training_data_fin %>%
+  separate_rows(Species, sep= ";")
 # Create lists for pathogen types
@@
 # Create two new columns: Pathogen_type & Simplified_Etiology
-training_data_fin=training_data_fin%>%
+training_data_fin <- training_data_fin %>%
   mutate(Pathogen_Type = case_when(
@@
 unique(training_data_fin$Location)
-training_data_fin=training_data_fin %>%
- separate_rows(Location,sep="; ")
+training_data_fin <- training_data_fin %>%
+ separate_rows(Location,sep = "; ")
+
+# Simplify Location
+training_data_fin <- training_data_fin %>% mutate(Location = case_when(
+  grepl("Buffet", Location, ignore.case = TRUE) ~ "Restaurant",
+  grepl("Nursing Home", Location, ignore.case = TRUE) ~ "Nursing Home/Assisted Living Facility",
+  grepl("Fast Food", Location, ignore.case = TRUE) ~ "Restaurant",
+  TRUE ~ Location))
 ```
@@
 ```{r}
-# unique(analysis_data_fin$`IFSAC Category`)
+# unique(testing_data_fin$`IFSAC Category`)
 # unique(training_data_fin$Food)
-Desserts=c("Cookie","Cream","Canoli","Cake","Pudding")
-Beef=c("Beef","Steak")
-Poultry=c("Chicken","Turkey")
-Pork=c("Pork","Sausage","Hot Dog","Deli","Spam","Strawberries","Blueberries","Pig","Ribs")
-Dairy=c("Milk","Cheese","Parsley")
-Crusteaceans=c("Lobster","Crab","Shrimp")
-Fish=c("Fish","Tuna","Salmon")
-Mollusks=c("Clams","Oysters")
-Fruits=c("Tomato","Banana","Melon","Grapes","Mango")
-`Grains-Beans`=c("Rice","Peas","Beans","wheat")
-`Vegetable-Row-Crops`=c("Lettuce","artichokes","broccoli","asparagus","celeries","spinach")
-`Root-underground-vegetables`=c("Roots","carrots","Beets","Onion","Tubers","yams","Garlic","Potatos","taro")
-Sprouts=c("alfalfa","mung")
-Salad=c("Salad")
+
+# Dairy
+Desserts <- c("Cookie","Cream","Canoli","Cake","Pudding")
+Dairy <- c("Milk","Cheese","Parsley")
+
+# Meat & Protein
+Beef <- c("Beef","Steak")
+Poultry <- c("Chicken","Turkey")
+Pork <- c("Pork", 
+          "Sausage",
+          "Hot Dog", 
+          "Deli", 
+          "Spam", 
+          "Strawberries", 
+          "Blueberries", 
+          "Pig", 
+          "Ribs")
+Crusteaceans <- c("Lobster","Crab","Shrimp")
+Fish <- c("Fish","Tuna","Salmon")
+Mollusks <- c("Clams","Oysters")
+Game <- c("Bear")
+`other meat` <- c("Lamb")
+
+# Plant-Based Foods
+Fruits <- c("Tomato","Banana","Melon","Grapes","Mango")
+`Grains-Beans` <- c("Rice","Peas","Beans","wheat")
+`Vegetable-Row-Crops` <- c("Lettuce", "artichokes", "broccoli", "asparagus", 
+                           "celeries", "spinach")
+`Root-underground-vegetables` <- c("Roots", "carrots", "Beets", "Onion", 
+                                   "Tubers", "yams", "Garlic", "Potatos", 
+                                   "taro")
+Sprouts <- c("alfalfa","mung")
+Salad <- c("Salad")
+
 # Other=c("Ice","Alcohol","Eggnog")
-Game=c("Bear")
-Herbs=c("basil","cilantro")
-Fungi=c("Mushroom","portabellas")
-`Nuts-Seeds`=c("peanuts","almonds","Seasames")
-Eggs=c("Egg")
-`Seeded-Vegetables`=c("Peppers","Squashes","Cucumbers","Okras","Corn")
-`other meat`=c("Lamb")
+
+Herbs <- c("basil","cilantro")
+Fungi <- c("Mushroom","portabellas")
+`Nuts-Seeds` <- c("peanuts","almonds","Seasames")
+Eggs <- c("Egg")
+`Seeded-Vegetables` <- c("Peppers","Squashes","Cucumbers","Okras","Corn")
 
@@
   mutate(`IFSAC Category` = case_when(
-    grepl("Salad",Food, ignore.case = TRUE) ~ "Salad",
+    grepl("Salad", Food, ignore.case = TRUE) ~ "Salad",
     grepl(paste(Desserts, 
@@
                 collapse = "|"), Food, ignore.case = TRUE) ~ "Seeded-Vegetables",
-    grepl("eggs",Food,ignore.case = TRUE)~"Eggs",
-    grepl("Lamb",Food)~"Other Meat",
+    grepl("eggs", Food, ignore.case = TRUE)~"Eggs",
+    grepl("Lamb", Food) ~ "Other Meat",
     TRUE ~ if_else(is.na(Food),"","Other")
@@
   relocate(`IFSAC Category`, .after = Food)
-training_data_fin=training_data_fin%>%
-  mutate(`IFSAC Category`=case_when(
-    grepl(";",Food)~"Multiple",TRUE~`IFSAC Category`))
+
+training_data_fin <- training_data_fin %>%
+  mutate(`IFSAC Category`= case_when(
+    grepl(";", Food) ~ "Multiple",
+    TRUE ~ `IFSAC Category`))
 ```
@@
 ```{r}
-
 testing_data_fin <- testing_data_fin %>%
@@
     )) %>%
-  mutate(Simplified_Etiology = word(Etiology,1)) %>%
+  mutate(Simplified_Etiology = word(Etiology, 1)) %>%
       mutate(Simplified_Etiology = case_when(
@@
 ```
-###Column for Location
+
+### Column for Location
 ```{r}
+testing_data_fin
+
+# Create Column
 testing_data_fin <- testing_data_fin %>%
-  rename(Location=Setting)%>%
-  mutate(Location=case_when(
-    grepl("Restaurant",Location)~"Restaurant",TRUE~Location))%>%
-    separate_rows(Location,sep="; ")
-testing_data_fin
+  rename(Location = Setting)%>%
+  mutate(Location = case_when(
+    grepl("Restaurant", Location) ~ "Restaurant", 
+    TRUE ~ Location)) %>%
+    separate_rows(Location, sep = "; ")
+
+# Match Training Data Format
+testing_data_fin <- testing_data_fin %>%
+  mutate(Location = case_when(
+    grepl("Private Home/Residence", Location, ignore.case = TRUE) ~ "Private Home/Residence",
+    grepl("Caterer", Location, ignore.case = TRUE) ~ "Catering Service",
+    grepl("Office/Indoor Workspace", Location, ignore.case = TRUE) ~ "Office/Indoor Workspace",
+    grepl("Banquet", Location, ignore.case = TRUE) ~ "Banquet Facility",
+    grepl("Fair", Location, ignore.case = TRUE) ~ "Fair/Festival",
+    grepl("Religious", Location, ignore.case = TRUE) ~ "Religious Facility",
+    grepl("Grocery Store", Location, ignore.case = TRUE) ~ "Grocery Store",
+    grepl("School/College/University", Location, ignore.case = TRUE) ~ "School/College/University",
+    grepl("Prison/Jail", Location, ignore.case = TRUE) ~ "Prison/Jail",
+    grepl("Long-Term Care", Location, ignore.case = TRUE) ~ "Nursing Home/Assisted Living Facility",
+    grepl("Child", Location, ignore.case = TRUE) ~ "Child Daycare",
+    grepl("Farm/Dairy", Location, ignore.case = TRUE) ~ "Farm/Dairy",
+    grepl("Hotel/Motel", Location, ignore.case = TRUE) ~ "Hotel/Motel",
+    TRUE ~ Location))
+```
+
+## Ensure Training and Test Data Have Same Columns
+
+To ensure compatibility for evaluating model or machine learning performance between the two datasets, column names must be the same.
+
+```{r}
+colnames(training_data_fin)
+colnames(testing_data_fin)
+
+training_data_fin <- training_data_fin %>%
+  select(Year, State, Month, Numeric_Month, State, Location, 
+         Food, `IFSAC Category`, 
+         Illnesses, Hospitalizations, Pathogen_Type, Simplified_Etiology)
+
+testing_data_fin <- testing_data_fin %>%
+  rename(Food = `Food Vehicle`) %>%
+  select(Year, State, Month, State, Location, 
+         Food, `IFSAC Category`, 
+         Illnesses, Hospitalizations, Pathogen_Type, Simplified_Etiology)
+
+colnames(training_data_fin)
+colnames(testing_data_fin)
 ```
@@
 ```{r}
-training_data_final=training_data_fin%>%
+training_data_final <- training_data_fin %>%
+  mutate(Simplified_Etiology = na_if(Simplified_Etiology, "")) %>%
   drop_na()
-testing_data_final=testing_data_fin%>%
+testing_data_final <- testing_data_fin %>%
+  mutate(Simplified_Etiology = na_if(Simplified_Etiology, "")) %>%
   drop_na()
-
 ```

### 2023-04-07 code/analysis_code/statistical_analysis.qmd
ml part completed

@@
 
-bag_training_illness
+# Get RMSE
+bag_training_illness$err
 ```
@@
 
-bag_training_hosp
+# Get RMSE
+bag_training_hosp$err
 ```
@@
 ```{r}
-# Linear Regression
-lm_complex_resampled_illness %>%
-  collect_metrics()
+# Make Dataframe of Model RMSEs
+illness_model_dataframe <- lm_complex_resampled_illness %>%
+  collect_metrics() %>% 
+  filter(.metric == "rmse") %>% select(mean) %>% rename(RMSE = mean) %>% 
+  mutate(Model = "Linear Regression") %>% 
+  full_join(poisson_complex_resampled_illness %>% 
+              collect_metrics() %>% 
+              filter(.metric == "rmse") %>% select(mean) %>% 
+              rename(RMSE = mean) %>% mutate(Model = "Poisson Regression")) %>%
+  full_join(random_forest_illness_tuned %>% 
+              show_best("rmse") %>% top_n(-1, mean) %>% select(mean) %>%
+              rename(RMSE = mean) %>% mutate(Model = "Random Forest")
+            ) %>%
+  full_join(data.frame(bag_training_illness$err, "Bagging") %>% 
+  rename(RMSE = bag_training_illness.err) %>% rename(Model = X.Bagging.)) %>%
+  arrange(RMSE)
 
-# Poisson Regression
-poisson_complex_resampled_illness %>%
-  collect_metrics()
-
-# Random Forest
-random_forest_illness_predict %>%
-  collect_metrics()
-
-# Make Dataframe of Model RMSEs
+illness_model_dataframe
 
@@
 ```{r}
-# Linear Regression
-# Poisson Regression
-# Random Forest
+hosp_model_dataframe <- lm_complex_resampled_hosp %>%
+  collect_metrics() %>% 
+  filter(.metric == "rmse") %>% select(mean) %>% rename(RMSE = mean) %>% 
+  mutate(Model = "Linear Regression") %>% 
+  full_join(poisson_complex_resampled_hosp %>% 
+              collect_metrics() %>% 
+              filter(.metric == "rmse") %>% select(mean) %>% 
+              rename(RMSE = mean) %>% mutate(Model = "Poisson Regression")) %>%
+  full_join(random_forest_hosp_tuned %>% 
+              show_best("rmse") %>% top_n(-1, mean) %>% select(mean) %>%
+              rename(RMSE = mean) %>% mutate(Model = "Random Forest")
+            ) %>%
+  full_join(data.frame(bag_training_hosp$err, "Bagging") %>% 
+  rename(RMSE = bag_training_hosp.err) %>% rename(Model = X.Bagging.)) %>%
+  arrange(RMSE)
+
+hosp_model_dataframe
 
@@
 
-## Fit Testing Data
+Given the above results, the random forest machine learning model is the best model for both health outcomes.
+
+# Modifying Test Data Variables to Maintain Compatibility
+```{r}
+test_data_final
+```
+
+# Fit Testing Data
+```{r}
+# Illness
+placeholder_illnesses <- random_forest_illness_final %>%
+  fit(test_data_final) %>% predict(test_data_final)
+placeholder_illnesses <- placeholder_illnesses[1:1922, ]
+
+test_data_final %>% select(Illnesses) %>% 
+  bind_cols(placeholder_illnesses) %>% rmse(Illnesses, .pred)
+
+# Hospitalizations
+placeholder_hosp <- random_forest_hosp_final %>%
+  fit(test_data_final) %>% predict(test_data_final)
+placeholder_hosp <- placeholder_hosp[1:1922, ]
+
+test_data_final %>% select(Hospitalizations) %>% 
+  bind_cols(placeholder_hosp) %>% rmse(Hospitalizations, .pred)
+
+```

### 2023-04-08 code/analysis_code/statistical_analysis.qmd
update again

@@
 
-# Modifying Test Data Variables to Maintain Compatibility
-```{r}
-test_data_final
-```
-
 # Fit Testing Data
@@
 ```
+
+# Save RDS Files
+
+```{r}
+# Illness
+save_location <- here("results", "illness_rmse_models.rds")
+saveRDS(illness_model_dataframe, file = save_location)
+
+# Hospitalizations
+save_location <- here("results", "hospital_rmse_models.rds")
+saveRDS(hosp_model_dataframe, file = save_location)
+```

### 2023-04-08 products/manuscript/Manuscript2.qmd
update again

---
title: "Manuscript for Foodborne-Outbreak Project"
author: "Weifan Wu & Kailin (Kai) Chen"
date: "May 2023"
format:
  docx
bibliography: ../references.bib
csl: ../apa.csl
---

```{r echo = FALSE, warning = FALSE, paged.print = TRUE}
# Load Libraries
library(here)
library(knitr)
library(tidyverse)
library(broom)
library(ggtext)
```

# Summary/Abstract

# Introduction

## General Background Information

Foodborne pathogens typically cause serious gastrointestinal infections. On average, 1 in 6 Americans will acquire foodborne illness each year; 128,000 will be hospitalized, and 3,000 will die [@centerforfoodsafetyandappliednutrition2022]. Economic loss associated with foodborne illness is estimated to be greater than 50 billion US dollars annually. Restaurants are frequently sites attributed to mass-outbreaks, and typical foods associated with these outbreaks include improperly-prepared meats and vegetables [@mun2020; @sarno2021]. Disease etiologies range from bacteria (ex: *Salmonella* spp. and *E. coli*) to viruses (norovirus and rotavirus), to parasites (*Cryptosporidium* spp. and *Cyclospora* spp. [@cdc2022]. To verify pathogens of greatest concern, common settings associated with foodborne illness, and foods typically associated with such outbreaks, this study was launched as a research project at the University of Georgia using information gathered over a 17-year period from the Centers for Disease Control and Prevention (CDC).

### Questions & Hypotheses

From this research project, we would like to answer the following questions: - What are the most frequent locations that foodborne outbreaks occur? - What pathogen species are most frequently associated with foodborne outbreaks? - What type of food is most often seen for foodborne outbreaks?

In other words, immediate outcomes of interest are location, disease etiology, and food category. These will act as predictors for the health outcomes of interest: illnesses and hospitalizations.

Our hypotheses include: - Outbreaks occur most often in homes rather than restaurants. - The most common pathogen associated with foodborne disease outbreaks is *E. coli*. - The food most commonly associated with foodborne outbreaks consists of salad.

# Methods

## Data Sources, Acquisition, and Description

Two datasets were adopted for this research. The first dataset--which was used to build statistical models for the study--was uploaded onto *Kaggle* in 2017, with 12 variables and more than 16,950 observations [@cdc2017a]. The second dataset--which was used for machine learning--was retrieved from the CDC's National Outbreak Reporting System (NORS) website. This second dataset covers reports of foodborne and waterborne disease outbreaks, as well as enteric disease outbreaks spread by contact with environmental sources, infected animals or persons, and other means from 2009 to 2020.

## Data Importation and Cleaning

The datasets were downloaded from their respective sites and imported into R/RStudio using the `tidyverse` and `readxl` packages. Afterwards, the data structure for each file was analyzed, including how many "NA" values were present for each column in the dataset. Only observations pertaining to foodborne disease were kept. Variables containing less than 80% "NA" values, including predictors and outcomes of interest, were retained. For the designated "training data", cases that did not have laboratory-confirmed tests were filtered out. Variables for pathogen type (virus, bacteria, parasite, and poison) and `Simplified Etiology` were created based on pathogen species. For both datasets, `Location` was formatted identically to maintain compatibility for eventual machine learning. `IFSAC Category` was also created in the training data based on the formatting of the "test data" and modified to include salads, desserts, and poultry. To address coinfections, observations containing multiple pathogen etiologies were separated into different rows. Finally, both datasets were modified to contain identical columns prior to being saved as "processed data" in the appropriate subdirectory.

For more information, please see the "processingfile_v1.qmd" file located in the "processing_code" folder of the "code" subdirectory.

## Statistical Analysis

### Exploratory

After processing procedures, an exploratory data analysis was performed to view general relationships between predictors (time, pathogen type, pathogen etiology, state, and food type) and the health outcomes of interest (illnesses and hospitalizations).

### Machine Learning

Linear and Poisson regression models were afterwards constructed using the aforementioned predictors and outcomes from the training data. Data was resampled using the bootstrapping approach 1,000 times for better estimates of accuracy and variability of model coefficients. Random forest and bagging models were also trained to predict health outcomes using the Kaggle dataset for the NORS-CDC data. RMSE was used to compare each of the models. The best model built from the training data was then employed for the test data.

For more information, please refer to the "statistical_analysis.qmd" file inside the "analysis_code" folder of the "code" subdirectory.

# Results

## Exploratory Analysis

### Training Data

![Figure 2: Hospitalizations by Food Type.](images/Food_type_v_hospitalization-01.png)

```{r Visualization of Hospitalizations by State}
rmarkdown::paged_table(readRDS("../../results/state_v_hospitalization_training.rds"))
```

### Test Data

![Figure 1. Year vs Illnesses Caused by Pathogen Type](images/time_v_illness_pathogen_type.png)

![Figure 3. Illnesses by outbreak location.](images/Location_v_Illness.png)

![Figure 4. Bacterial etiology by outbreak frequency.](images/bacterial_etiology_v_outbreak_frequency.png)

## Statistical Analysis

### Basic

Between simple and multivariate regressions, multivariate models performed better with lower associated RMSEs. For both health outcomes, the multivariate linear regression models (RMSE = 58.4, 5.50) constructed performed better than multivariate poisson regression models (RMSE = 59.7, 5.64).

![Figure 5. Predicted illnesses versus actual illness count for multivariate linear regression model.](images/Predicted%20Illnesses%20VS%20Illnesses%20by%20linear%20regression%20with%20all%20predictors.png){fig-align="center"}

### Machine Learning

```{r Model RMSE Comparison, paged.print=TRUE} 
# Illness Models 
rmarkdown::paged_table(readRDS("../../results/illness_rmse_models.rds"))

# Hospitalization Models
rmarkdown::paged_table(readRDS("../../results/hospital_rmse_models.rds"))
```

Overall, between itself and the other models (linear regression, poisson regression, and bagging) generated, the random forest model performed the best for both health outcomes with the lowest RMSEs (55.7, 5.05).

### Final Fit with Test Data

Fitting the finalized random forest model to the test data resulted in RMSEs of 36.6 and 6.70 for the health outcomes of illnesses and hospitalizations, respectively.

# Discussion

## Summary and Interpretations

## Strengths and Limitations

# Conclusion

# References

### 2023-04-21 code/analysis_code/exploratory_analysis.qmd
update

@@
 
+### Pathogen Type + Etiology vs Hospitalization
 ```{r fig.height=8,fig.width=8}
-# Time vs Hospitalizations
+# Pathogen Type + Etiology vs Hospitalization
+ggplot(training_data_final %>% 
+         group_by(Pathogen_Type, Simplified_Etiology, Hospitalizations) %>% 
+         count() %>% mutate(true_hospitalization = n * Hospitalizations), 
+       aes(x = Pathogen_Type, y = true_hospitalization, 
+           group = Simplified_Etiology, fill = Simplified_Etiology)) +
+  geom_col() + 
+  labs(x = "Pathogen Type", y = "Hospitalizations", 
+       title = "Stacked Bar Chart of Hospitalizations by Pathogen & Etiology") +
+  theme_classic() + 
+  theme(plot.title = element_text(face = "bold", size = 11, hjust = 0.5))
+
+## Save Graph
+image_path <- here("results","images_EDA")
+ggsave("pathogen_type_etiology_v_hospitalizations.png", path = image_path, width = 3000, height = 2000, units = "px")
+```
+
+### Template Setup
+```{r}
+# Template Graph
+theme_set(theme_bw())
+theme_update(plot.title = element_text(face = "bold", size = 12))
+theme_update(plot.title = element_text(hjust = 0.5))
+```
+
+### Time vs Hospitalizations
+```{r}
+# Data Visualization
+## Illnesses
+# ggplot(training_data_final %>%
+#          group_by(Year, Pathogen_Type, Illnesses) %>%
+#          count() %>%
+#          mutate(true_illnesses = n * Illnesses),
+#        aes(x = Year, y = true_illnesses,
+#            group = Pathogen_Type, color = Pathogen_Type)) +
+#   geom_point() + geom_jitter() +
+#   labs(x = "Year", y = "Illnesses",
+#        title = "Scatter Plot of Illnesses by Year and Pathogen Type",
+#        color = "Pathogen Type") +
+#   facet_wrap(~ Pathogen_Type) +
+#   theme(legend.position = "none")
+
+## Hospitalizations
 ggplot(training_data_final %>% 
@@
   labs(x = "Year", y = "Hospitalizations", 
-       title = "Scatter Plot of Hospitalizations by Year and Pathogen Type")
+       title = "Scatter Plot of Hospitalizations by Year and Pathogen Type",
+       color = "Pathogen Type") + 
+  facet_wrap(~ Pathogen_Type) + 
+  theme(legend.position = "none")
 
 ## Save Graph
-image_path <- here("results","images_EDA")
 ggsave("time_v_hospitalization_pathogen_type.png", path = image_path, width = 3000, height = 2000, units = "px")
+```
 
-# Pathogen Type + Etiology vs Hospitalization
+### Bacterial Etiologies for Outbreaks
+```{r}
+# General
 ggplot(training_data_final %>% 
-         group_by(Pathogen_Type, Simplified_Etiology, Hospitalizations) %>% count() %>% 
-         mutate(true_hospitalization = n * Hospitalizations), 
-       aes(x = Pathogen_Type, y = true_hospitalization, 
-           group = Simplified_Etiology, fill = Simplified_Etiology)) +
-  geom_col() + 
-  labs(x = "Pathogen Type", y = "Hospitalizations", 
-       title = "Stacked Bar Chart of Hospitalizations by Pathogen Type and Etiology")
-
-## Save Graph
-ggsave("pathogen_type_etiology_v_hospitalizations.png", path = image_path, width = 3000, height = 2000, units = "px")
-
-## Closer Look at Bacterial Etiologies for Outbreaks
-ggplot(training_data_final %>% 
-         select(Pathogen_Type, Simplified_Etiology, Hospitalizations) %>% filter(Pathogen_Type == "Bacteria"), aes(x = Simplified_Etiology)) + geom_bar() + 
+         select(Pathogen_Type, Simplified_Etiology, Hospitalizations) %>%
+         filter(Pathogen_Type == "Bacteria"), 
+       aes(x = Simplified_Etiology)) + geom_bar() + 
   labs(x = "Bacteria", y = "Hospitalizations", 
-       title = "Bar Chart of Bacteria Frequency") +
-  coord_flip()
+       title = "Bar Chart of Bacteria Frequency") + coord_flip()
 
@@
 
-### Even Closer Look at Salmonella Cases
-
+# Salmonella
 Salmonella_Hosp <- training_data_final %>%
@@
 ggplot(Salmonella_Hosp, aes(x=`IFSAC Category`, y = total_hospitalizations))+
-  geom_boxplot() +
+  geom_boxplot() + labs(x = "Total Hospitalizations", y = "Food Type", 
+                        title = "Hospitalizations by Food Type for Salmonella") +
   coord_flip()
-
+```
+### Hospitalizations by State
+```{r}
 # States vs Hospitalization
-
 state_v_hospitalization_training <- training_data_final %>% 
@@
   select(State, all_hospitalizations) %>% arrange(desc(all_hospitalizations))
+
+training_data_final %>% 
+  group_by(State) %>% 
+  mutate(all_illnesses = sum(Illnesses)) %>% 
+  count(all_illnesses) %>% 
+  select(State, all_illnesses) %>% arrange(desc(all_illnesses))
 
@@
 
-## Project Total Hospitalizations onto State Map
+```
+
+### State Map of Hospitalizations
+```{r}
 training_hosp <- training_data_final%>%
@@
                        limits = c(0,NA)) +
-  labs(title = "Cases of Hospitalizations Caused by Foodborne Illnesses\n in Each State During 1998-2015") +
-  theme_void() +
+  labs(title = "Cases of Hospitalizations Caused by Foodborne Illnesses in Each State (1998-2015)") +
+  theme_void() + theme(plot.title = element_text(hjust = 0.5)) + 
+  theme(plot.title = element_text(face = "bold", size = 12)) +
   coord_quickmap()
 
-ggsave("State_hospitalization_map.png", path = image_path, width=3000, height=2000, units="px")
+ggsave("State_hospitalization_map.png", path = image_path, width = 3000, height = 2000, units="px")
+```
 
+### Food Type and Hospitalizations
+```{r}
 # Food Type and Hospitalizations
@@
          filter(`IFSAC Category` != "Multiple") %>% 
-         filter(`IFSAC Category` != "Other"), 
-       aes(x = `IFSAC Category`, y = Hospitalizations)) + 
+         filter(`IFSAC Category` != "Other") %>%
+         group_by(`IFSAC Category`) %>% 
+         mutate(all_hospitalizations = sum(Hospitalizations)) %>%
+         count(all_hospitalizations) %>% arrange(desc(all_hospitalizations)), 
+       aes(x = `IFSAC Category`, y = all_hospitalizations, 
+           fill = `IFSAC Category`)) + 
   geom_col() + 
-  labs(x = "Food Type", y = "Hospitalizations", title = "Hospitalizations by Food Type, Excluding 'Multiple' and 'Other'")+
-  coord_flip()
+  labs(x = "Food Type", y = "Hospitalizations", title = "Hospitalizations by Food Type, Excluding 'Multiple' and 'Other'") +
+  coord_flip() + theme(legend.position = "none")
 
@@
 
-# Locations and Hospitalizations
+# Extra: Food Type & Illnesses
+training_data_food_and_illnesses <- training_data_final %>% 
+         filter(`IFSAC Category` != "Multiple") %>% 
+         filter(`IFSAC Category` != "Other") %>%
+         group_by(`IFSAC Category`) %>% 
+         mutate(all_illnesses = sum(Illnesses)) %>%
+         count(all_illnesses) %>% 
+  arrange(desc(all_illnesses))
+table_path <- here("results", "training_food_vs_illnesses.rds")
+saveRDS(training_data_food_and_illnesses, file = table_path)
+
+```
+
+### Locations and Hospitalizations
+```{r fig.height=8,fig.width=8}
 Location_v_Hospitalization <- ggplot(training_data_final, aes(x = Location, y = Hospitalizations)) +
   geom_col() + 
-  labs(x = "Locations", y = "Hospitalizations", title = "Hospitalizations by Food Type, Excluding 'Multiple' and 'Other'")+
+  labs(x = "Locations", y = "Hospitalizations", title = "Outbreak Location versus Hospitalizations") +
   coord_flip()
@@
 ## Testing Data
+
+### Pathogen Type vs Illnesses Over Time
 ```{r}
-# Illnesses Caused by Pathogen Type Over Time
 ggplot(analysis_data_final %>% 
@@
 
-## Save Graph
 ggsave("time_v_illness_pathogen_type.png", path = image_path, width = 3000, height = 2000, units = "px")
 
-# Pathogen Type + Etiology vs Illnesses Caused
+```
+
+### Pathogen Type and Etiology vs Illnesses
+```{r}
+# Visualization
 ggplot(analysis_data_final %>% 
@@
 
-## Save Graph
+# Save Graph
 ggsave("pathogen_type_etiology_v_illness.png", path = image_path, width = 3000, height = 2000, units = "px")
+```
 
-## Closer Look at Bacterial Etiologies
+### Bacterial Etiologies
+```{r}
 ggplot(analysis_data_final %>% 
@@
   geom_bar() + 
-  labs(x = "Bacteria", y = "Cases", title = "Bar Chart of Bacteria Frequency")+
+  labs(x = "Bacteria", y = "Cases", title = "Bar Chart of Bacteria Frequency for Illnesses")+
   coord_flip()
@@
 ggsave("bacterial_etiology_v_outbreak_frequency.png", path = image_path, width = 3000, height = 2000, units = "px")
+```
 
-# States vs Illnesses
-
+### Illnesses by State
+```{r}
 state_v_illness_analysis <- analysis_data_final %>% 
@@
 
+# Save Table
+table_path <- here("results", "state_v_illness_training.rds")
 saveRDS(state_v_illness_analysis, file = table_path)
@@
 rmarkdown::paged_table(state_v_illness_analysis)
+```
 
-# Food Type and Illnesses
+### Food Type and Illnesses
+```{r}
 ggplot(analysis_data_final %>% filter(`IFSAC Category` != "Multiple") %>% filter(`IFSAC Category` != "Other"), aes(x = `IFSAC Category`, y = Illnesses)) + 
@@
 
-# Save Graph
 ggsave("food_type_v_illness_frequency.png", path = image_path, width = 3000, height = 2000, units = "px")
 
+```
+
+### Locations and Illnesses
+```{r}
 # Locations and Illnesses
-ggplot(analysis_data_final, aes(x = Location, y = Illnesses)) +
+ggplot(analysis_data_final %>% group_by(Location) %>% 
+         mutate(all_illnesses = sum(Illnesses)) %>%
+         count(all_illnesses) %>% arrange(desc(all_illnesses)), aes(x = Location, y = all_illnesses)) +
   geom_col() + 
-  labs(x = "Locations", y = "Illnesses", title = "Illnesses by Locations")+
+  labs(x = "Locations", y = "Illnesses", title = "Illnesses by Location")+
   coord_flip()
@@
 ggsave("Location_v_Illness.png", path = image_path, width = 3000, height = 2000, units = "px")
-
 ```
@@
 Resturants and Private home are the most frequent locations that foodborne outbreaks occur.
+
 Overall, the relationships between predictors and outcomes of training dataset share similar patterns with those of analysis dataset.

### 2023-04-21 code/processing_code/processingfile_v1.qmd
update

@@
     grepl("Caterer", Location, ignore.case = TRUE) ~ "Catering Service",
-    grepl("Office/Indoor Workspace", Location, ignore.case = TRUE) ~ "Office/Indoor Workspace",
+    grepl("Office/Indoor", Location, ignore.case = TRUE) ~ "Office/Indoor Workplace",
     grepl("Banquet", Location, ignore.case = TRUE) ~ "Banquet Facility",
@@
     grepl("Hotel/Motel", Location, ignore.case = TRUE) ~ "Hotel/Motel",
+    grepl("Ship/Boat", Location, ignore.case = TRUE) ~ "Ship/Boat",
     TRUE ~ Location))
+
+unique(testing_data_fin$Location)
+unique(training_data_fin$Location)
 ```

### 2023-04-21 products/manuscript/Manuscript2.qmd
update

@@
 format:
-  docx
+  html
 bibliography: ../references.bib
@@
 library(ggtext)
+library(flextable)
 ```
@@
 
-Foodborne pathogens typically cause serious gastrointestinal infections.
-On average, 1 in 6 Americans will acquire foodborne illness each year; 128,000 will be hospitalized, and 3,000 will die [@centerforfoodsafetyandappliednutrition2022].
-Economic loss associated with foodborne illness is estimated to be greater than 50 billion US dollars annually.
+Foodborne pathogens can cause serious gastrointestinal infections.
+About 600 million cases of foodborne disease are reported worldwide, with almost 1/10 of these cases caused by contaminated food [@he2023].
+On average within the United States each year, 1 in 6 Americans will acquire foodborne illness; 128,000 will be hospitalized, and 3,000 will die [@centerforfoodsafetyandappliednutrition2022].
+Economic loss associated with food poisoning is estimated to be greater than 50 billion US dollars annually.
 Restaurants are frequently sites attributed to mass-outbreaks, and typical foods associated with these outbreaks include improperly-prepared meats and vegetables [@mun2020; @sarno2021].
@@
 [@cdc2022].
-To verify pathogens of greatest concern, common settings associated with foodborne illness, and foods typically associated with such outbreaks, this study was launched as a research project at the University of Georgia using information gathered over a 17-year period from the Centers for Disease Control and Prevention (CDC).
+Environmental factors such as heavy metals can also cause foodborne illness.
 
@@
 
-From this research project, we would like to answer the following questions: - What are the most frequent locations that foodborne outbreaks occur?
+To verify pathogens of greatest concern, common settings associated with food poisoning, and foods typically associated with such outbreaks, this study was launched as a research project at the University of Georgia using information gathered over a 17-year period from the Centers for Disease Control and Prevention (CDC).
+
+From this research project, we would like to answer the following questions:
+
+-   What are the most frequent locations that foodborne outbreaks occur?
+
 -   What pathogen species are most frequently associated with foodborne outbreaks?
+
 -   What type of food is most often seen for foodborne outbreaks?
@@
 
-Our hypotheses include: - Outbreaks occur most often in homes rather than restaurants.
-- The most common pathogen associated with foodborne disease outbreaks is *E.
-coli*.
+Our hypotheses include:
+
+-   Outbreaks occur most often in homes rather than restaurants.
+
+-   The most common pathogen associated with foodborne disease outbreaks is *E. coli*.
+
 -   The food most commonly associated with foodborne outbreaks consists of salad.
@@
 Two datasets were adopted for this research.
-The first dataset--which was used to build statistical models for the study--was uploaded onto *Kaggle* in 2017, with 12 variables and more than 16,950 observations [@cdc2017a].
+The first dataset--which was used to build statistical models for the study--was uploaded onto *Kaggle* in 2017, with 12 variables and more than 16,950 observations spanning almost two decades from 1998 to 2015 [@cdc2017a].
 The second dataset--which was used for machine learning--was retrieved from the CDC's National Outbreak Reporting System (NORS) website.
@@
 
-![Figure 2: Hospitalizations by Food Type.](images/Food_type_v_hospitalization-01.png)
+```{r Table of Hospitalizations by State}
+first_table <- readRDS("../../results/state_v_hospitalization_training.rds") %>% 
+  rename(Hospitalizations = all_hospitalizations)
+first_table <- flextable(first_table)
+first_table <- theme_vanilla(first_table)
+first_table <- set_caption(first_table, caption = as_paragraph(colorize("Table 1. Hospitalizations by state from 1998 to 2015.", color = "black")))
+first_table <- set_table_properties(first_table, layout = "autofit")
+first_table
+```
 
-```{r Visualization of Hospitalizations by State}
-rmarkdown::paged_table(readRDS("../../results/state_v_hospitalization_training.rds"))
+![Figure 1. Hospitalizations by food type from 1998-2015.](images/Food_type_v_hospitalization-01.png)
+
+```{r}
+second_table <- readRDS("../../results/training_food_vs_illnesses.rds") %>% 
+  select(`IFSAC Category`, all_illnesses) %>% 
+  rename(Illnesses = all_illnesses)
+second_table <- flextable(second_table)
+second_table <- theme_vanilla(second_table)
+second_table <- set_caption(second_table, caption = as_paragraph(colorize("Table 2. Illnesses by food type from 1998 to 2015.", color = "black")))
+second_table
 ```
@@
 
-![Figure 1. Year vs Illnesses Caused by Pathogen Type](images/time_v_illness_pathogen_type.png)
+![Figure 2. Illness over time by pathogen type (2009-2020).](images/time_v_illness_pathogen_type.png)
 
-![Figure 3. Illnesses by outbreak location.](images/Location_v_Illness.png)
+![Figure 3. Illnesses by outbreak location (2009-2020).](images/Location_v_Illness-01.png)
 
-![Figure 4. Bacterial etiology by outbreak frequency.](images/bacterial_etiology_v_outbreak_frequency.png)
+![Figure 4. Bacterial etiology by outbreak frequency (2009-2020).](images/bacterial_etiology_v_outbreak_frequency.png)
 
@@
 
+Exploratory data visualization demonstrates bacteria contribute to the majority outbreaks, and all outbreaks with illnesses numbering over 1,000 and hospitalizations over 100 are attributed to bacteria.
+From all pathogen types, bacteria are thus associated with worse health outcomes when it comes to food poisoning.
+
+From 1998 to 2015, multistate outbreaks caused the most illnesses (N = 18,617) and hospitalizations (3,506) than any in-state outbreak.
+The fact that multistate outbreaks were seen more often than in any individual state for illnesses and hospitalizations suggests that these disease outbreaks typically involve a regional or national ingredient distributor.
+Hence, it would be reasonable to suggest laboratory sampling of goods, especially for bacteria, as a safety benchmark upon arrival at their destinations.
+
+In terms of food categories, poultry was identified to have caused the most hospitalizations (N = 770), followed closely by fruits and beef.
+Interestingly, however, salads were found to cause the most illnesses (N = 11,493).
+Upon addition of total hospitalizations and illnesses for both poultry and salads, salads were determined to have the highest number of cases in total, which confirms our original hypothesis.
+
+The remaining two of the three initial hypotheses were proved incorrect.
+Instead of *E.
+coli*, results overall indicated that *Salmonella* spp.
+are the most common pathogens linked to foodborne disease; these bacteria contribute to the greatest amount of illnesses and hospitalizations in comparison to other disease agents for our data.
+Finally, we also see that food poisoning is most frequently linked to restaurants instead of private households.
+
+\[Something about machine learning here???\]
+
+These results offer insight into common disease characteristics across food poisoning cases, which is greatly useful for epidemiologists and other public health professionals to anticipate which areas to investigate first for future outbreaks.
+
 ## Strengths and Limitations
@@
 
+Food poisoning in the United States poses a great concern to public health in terms of both morbidity and economic costs.
+Previous surveillance data points to *Salmonella* bacteria and salads as the main cause and medium for foodborne disease outbreaks, along with the most common setting for such outbreaks being a restaurant.
+These findings will also give future outbreak investigators a first direction to consider in the field.
+
 # References

### 2023-04-21 code/analysis_code/exploratory_analysis.qmd
more work

@@
 The outcome of interest we will explore for the training data is hospitalizations.
+### Template Setup
+```{r}
+# Template Graph
+theme_set(theme_classic())
+theme_update(plot.title = element_text(face = "bold", size = 12))
+theme_update(plot.title = element_text(hjust = 0.5))
+```
 
 ### Pathogen Type + Etiology vs Hospitalization
-```{r fig.height=8,fig.width=8}
-# Pathogen Type + Etiology vs Hospitalization
+```{r}
+# Pathogen Type vs Hospitalization
 ggplot(training_data_final %>% 
-         group_by(Pathogen_Type, Simplified_Etiology, Hospitalizations) %>% 
+         group_by(Pathogen_Type, Hospitalizations) %>% 
          count() %>% mutate(true_hospitalization = n * Hospitalizations), 
        aes(x = Pathogen_Type, y = true_hospitalization, 
-           group = Simplified_Etiology, fill = Simplified_Etiology)) +
+           )) +
   geom_col() + 
   labs(x = "Pathogen Type", y = "Hospitalizations", 
-       title = "Stacked Bar Chart of Hospitalizations by Pathogen & Etiology") +
+       title = "Bar Chart of Hospitalizations by Pathogen type") +
   theme_classic() + 
@@
 image_path <- here("results","images_EDA")
-ggsave("pathogen_type_etiology_v_hospitalizations.png", path = image_path, width = 3000, height = 2000, units = "px")
-```
-
-### Template Setup
-```{r}
-# Template Graph
-theme_set(theme_bw())
-theme_update(plot.title = element_text(face = "bold", size = 12))
-theme_update(plot.title = element_text(hjust = 0.5))
+ggsave("pathogen_type_v_hospitalizations.png", path = image_path, width = 3000, height = 2000, units = "px")
 ```
@@
 
-## Hospitalizations
+## Hospitalizations VS year
 ggplot(training_data_final %>% 
@@
 ## Save Graph
-ggsave("time_v_hospitalization_pathogen_type.png", path = image_path, width = 3000, height = 2000, units = "px")
+ggsave("year_v_hospitalization.png", path = image_path, width = 3000, height = 2000, units = "px")
+
+## Hospitalizations VS month
+ggplot(training_data_final %>% 
+         group_by(Numeric_Month, Hospitalizations) %>% 
+         count() %>% 
+         mutate(true_hospitalization = n * Hospitalizations), 
+       aes(x = as.factor(Numeric_Month), y = true_hospitalization)) +
+  geom_boxplot() +
+  labs(x = "Month", y = "Hospitalizations", 
+       title = "Scatter Plot of Hospitalizations by Month") + 
+  theme(legend.position = "none")
+
+## Save Graph
+ggsave("month_v_hospitalization.png", path = image_path, width = 3000, height = 2000, units = "px")
 ```
@@
          select(Pathogen_Type, Simplified_Etiology, Hospitalizations) %>%
-         filter(Pathogen_Type == "Bacteria"), 
-       aes(x = Simplified_Etiology)) + geom_bar() + 
+         filter(Pathogen_Type == "Bacteria")%>%
+         group_by(Simplified_Etiology)%>%
+         mutate(all_hospitalizations=sum(Hospitalizations)),
+       aes(x = fct_reorder(Simplified_Etiology,all_hospitalizations))) + geom_bar() + 
   labs(x = "Bacteria", y = "Hospitalizations", 
@@
 
-# Salmonella
-Salmonella_Hosp <- training_data_final %>%
+#  Top 3 Bacterial Etiology: Salmonella, E. coli, and Clostridium
+Top_Hosp <- training_data_final %>%
   select(Simplified_Etiology, Hospitalizations, `IFSAC Category`) %>%
   group_by(Simplified_Etiology, Hospitalizations, `IFSAC Category`) %>%
-  filter(Simplified_Etiology == "Salmonella") %>%
+  filter(Simplified_Etiology %in% c("Salmonella","Clostridium","E. coli")) %>%
   filter(`IFSAC Category` != "Multiple") %>%
@@
   count() %>%
-  mutate(total_hospitalizations = n * Hospitalizations) %>%
-  arrange(desc(total_hospitalizations))
+  mutate(total_hospitalizations = n * Hospitalizations)
 
-ggplot(Salmonella_Hosp, aes(x=`IFSAC Category`, y = total_hospitalizations))+
-  geom_boxplot() + labs(x = "Total Hospitalizations", y = "Food Type", 
-                        title = "Hospitalizations by Food Type for Salmonella") +
+ggplot(Top_Hosp, aes(x=fct_reorder(`IFSAC Category`, total_hospitalizations),y = total_hospitalizations))+
+  geom_boxplot() + labs(x = "Food Type", y = "Total Hospitalizations", 
+                        title = "Hospitalizations by Food Type for Top 3 Bacterial Etiology") +
+  facet_wrap(vars(Simplified_Etiology))+
   coord_flip()
+ggsave("top3_bacterial_etiology_v_food_hospitalization_frequency.png", path = image_path, width = 3000, height = 2000, units = "px")
 ```
+
 ### Hospitalizations by State
@@
          mutate(all_hospitalizations = sum(Hospitalizations)) %>%
-         count(all_hospitalizations) %>% arrange(desc(all_hospitalizations)), 
-       aes(x = `IFSAC Category`, y = all_hospitalizations, 
-           fill = `IFSAC Category`)) + 
+         count(all_hospitalizations), 
+       aes(x = fct_reorder(`IFSAC Category`,all_hospitalizations), y = all_hospitalizations, 
+           )) + 
   geom_col() + 
@@
 ### Locations and Hospitalizations
-```{r fig.height=8,fig.width=8}
-Location_v_Hospitalization <- ggplot(training_data_final, aes(x = Location, y = Hospitalizations)) +
+```{r}
+Location_v_Hospitalization <- training_data_final%>%
+  group_by(Location)%>%
+  mutate(all_hospitalizations = sum(Hospitalizations)) %>%
+         count(all_hospitalizations)%>%
+  ggplot(aes(x = fct_reorder(Location,all_hospitalizations), y = all_hospitalizations)) +
   geom_col() + 
@@
 ggplot(analysis_data_final %>% 
-         group_by(Pathogen_Type, Simplified_Etiology, Illnesses) %>% 
-         filter(Pathogen_Type == "Bacteria"), aes(x = Simplified_Etiology)) + 
+         group_by(Simplified_Etiology) %>% 
+         mutate(Illnesses=sum(Illnesses))%>%
+         filter(Pathogen_Type == "Bacteria"), aes(x = fct_reorder(Simplified_Etiology,Illnesses))) + 
   geom_bar() + 
@@
 ```{r}
-ggplot(analysis_data_final %>% filter(`IFSAC Category` != "Multiple") %>% filter(`IFSAC Category` != "Other"), aes(x = `IFSAC Category`, y = Illnesses)) + 
+analysis_data_final %>% filter(!`IFSAC Category` %in% c("Multiple", "Other"))%>%
+  filter(Illnesses>0)%>%
+  group_by(`IFSAC Category`)%>%
+  mutate(Illnesses=sum(Illnesses))%>%
+ggplot(aes(x = fct_reorder(`IFSAC Category`,Illnesses), y = Illnesses)) + 
   geom_col() + 
+  scale_y_log10()+
   labs(x = "Food Type", y = "Illnesses", title = "Illnesses by Food Type, Excluding 'Multiple' and 'Other'")+
@@
          mutate(all_illnesses = sum(Illnesses)) %>%
-         count(all_illnesses) %>% arrange(desc(all_illnesses)), aes(x = Location, y = all_illnesses)) +
+         count(all_illnesses), aes(x = fct_reorder(Location,all_illnesses), y = all_illnesses)) +
   geom_col() + 

### 2023-04-21 code/analysis_code/statistical_analysis.qmd
more work

@@
   autoplot()
-
 # Final Specifications
@@
   fit(illness_data)
-
 # Plotting Predicted Illness VS Illnesses

### 2023-04-21 products/manuscript/Manuscript2.qmd
more work

@@
 date: "May 2023"
-format:
-  html
+output: word_document
 bibliography: ../references.bib
@@
 
-Foodborne pathogens can cause serious gastrointestinal infections.
-About 600 million cases of foodborne disease are reported worldwide, with almost 1/10 of these cases caused by contaminated food [@he2023].
-On average within the United States each year, 1 in 6 Americans will acquire foodborne illness; 128,000 will be hospitalized, and 3,000 will die [@centerforfoodsafetyandappliednutrition2022].
-Economic loss associated with food poisoning is estimated to be greater than 50 billion US dollars annually.
-Restaurants are frequently sites attributed to mass-outbreaks, and typical foods associated with these outbreaks include improperly-prepared meats and vegetables [@mun2020; @sarno2021].
-Disease etiologies range from bacteria (ex: *Salmonella* spp.
+Foodborne pathogens typically cause serious gastrointestinal infections.
+On average, 1 in 6 Americans acquire foodborne illness each year, leading to 128,000 hospitalizations, and 3,000 mortality [@centerforfoodsafetyandappliednutrition2022].
+Economic loss associated with foodborne illness is estimated to be greater than 50 billion US dollars annually.
+Moreover, it can be particularly challenging to trace the food sources and exposure settings that are associated with these illnesses .
+Nevertheless, foodborne disease outbreak surveillance enables the identification of crucial agents that are commonly involved in outbreaks.
+In the United States, CDC conducts surveillance for foodborne disease outbreaks through the Foodborne Disease Outbreak Surveillance System [@gould_surveillance_2013].
+The primary responsibility for the identification, investigation, and reporting of foodborne disease outbreaks caused by various enteric microorganism, and chemical/toxic agents lies with state, local, and territorial public health departments, in accordance with established scientific protocols and procedures [@hoffman_capacity_2005].
+Previous studies have identified that restaurants are sites frequently attributed to mass-outbreaks, and typical foods associated with these outbreaks include improperly-prepared meats and vegetables [@mun2020;@sarno2021].
+The etiologies leading to disease encompass a wide range of microorganisms, ranging from bacteria such as *Salmonella* spp.
 and *E.
-coli*) to viruses (norovirus and rotavirus), to parasites (*Cryptosporidium* spp.
-and *Cyclospora* spp.
-[@cdc2022].
-Environmental factors such as heavy metals can also cause foodborne illness.
+coli*, to viruses (norovirus and rotavirus), to parasites (*Cryptosporidium* spp.
+and *Cyclospora* spp), as well as toxins like marine toxins or metal toxins [@cdc2022].
+To verify pathogens of greatest concern, common settings associated with foodborne illness, and foods typically associated with such outbreaks, this study was launched as a research project at the University of Georgia using information gathered over a 17-year period from the Centers for Disease Control and Prevention (CDC).
 
@@
 
-To verify pathogens of greatest concern, common settings associated with food poisoning, and foods typically associated with such outbreaks, this study was launched as a research project at the University of Georgia using information gathered over a 17-year period from the Centers for Disease Control and Prevention (CDC).
-
-From this research project, we would like to answer the following questions:
-
--   What are the most frequent locations that foodborne outbreaks occur?
-
--   What pathogen species are most frequently associated with foodborne outbreaks?
-
--   What type of food is most often seen for foodborne outbreaks?
-
-In other words, immediate outcomes of interest are location, disease etiology, and food category. These will act as predictors for the health outcomes of interest: illnesses and hospitalizations.
-
-Our hypotheses include:
-
--   Outbreaks occur most often in homes rather than restaurants.
-
--   The most common pathogen associated with foodborne disease outbreaks is *E. coli*.
-
--   The food most commonly associated with foodborne outbreaks consists of salad.
+This research project is designed to answer the following questions, 1).
+the most frequent locations that foodborne outbreaks occur, 2).
+the most associated etiology agents, and 3).
+the food vehicle linked to foodborne outbreaks.
+In other words, immediate outcomes of interest are location, disease etiology, and food category.
+Furthermore, these were employed as predictors for building a model aimed at forecasting the the health outcomes of interest: illnesses and hospitalizations.
+Our research postulates three hypotheses based on empirical evidence, 1).
+foodborne disease outbreaks exhibit higher incidence rates in home settings compared to restaurant settings, 2).
+*E.
+coli* represents the most frequently detected pathogenic agent in cases of foodborne disease outbreaks, and 3).
+salad is the food item most commonly associated with foodborne disease outbreaks.
 
@@
 Two datasets were adopted for this research.
-The first dataset--which was used to build statistical models for the study--was uploaded onto *Kaggle* in 2017, with 12 variables and more than 16,950 observations spanning almost two decades from 1998 to 2015 [@cdc2017a].
-The second dataset--which was used for machine learning--was retrieved from the CDC's National Outbreak Reporting System (NORS) website.
-This second dataset covers reports of foodborne and waterborne disease outbreaks, as well as enteric disease outbreaks spread by contact with environmental sources, infected animals or persons, and other means from 2009 to 2020.
+The first dataset, which was used to build statistical models for the study, was uploaded onto Kaggle in 2017 and originated in CDC’s National Outbreak Reporting System (NORS) website, with 12 variables and more than 16,950 observations (CDC, 2017).
+The second dataset, which was used for evaluating the model performance, was retrieved from the CDC’s NORS website.
+The second data set covers reports of foodborne and waterborne disease outbreaks, as well as enteric disease outbreaks spread by contact with environmental sources, infected animals or persons, and other means from 2009 to 2020.
 
@@
 After processing procedures, an exploratory data analysis was performed to view general relationships between predictors (time, pathogen type, pathogen etiology, state, and food type) and the health outcomes of interest (illnesses and hospitalizations).
+We observed certain patterns between preditors and outcomes and selected
 
-### Machine Learning
+### Statistical Analysis and Machine Learning
 
 Linear and Poisson regression models were afterwards constructed using the aforementioned predictors and outcomes from the training data.
-Data was resampled using the bootstrapping approach 1,000 times for better estimates of accuracy and variability of model coefficients.
-Random forest and bagging models were also trained to predict health outcomes using the Kaggle dataset for the NORS-CDC data.
-RMSE was used to compare each of the models.
-The best model built from the training data was then employed for the test data.
+Specifically, We first assessed the univariate relationships between the main outcome `Simple Etiology` and each health outcome `Illnesses` and `Hospitalizations` using linear regression.
+Multiple linear and Poisson regression models were also performed with each health outcome.
+Machine learning approaches including random forest and bagging models were also trained to predict health outcomes using the Kaggle dataset for the NORS-CDC data.
+For machine learning models, we tuned hyperparameters using randomized grid search.
+The model with the lowest RMSE among all the models was chosen as the best model from the training data, and subsequently utilized for the test data.
 
@@
 
-### Training Data
-
-```{r Table of Hospitalizations by State}
-first_table <- readRDS("../../results/state_v_hospitalization_training.rds") %>% 
-  rename(Hospitalizations = all_hospitalizations)
-first_table <- flextable(first_table)
-first_table <- theme_vanilla(first_table)
-first_table <- set_caption(first_table, caption = as_paragraph(colorize("Table 1. Hospitalizations by state from 1998 to 2015.", color = "black")))
-first_table <- set_table_properties(first_table, layout = "autofit")
-first_table
-```
-
-![Figure 1. Hospitalizations by food type from 1998-2015.](images/Food_type_v_hospitalization-01.png)
-
-```{r}
-second_table <- readRDS("../../results/training_food_vs_illnesses.rds") %>% 
-  select(`IFSAC Category`, all_illnesses) %>% 
-  rename(Illnesses = all_illnesses)
-second_table <- flextable(second_table)
-second_table <- theme_vanilla(second_table)
-second_table <- set_caption(second_table, caption = as_paragraph(colorize("Table 2. Illnesses by food type from 1998 to 2015.", color = "black")))
-second_table
-```
-
-### Test Data
-
-![Figure 2. Illness over time by pathogen type (2009-2020).](images/time_v_illness_pathogen_type.png)
-
-![Figure 3. Illnesses by outbreak location (2009-2020).](images/Location_v_Illness-01.png)
-
-![Figure 4. Bacterial etiology by outbreak frequency (2009-2020).](images/bacterial_etiology_v_outbreak_frequency.png)
+After applying the aforementioned data processing method to the raw data, a total of 3985 observations were obtained in the training data set.
+In the United States, Multi-state outbreaks were the most commonly associated with hospitalizations caused by foodborne illnesses, followed by California, Ohio, and New York.
+Poultry, fruits, beef, and pork were identified as the most frequent food vehicles linked to mass hospitalizations (involving over 500 individuals) due to foodborne illnesses (Figure 2).
+Nevertheless, despite poultry being the top food vehicle, salad was identified as the primary culprit for causing the most foodborne illnesses from 1998 to 2005, as shown in Table 2.
+Figure 1 unveiled patterns concerning the seasonality of foodborne illness outbreaks.
+Hospitalizations due to foodborne illnesses peaked in July, while reaching their lowest point between December and February.
+Intriguingly, bacteria is the predominant etiologic agent that causes hospitalization (Figure 3).
+Among these bacteria, the genus *Salmonella* is the most frequently identified etiologic agent responsible for foodborne illness-related hospitalizations.Most commonly implicated food vehicles for *Salmonella* include fruits, pork, and poultry (Figure 5).
+In contrast, *E.
+coli* and *Clostridium*, two other bacterial species associated with foodborne hospitalizations, were most frequently linked to vegetables as the primary food carrier.
+Restaurant and private home were locations where the majority of foodborne hospitalizations occurred (Figure 6).
+The testing data set revealed a number of trends that were consistent with those found in the training data set.
+For instance, *Salmonella* continued to be the most common bacterial pathogen responsible for foodborne illnesses, and restaurants remained the most frequently implicated exposure setting (Figure 7 and Figure 8).
+However, some shifts were also observed between the two data sets.
+While Table 2 had previously shown salad to be the leading food carrier of foodborne illnesses, the testing data set revealed that poultry was actually the most common source of illness between 2009 and 2020 (Figure 9).
+In addition, the testing data set revealed that *E.
+coli* had replaced the previously identified second most common bacterial pathogen *Clostridium*, causing a significant number of illnesses between 2009 and 2020 (Figure 8).
 
@@
 
+Based on the findings of the exploratory analysis, there are clear associations between time (Month, Year), pathogen type, etiology, location, state, and food type (IFSAC Category) with health outcomes.
+However, since there is a strong correlation between etiology and pathogen types, the latter was removed.
+Consequently, our predictive models were constructed using the chosen predictors of time (Month, Year), etiology, location, state, and food type (IFSAC Category).
+
 ### Basic
 
-Between simple and multivariate regressions, multivariate models performed better with lower associated RMSEs.
-For both health outcomes, the multivariate linear regression models (RMSE = 58.4, 5.50) constructed performed better than multivariate poisson regression models (RMSE = 59.7, 5.64).
-
-![Figure 5.
-Predicted illnesses versus actual illness count for multivariate linear regression model.](images/Predicted%20Illnesses%20VS%20Illnesses%20by%20linear%20regression%20with%20all%20predictors.png){fig-align="center"}
+Between univariate and multivariate regressions, multivariate models performed better with lower associated RMSEs.
+For both health outcomes, the multivariate linear regression models (RMSE = 58.4, 5.50) constructed performed better with  and than multivariate poisson regression models `r $RMSE = 59.7, 5.64).
 
@@
 
+In addition to the above information, it's worth noting that the data also revealed some interesting trends related to the seasonality and demographics of foodborne illness outbreaks.
+For instance, the incidence of outbreaks tends to be highest during the summer months, likely due to factors such as increased outdoor food preparation and consumption of raw or undercooked foods.
+
+In exploratory analysis, slight shift regarding the most common food vehicle that led to most foodborne illnesses (Table 2 and Figure 9) was obtained.
+This highlighted the dynamic nature of foodborne illness outbreaks, and the need for ongoing monitoring and surveillance to identify emerging trends and adapt food safety practices accordingly.
+Efforts to improve food safety must be responsive to these changing patterns to prevent and control the spread of pathogens.
+
 The remaining two of the three initial hypotheses were proved incorrect.
@@
 Previous surveillance data points to *Salmonella* bacteria and salads as the main cause and medium for foodborne disease outbreaks, along with the most common setting for such outbreaks being a restaurant.
+This study allowed for a comprehensive understanding of the potential variables that may impact the occurrence and severity of foodborne disease outbreaks.
+Identifying key patterns and associations enabled the development of more targeted and effective intervention strategies to mitigate the risks of foodborne illness,and enhance public health outcomes.
 These findings will also give future outbreak investigators a first direction to consider in the field.
@@
 # References
+# Appendix: Tables and Plots
+
+![Figure 1. Hospitalizations by month from 1998-2015.](images/month_v_hospitalization.png)
+
+![Figure 2. Common contaminated food linked to hospitalizations from 1998-2015.](images/Food_type_v_hospitalization.png)
+
+![Figure 3. Pathogen types and Hospitalizations.](images/pathogen_type_v_hospitalizations.png)
+![Figure 4. Common bacterial etiology causing hospitalizations from 1998-2015.](images/bacterial_etiology_v_hospitalization_frequency.png)
+![Figure 5. Top 3 bacterial pathogens and their common food carrier month from 1998-2015.](images/top3_bacterial_etiology_v_food_hospitalization_frequency.png)
+![Figure 5. Leading locations that result in hospitalizations ](images/Location_v_hospitalization.png)
+![Figure 6. Illness over time by pathogen type (2009-2020).](images/time_v_illness_pathogen_type.png)
+![Figure 7. Illnesses by outbreak location (2009-2020).](images/Location_v_Illness.png)
+![Figure 8. Bacterial etiology by outbreak frequency (2009-2020).](images/bacterial_etiology_v_outbreak_frequency.png)
+![Figure 9. Food by outbreak frequency (2009-2020).](images/food_type_v_illness_frequency.png)
+
+```{r}
+ first_table <- readRDS("../../results/state_v_hospitalization_training.rds") %>% 
+   rename(Hospitalizations = all_hospitalizations)
+first_table <- flextable(first_table) 
+first_table <- theme_vanilla(first_table) 
+first_table <- set_caption(first_table, caption = as_paragraph(colorize("Table 1. Hospitalizations by state from 1998 to 2015.", color = "black")))
+first_table <- set_table_properties(first_table, layout = "autofit")
+first_table
+```
+
+```{r}
+second_table <- readRDS("../../results/training_food_vs_illnesses.rds") %>% 
+  select(`IFSAC Category`, all_illnesses) %>% 
+  rename(Illnesses = all_illnesses)
+second_table <- flextable(second_table)
+second_table <- theme_vanilla(second_table)
+second_table <- set_caption(second_table, caption = as_paragraph(colorize("Table 2. Illnesses by food type from 1998 to 2015.", color = "black")))
+second_table
+```

### 2023-04-22 code/analysis_code/exploratory_analysis.qmd
Weifan_final

@@
 # General
+## Hospitalizations
 ggplot(training_data_final %>% 
@@
 ggsave("bacterial_etiology_v_hospitalization_frequency.png", path = image_path, width = 3000, height = 2000, units = "px")
+
+## Illnesses
+ggplot(training_data_final %>% 
+         select(Pathogen_Type, Simplified_Etiology, Illnesses) %>%
+         filter(Pathogen_Type == "Bacteria")%>%
+         group_by(Simplified_Etiology)%>%
+         mutate(all_Illnesses=sum(Illnesses)),
+       aes(x = fct_reorder(Simplified_Etiology,all_Illnesses))) + geom_bar() + 
+  labs(x = "Bacteria", y = "Illnesses", 
+       title = "Bar Chart of Bacteria Frequency") + coord_flip()
+
+ggsave("bacterial_etiology_v_Illnesses_frequency.png", path = image_path, width = 3000, height = 2000, units = "px")
 
@@
 # Food Type and Hospitalizations
-ggplot(training_data_final %>% 
+food_hosp=training_data_final %>% 
          filter(`IFSAC Category` != "Multiple") %>% 
@@
          mutate(all_hospitalizations = sum(Hospitalizations)) %>%
-         count(all_hospitalizations), 
-       aes(x = fct_reorder(`IFSAC Category`,all_hospitalizations), y = all_hospitalizations, 
+         count(all_hospitalizations)%>%
+  arrange(desc(all_hospitalizations))
+save_location <- here("results","tables", "food_v_hospitalizations.rds")
+saveRDS(food_hosp,file = save_location)
+
+  ggplot(food_hosp,aes(x = fct_reorder(`IFSAC Category`,all_hospitalizations), y = all_hospitalizations, 
            )) + 

### 2023-04-22 code/analysis_code/statistical_analysis.qmd
Weifan_final

@@
 ```
+### Organize performances metrics for single variate models
+```{r}
+
+simp_illness_metric=lm_simple_resampled_illness%>%
+    collect_metrics() %>%
+  mutate(model="linear")%>%
+  filter(.metric=='rmse')%>%
+  bind_rows(poisson_simple_resampled_illness %>%
+  collect_metrics() %>%
+  filter(.metric=='rmse')%>%
+    mutate(model="Poisson"))
+
+save_location <- here("results", "tables","illness_simple_rmse.rds")
+saveRDS(simp_illness_metric, file = save_location)
+```
+
 According to the result, the model built using multivariate linear regression seems to have the smallest root mean squared error and highest R^2.
@@
 ```
+
+### Organize performances metrics for single variate models
+```{r}
+simp_hospital_metric=lm_simple_resampled_hosp%>%
+    collect_metrics() %>%
+  filter(.metric=='rmse')%>%
+  mutate(model="linear")%>%
+  bind_rows(poisson_simple_resampled_hosp%>%
+  collect_metrics() %>%
+  filter(.metric=='rmse')%>%
+    mutate(model="poisson"))
+
+save_location <- here("results", "tables","hospital_simple_rmse.rds")
+saveRDS(simp_hospital_metric, file = save_location)
+```
+
 According to the results, the model built using multivariate linear regression seems to have the smallest root mean squared error and highest R^2 (14%).
@@
 # Illness
-save_location <- here("results", "illness_rmse_models.rds")
+# Univariate
+
+# Multivariate
+save_location <- here("results","tables", "illness_rmse_models.rds")
 saveRDS(illness_model_dataframe, file = save_location)
@@
 # Hospitalizations
-save_location <- here("results", "hospital_rmse_models.rds")
+save_location <- here("results", "tables","hospital_rmse_models.rds")
 saveRDS(hosp_model_dataframe, file = save_location)

### 2023-04-22 products/manuscript/Manuscript2.qmd
Weifan_final

@@
 date: "May 2023"
-output: word_document
+output:
+  bookdown::word_document2:
+    toc: false
 bibliography: ../references.bib
@@
 
+```{r setup, include=FALSE, echo=FALSE, cache.lazy=FALSE}
+# Setting options for the rest of the document
+knitr::opts_chunk$set(warning=FALSE)
+
+# Rounding rules
+options(digits = 3)
+#Load clean data from "processed_training_data.rds"
+# Path to Data
+data_location <- here(
+  "data","processed_data","processed_training_data.rds")
+# Import Data
+training_data_final <- readRDS(data_location)
+
+#Load table from "state_v_hospitalization_training.rds"
+# Path to Data
+data_location <- here(
+  "results","tables","state_v_hospitalization_training.rds")
+# Import Data
+state_hosp <- readRDS(data_location)
+
+# Load table from "state_v_illness_training.rds"
+# Path to Data
+data_location <- here(
+  "results","tables","state_v_illness_training.rds")
+# Import Data
+state_illness <- readRDS(data_location)
+
+# Load table from "food_v_hospitalizations.rds"
+# Path to Data
+data_location<-here("results","tables", "food_v_hospitalizations.rds")
+# Import Data
+food_hosp<-readRDS(data_location)
+
+# Load table from "training_food_vs_illnesses.rds"
+# Path to Data
+data_location<-here("results","tables", "training_food_vs_illnesses.rds")
+# Import Data
+food_illness<-readRDS(data_location)
+
+# Load rmse data for outcome `Illnesses`
+
+# Univariate for linear and poisson regression models
+# Path to Data
+data_location<-here("results","tables","illness_simple_rmse.rds")
+# Import Data
+illness_uni_rmse<-readRDS(data_location)
+
+# Multivariate
+# Path to Data
+data_location=here("results","tables","illness_rmse_models.rds")
+# Import Data
+illness_rmse<-readRDS(data_location)
+
+# Load rmse data for outcome `Hospitalizations`
+# Univariate for linear and poisson regression models
+# Path to Data
+data_location<-here("results","tables","hospital_simple_rmse.rds")
+# Import Data
+hosp_uni_rmse<-readRDS(data_location)
+
+# Multivariate
+# Path to Data
+data_location=here("results","tables","hospital_rmse_models.rds")
+# Import Data
+hosp_rmse<-readRDS(data_location)
+
+# Final model metrics
+
+```
+
 # Summary/Abstract
+
+In the United States, it has been reported that one in six Americans acquire foodborne illnesses annually, leading to around 128,000 hospitalizations and 3,000 deaths.
+Our project aims to identify the major etiology agents responsible for foodborne illnesses and their associated factors in the U.S.
+We utilized two datasets collected by the CDC, which contain foodborne outbreak information from 1998-2015 and 2009-2020, respectively.
+The first dataset was used for model training, while the second dataset was reserved for testing purposes.
+We developed regression models, including linear and Poisson regression, to predict health outcomes using selected predictors from the training dataset.
+Additionally, random forest model and bagging approaches were employed to refine the model.
+Results showed that multi-state outbreaks were the most commonly associated with total hospitalizations.
+Poultry was identified as the most frequent food vehicles linked to mass hospitalizations due to foodborne illnesses while salad was identified as the primary culprit for causing the most foodborne illnesses from 1998 to 2015.
+Bacteria is the predominant etiologic agent that causes hospitalization, among which the genus *Salmonella* is the most frequently identified etiologic agent responsible for foodborne illness-related hospitalizations.
+Most commonly implicated food vehicles for *Salmonella* include fruits, pork, and poultry.
+The best model constructed based on time, pathogen type, pathogen etiology, state, and food type was random forest model, with the lowest RMSE of `r illness_rmse$RMSE[1]`, `r hosp_rmse$RMSE[1]` for illnesses and hospitalizations, respectively.
 
@@
 After processing procedures, an exploratory data analysis was performed to view general relationships between predictors (time, pathogen type, pathogen etiology, state, and food type) and the health outcomes of interest (illnesses and hospitalizations).
-We observed certain patterns between preditors and outcomes and selected
+We observed certain patterns between predictors and outcomes and selected predictors accordingly for further model construction.
 
@@
 After applying the aforementioned data processing method to the raw data, a total of 3985 observations were obtained in the training data set.
-In the United States, Multi-state outbreaks were the most commonly associated with hospitalizations caused by foodborne illnesses, followed by California, Ohio, and New York.
+The illnesses caused by a single outbreak ranged from `r min (training_data_final$Illnesses)` to `r max (training_data_final$Illnesses)` while resulted hospitalizations varied from `r min (training_data_final$Hospitalizations)` to `r max (training_data_final$Hospitalizations)` from 1998 to 2015.
+In the United States, Multi-state outbreaks were the most commonly associated with total hospitalizations (N= `r max (state_hosp$all_hospitalizations)`) caused by foodborne illnesses, followed by California (N=`r state_hosp$all_hospitalizations[2]`), Ohio (N=`r state_hosp$all_hospitalizations[3]`), and New York (n=`r state_hosp$all_hospitalizations[4]`).
 Poultry, fruits, beef, and pork were identified as the most frequent food vehicles linked to mass hospitalizations (involving over 500 individuals) due to foodborne illnesses (Figure 2).
-Nevertheless, despite poultry being the top food vehicle, salad was identified as the primary culprit for causing the most foodborne illnesses from 1998 to 2005, as shown in Table 2.
+Nevertheless, despite poultry being the top food vehicle, salad was identified as the primary culprit for causing the most foodborne illnesses from 1998 to 2015, as shown in Table 2.
 Figure 1 unveiled patterns concerning the seasonality of foodborne illness outbreaks.
@@
 Between univariate and multivariate regressions, multivariate models performed better with lower associated RMSEs.
-For both health outcomes, the multivariate linear regression models (RMSE = 58.4, 5.50) constructed performed better with  and than multivariate poisson regression models `r $RMSE = 59.7, 5.64).
+For both health outcomes, the multivariate linear regression models (RMSE = `r illness_rmse$RMSE[2]`, `r hosp_rmse$RMSE[3]`) constructed performed better with  and than multivariate poisson regression models (RMSE = `r illness_rmse$RMSE[3]`, `r hosp_rmse$RMSE[4]`).
 
@@
 # Illness Models 
-rmarkdown::paged_table(readRDS("../../results/illness_rmse_models.rds"))
+rmarkdown::paged_table(readRDS("../../results/tables/illness_rmse_models.rds"))
 
 # Hospitalization Models
-rmarkdown::paged_table(readRDS("../../results/hospital_rmse_models.rds"))
+rmarkdown::paged_table(readRDS("../../results/tables/hospital_rmse_models.rds"))
 ```
 
-Overall, between itself and the other models (linear regression, poisson regression, and bagging) generated, the random forest model performed the best for both health outcomes with the lowest RMSEs (55.7, 5.05).
+Overall, between itself and the other models (linear regression, poisson regression, and bagging) generated, the random forest model performed the best for both health outcomes with the lowest RMSEs (`r illness_rmse$RMSE[1]`, `r hosp_rmse$RMSE[1]`).
 
@@
 
-From 1998 to 2015, multistate outbreaks caused the most illnesses (N = 18,617) and hospitalizations (3,506) than any in-state outbreak.
+From 1998 to 2015, multistate outbreaks caused the most illnesses (N = `r state_illness$all_illnesses[1]`) and hospitalizations (N = `r max (state_hosp$all_hospitalizations)`) than any in-state outbreak.
 The fact that multistate outbreaks were seen more often than in any individual state for illnesses and hospitalizations suggests that these disease outbreaks typically involve a regional or national ingredient distributor.
@@
 
-In terms of food categories, poultry was identified to have caused the most hospitalizations (N = 770), followed closely by fruits and beef.
-Interestingly, however, salads were found to cause the most illnesses (N = 11,493).
+In terms of food categories, poultry was identified to have caused the most hospitalizations (N = `r food_hosp$all_hospitalizations[1]`), followed closely by fruits (N = `r food_hosp$all_hospitalizations[2]`) and beef (N = `r food_hosp$all_hospitalizations[3]`).
+Interestingly, however, salads were found to cause the most illnesses (N = `r max(food_illness$all_illnesses)`).
 Upon addition of total hospitalizations and illnesses for both poultry and salads, salads were determined to have the highest number of cases in total, which confirms our original hypothesis.
@@
 coli*, results overall indicated that *Salmonella* spp.
-are the most common pathogens linked to foodborne disease; these bacteria contribute to the greatest amount of illnesses and hospitalizations in comparison to other disease agents for our data.
-Finally, we also see that food poisoning is most frequently linked to restaurants instead of private households.
+were the most common pathogens linked to foodborne disease; these bacteria contribute to the greatest amount of illnesses and hospitalizations in comparison to other disease agents for our data.
+As a matter of fact, contaminated poultry, eggs, beef, and pork are the most common food vehicles that contribute to *Salmonella* outbreaks (@jackson_outbreak-associated_2013).
+The result obtained from a previous study conducted in Poland from 2010 to 2015 studying the distribution of  *Salmonella* serovars identified from food, animals, and feedingstuffs is consistent with our findings (@skarzynska_distribution_2017).
+The study revealed that *Salmonella* most frequently recovered from poultry samples, and *S*.
+Typhimurium being the most prevalent in pork products.
 
-\[Something about machine learning here???\]
+Finally, we observed that food poisoning is most frequently linked to restaurants instead of private households.
+This result is in agreement with a previous study that revealed over 40 % of outbreaks could be traced back to restaurants (@hedberg_systematic_2006).
+In contrast, some Chinese researchers reported that the most common exposure setting was household, followed by restaurants (@li_surveillance_2020).
+
+\[Something about machine learning here???\] #you can briefly talk about the algorithm or working principle behind random forest, what attributes may make it the best model?
 
@@
 ## Strengths and Limitations
+
+Future research can focus on the associations between subspecies of identified common bacterial etiology and crucial factors.
+Text mining approach might be implemented to explore more food carriers linked to foodborne outbreaks.
 
@@
 ```{r}
- first_table <- readRDS("../../results/state_v_hospitalization_training.rds") %>% 
+ first_table <- readRDS("../../results/tables/state_v_hospitalization_training.rds") %>% 
    rename(Hospitalizations = all_hospitalizations)
@@
 ```{r}
-second_table <- readRDS("../../results/training_food_vs_illnesses.rds") %>% 
+second_table <- readRDS("../../results/tables/training_food_vs_illnesses.rds") %>% 
   select(`IFSAC Category`, all_illnesses) %>% 

### 2023-04-23 products/manuscript/Manuscript2.qmd

@@
 The primary responsibility for the identification, investigation, and reporting of foodborne disease outbreaks caused by various enteric microorganism, and chemical/toxic agents lies with state, local, and territorial public health departments, in accordance with established scientific protocols and procedures [@hoffman_capacity_2005].
-Previous studies have identified that restaurants are sites frequently attributed to mass-outbreaks, and typical foods associated with these outbreaks include improperly-prepared meats and vegetables [@mun2020;@sarno2021].
+Previous studies have identified that restaurants are sites frequently attributed to mass-outbreaks, and typical foods associated with these outbreaks include improperly-prepared meats and vegetables [@mun2020; @sarno2021].
 The etiologies leading to disease encompass a wide range of microorganisms, ranging from bacteria such as *Salmonella* spp.
@@
 Two datasets were adopted for this research.
-The first dataset, which was used to build statistical models for the study, was uploaded onto Kaggle in 2017 and originated in CDC’s National Outbreak Reporting System (NORS) website, with 12 variables and more than 16,950 observations (CDC, 2017).
-The second dataset, which was used for evaluating the model performance, was retrieved from the CDC’s NORS website.
+The first dataset, which was used to build statistical models for the study, was uploaded onto Kaggle in 2017 and originated in CDC's National Outbreak Reporting System (NORS) website, with 12 variables and more than 16,950 observations (CDC, 2017).
+The second dataset, which was used for evaluating the model performance, was retrieved from the CDC's NORS website.
 The second data set covers reports of foodborne and waterborne disease outbreaks, as well as enteric disease outbreaks spread by contact with environmental sources, infected animals or persons, and other means from 2009 to 2020.
@@
 were the most common pathogens linked to foodborne disease; these bacteria contribute to the greatest amount of illnesses and hospitalizations in comparison to other disease agents for our data.
-As a matter of fact, contaminated poultry, eggs, beef, and pork are the most common food vehicles that contribute to *Salmonella* outbreaks (@jackson_outbreak-associated_2013).
-The result obtained from a previous study conducted in Poland from 2010 to 2015 studying the distribution of  *Salmonella* serovars identified from food, animals, and feedingstuffs is consistent with our findings (@skarzynska_distribution_2017).
+As a matter of fact, contaminated poultry, eggs, beef, and pork are the most common food vehicles that contribute to *Salmonella* outbreaks [@jackson_outbreak-associated_2013].
+The result obtained from a previous study conducted in Poland from 2010 to 2015 studying the distribution of *Salmonella* serovars identified from food, animals, and feedingstuffs is consistent with our findings [@skarzynska_distribution_2017].
 The study revealed that *Salmonella* most frequently recovered from poultry samples, and *S*.
@@
 Finally, we observed that food poisoning is most frequently linked to restaurants instead of private households.
-This result is in agreement with a previous study that revealed over 40 % of outbreaks could be traced back to restaurants (@hedberg_systematic_2006).
-In contrast, some Chinese researchers reported that the most common exposure setting was household, followed by restaurants (@li_surveillance_2020).
+This result is in agreement with a previous study that revealed over 40% of outbreaks could be traced back to restaurants [@hedberg_systematic_2006].
+In contrast, some Chinese researchers reported that the most common exposure setting was household, followed by restaurants [@li_surveillance_2020].
 
-\[Something about machine learning here???\] #you can briefly talk about the algorithm or working principle behind random forest, what attributes may make it the best model?
+Lastly, the random forest model outperformed all other models.
+It is also interesting to note that bagging resulted in a higher RMSE than random forest each time despite using all predictors in its splitting, which should have greatly reduced variance.
+However, random forest models take a random subset of the 10 predictors for splitting; average variance is thus reduced from high to moderate--meaning that random forest should have performed worse.
+Additional future investigation is suggested to uncover why.
 
-These results offer insight into common disease characteristics across food poisoning cases, which is greatly useful for epidemiologists and other public health professionals to anticipate which areas to investigate first for future outbreaks.
+Overall, these results offer insight into common disease characteristics across food poisoning cases, which is greatly useful for epidemiologists and other public health professionals to anticipate which areas to investigate first for future outbreaks.
 
@@
 
-# References
 # Appendix: Tables and Plots
@@
 
-![Figure 3. Pathogen types and Hospitalizations.](images/pathogen_type_v_hospitalizations.png)
-![Figure 4. Common bacterial etiology causing hospitalizations from 1998-2015.](images/bacterial_etiology_v_hospitalization_frequency.png)
-![Figure 5. Top 3 bacterial pathogens and their common food carrier month from 1998-2015.](images/top3_bacterial_etiology_v_food_hospitalization_frequency.png)
-![Figure 5. Leading locations that result in hospitalizations ](images/Location_v_hospitalization.png)
-![Figure 6. Illness over time by pathogen type (2009-2020).](images/time_v_illness_pathogen_type.png)
-![Figure 7. Illnesses by outbreak location (2009-2020).](images/Location_v_Illness.png)
-![Figure 8. Bacterial etiology by outbreak frequency (2009-2020).](images/bacterial_etiology_v_outbreak_frequency.png)
-![Figure 9. Food by outbreak frequency (2009-2020).](images/food_type_v_illness_frequency.png)
+![Figure 3.
+Pathogen types and Hospitalizations.](images/pathogen_type_v_hospitalizations.png) ![Figure 4.
+Common bacterial etiology causing hospitalizations from 1998-2015.](images/bacterial_etiology_v_hospitalization_frequency.png) ![Figure 5.
+Top 3 bacterial pathogens and their common food carrier month from 1998-2015.](images/top3_bacterial_etiology_v_food_hospitalization_frequency.png) ![Figure 5.
+Leading locations that result in hospitalizations](images/Location_v_hospitalization.png) ![Figure 6.
+Illness over time by pathogen type (2009-2020).](images/time_v_illness_pathogen_type.png) ![Figure 7.
+Illnesses by outbreak location (2009-2020).](images/Location_v_Illness.png) ![Figure 8.
+Bacterial etiology by outbreak frequency (2009-2020).](images/bacterial_etiology_v_outbreak_frequency.png) ![Figure 9.
+Food by outbreak frequency (2009-2020).](images/food_type_v_illness_frequency.png)
 
@@
 ```
+
+# References

### 2023-04-23 code/analysis_code/statistical_analysis.qmd
update
final RMSEs

@@
 
-test_data_final %>% select(Illnesses) %>% 
+final_illness_fit <- test_data_final %>% select(Illnesses) %>% 
   bind_cols(placeholder_illnesses) %>% rmse(Illnesses, .pred)
@@
 
-test_data_final %>% select(Hospitalizations) %>% 
+final_hosp_fit <- test_data_final %>% 
+  select(Hospitalizations) %>% 
   bind_cols(placeholder_hosp) %>% rmse(Hospitalizations, .pred)
-
 ```
@@
 # Illness
-# Univariate
-
-# Multivariate
 save_location <- here("results","tables", "illness_rmse_models.rds")
 saveRDS(illness_model_dataframe, file = save_location)
+
+save_location <- here("results", "tables", "final_illness_fit.rds")
+saveRDS(final_illness_fit, file = save_location)
 
@@
 saveRDS(hosp_model_dataframe, file = save_location)
+
+save_location <- here("results", "tables", "final_hosp_fit.rds")
+saveRDS(final_hosp_fit, file = save_location)
 ```

### 2023-04-23 products/manuscript/Manuscript2.qmd
final RMSEs

@@
 # Final model metrics
+data_location <- here("results", "tables", "final_illness_fit.rds")
+final_illness_fit <- readRDS(data_location)
 
+data_location <- here("results", "tables", "final_hosp_fit.rds")
+final_hosp_fit <- readRDS(data_location)
 ```
@@
 
-Fitting the finalized random forest model to the test data resulted in RMSEs of 36.6 and 6.70 for the health outcomes of illnesses and hospitalizations, respectively.
+Fitting the finalized random forest model to the test data resulted in RMSEs of `r final_illness_fit$.estimate[1]` and `r final_hosp_fit$.estimate[1]` for the health outcomes of illnesses and hospitalizations, respectively.
 

### 2023-04-24 products/manuscript/Manuscript.qmd
add vip to analysis and manuscript

--- 2023-03-17 products/manuscript/Manuscript.qmd
@@
 ---
-title: "Manuscript for foodborne-outbreak Project"
-subtitle: ""
-author: Andreas Handel
-date: "`r Sys.Date()`"
+title: "Manuscript for Foodborne-Outbreak Project"
+# subtitle: ""
+author: "Weifan Wu and Kailin Chen"
+# date: "`r Sys.Date()`"
 format: 
-  docx:
-    toc: false
-    number-sections: true
-    highlight-style: github
-bibliography: ../references.bib
-csl: ../apa.csl
-editor: 
-  markdown: 
-    wrap: sentence
+  docx
+#    toc: false
+#    number-sections: true
+#    highlight-style: github
+# bibliography: ../references.bib
+# csl: ../apa.csl
+# editor: 
+#  markdown: 
+#    wrap: sentence
 ---

### 2023-04-24 products/manuscript/Manuscript2.qmd
add vip to analysis and manuscript

@@
 
-```{r echo = FALSE, warning = FALSE, paged.print = TRUE}
+```{r include=FALSE,echo = FALSE, warning = FALSE}
 # Load Libraries
@@
 Nevertheless, despite poultry being the top food vehicle, salad was identified as the primary culprit for causing the most foodborne illnesses from 1998 to 2015, as shown in Table 2.
+
 Figure 1 unveiled patterns concerning the seasonality of foodborne illness outbreaks.
@@
 
-Lastly, the random forest model outperformed all other models.
+Lastly, the random forest model outperformed all other models in minimizing RMSE.
+Our findings reveal that the 'Multistate' variable holds the highest importance in determining hospitalization predictions, followed by 'Year', 'Month', and 'IFSAC Category_Fruits'.
+'Location' also plays a significant role in the model's predictive power (Figure 10).
+These results suggest that geographical factors, such as multistate outbreaks, and temporal factors, such as year and month, play a crucial role in predicting hospitalizations.
+Furthermore, the specific IFSAC category, particularly 'Fruits', and the presence of *Salmonella* contribute significantly to the model's performance, indicating that certain food categories and bacterial etiologies may have a higher association with hospitalization rates.
+
 It is also interesting to note that bagging resulted in a higher RMSE than random forest each time despite using all predictors in its splitting, which should have greatly reduced variance.
@@
 Food by outbreak frequency (2009-2020).](images/food_type_v_illness_frequency.png)
+![Figure 10. Variable importance in determining hospitalization predictions using random forest.](images/variable importance in predicting hospitalizations by random forest model.png)
 

### 2023-04-24 code/analysis_code/statistical_analysis.R

@@ -1,0 +1,55 @@
+###############################
+# analysis script
+#
+#this script loads the processed, cleaned data, does a simple analysis
+#and saves the results to the results folder
+
+#load needed packages. make sure they are installed.
+library(ggplot2) #for plotting
+library(broom) #for cleaning up output from lm()
+library(here) #for data loading/saving
+
+#path to data
+#note the use of the here() package and not absolute paths
+data_location <- here::here("data","processed_data","processeddata.rds")
+
+#load data. 
+mydata <- readRDS(data_location)
+
+
+######################################
+#Data fitting/statistical analysis
+######################################
+
+############################
+#### First model fit
+# fit linear model using height as outcome, weight as predictor
+
+lmfit1 <- lm(Height ~ Weight, mydata)  
+
+# place results from fit into a data frame with the tidy function
+lmtable1 <- broom::tidy(lmfit1)
+
+#look at fit results
+print(lmtable1)
+
+# save fit results table  
+table_file1 = here("results", "resulttable1.rds")
+saveRDS(lmtable1, file = table_file1)
+
+############################
+#### Second model fit
+# fit linear model using height as outcome, weight and sex as predictor
+
+
+# place results from fit into a data frame with the tidy function
+lmtable2 <- broom::tidy(lmfit2)
+
+#look at fit results
+print(lmtable2)
+
+# save fit results table  
+table_file2 = here("results", "resulttable2.rds")
+saveRDS(lmtable2, file = table_file2)
+
+  

### 2023-04-28 Project_Review_DeannaL.qmd

---
title: Project Review Weifan and Kai
author: Deanna Lanier
date: "`r file.mtime(knitr::current_input())`"
format: 
  html:
    toc: true
    toc-depth: 3
    number-sections: true
---

# Overview

Title of project: Foodborne-Outbreak

Name of project author(s): Weifan Wu & Kailin (Kai) Chen

Name of project reviewer: Deanna Lanier

# Specific project content evaluation
Evaluate the different parts of the project by filling in the sections below.

## Background, Context and Motivation
How well is the context of the project described? Is a comprehensive background, including summary of previous/related work given? Is the project well placed into the context of existing work (including proper referencing of existing work). Is it clear why the project was undertaken and what new information it hopes to provide?

### Feedback and Comments

The background does a good job of describing your topic. I would have liked to see more on previous or related work, or a statement saying that there is no previous work prior to this study.

### Summary assessment (PICK ONE, DELETE THE OTHERS)
* some contextualization and motivation

## Question description
How well and clear are the question(s)/hypotheses the project aims to address described? Is it clear how the questions relate to the data?

### Feedback and Comments

The questions and hypothesis were clearly stated and well explained. They relate to the data and the background well and are applicable to the analysis.

### Summary assessment
* question/hypotheses fully clear

## Data description
How well is the data overall described? Is the source provided? Is a codebook or other meta-information available that makes it clear what the data is?

### Feedback and Comments

The data are described in the data sources section and clearly explain where the data were obtained. There is no meta-information provided on the data sources so it is hard to be clear on what the variables are. I would recommend including an additional spreadsheet with your data dictionary so it is easier for other users to interpret.

### Summary assessment
* source and overall structure of data somewhat explained

## Data wrangling and exploratory analysis
How well is the data cleaned/processed and explored? Are all steps reasonable and well explained? Are alternatives discussed and considered? Are meaningful exploratory results shown (e.g. in the supplementary materials)?

### Feedback and Comments

The data processing was very comprehensive and well explained. I especially liked the Pareto Chart with Missing Values figures in your processing script. While I do think there could have been more documentation for the steps you took to clean with some rational, I think all of the steps were reasonable. For the exploratory analysis, the figures look great and all the results presented are related to the overarching study. I would like to have seen some sentences to explain the exploration and the figures provided. There were no 'takeaways' from the exploratory analysis just by reading the manuscript. The sentences found at the end of your exploratory analysis code would be a great addition to the manuscript to just give some general overview of what you are visualizing and the importance.

### Summary assessment
* essentially no weaknesses in wrangling and exploratory component

## Appropriateness of Analysis
Were the analysis methods appropriate for the data? Was the analysis done properly? Were different components of the analysis (e.g. performance measure, variable selection, data pre-processing, model evaluation) done in the best way possible and explained well?

### Feedback and Comments

Great analysis! You this statistical analysis was very appropriate for the questions and hypothesis and you utilized multiple models and included an great model interpretation with some performance metrics. I would help to include some model interpretation that relates back to the questions and hypothesis.

### Summary assessment
* strong and reasonable analysis

## Presentation
How well are results presented? Are tables and figures easy to read and understand? Are the main figures/tables publication level quality?

### Feedback and Comments

The figures are off to a great start. For all of you figures, I would increase the size of the text for the title, axis labels, axis titles and legend. There are too small to read without having to zoom in. Additionally, for figures 2 and 3, I would make the axis labels at an angle so that they are not overlapping. THe overlap makes the figure difficult to read.

### Summary assessment
* results are presented ok, with room for improvement

## Discussion/Conclusions
Are the study findings properly discussed? Are strengths and limitations acknowledged? Are findings interpreted properly?

### Feedback and Comments

The discussion was not included in the manuscript so I am unable to give feedback.

### Summary assessment
* major parts of discussion missing or wrong

## Further comments

Overall I think the project is great and I love the topic and dataset you all chose. You are asking great questions and have a great start to your analysis. I would only recommend making those edits to the figures and tying everything together in the discussion and results section so we can see the findings of your analysis.

# Overall project content evaluation
Evaluate overall features of the project  by filling in the sections below.

## Structure
Is the project well structured? Are files in well labeled folders? Do files have reasonable names? Are all "junk" files not needed for analysis/reproduction removed? By just looking at files and folders, can you get an idea of how things fit together?

### Feedback and Comments

The project repository is well structured and all of the files and folders are given reasonable names. There are no junk files or folders that do not need to be there. Even without the instructions in the readme, it was very easy to know how everything fit together and which code to run first just based on the names.

### Summary assessment
* well structured

## Documentation 
How well is the project documented? Are you able to understand each step of the whole analysis, each decision that was made, and each line of code? Is enough information provided as comments in code or as part of Rmd files?

### Feedback and Comments

This project could benefit from a lot more comments in the scripts. There are very few comments that explain what you are doing in the code chunks and that would be something helpful to add.

### Summary assessment
* decently documented with some gaps

## Reproducibility
Are all results fully reproducible? Is documentation provided which clearly explains how to reproduce things, and does it work without the need for any manual intervention? Are you able to re-run the whole analysis without having to do manual interventions/edits?

### Feedback and Comments

The README.md files could be more comprehensive. The one in the primary folder gives no information on the steps required to reproduce the results and the ones in the sub folders are not helpful either. All of the results are fully reproducible and there were no manual additions required for the scripts to run in their entirety. The only error I had was with the last lines in the statistical analysis (starting at line 756) which was saving the files. This did not affect any of the manuscript or figures though. I would fix that chunk so there are no errors and all results save properly.

### Summary assessment
* fully reproducible without issues

## Thoroughness
How thorough was the overall study? Were alternatives (e.g. different ways of processing the data or different models) considered? Were alternatives discussed? Were the questions/hypotheses fully and thoroughly addressed?

### Feedback and Comments

I think this is a thorough study up to this point and there were different models used. The questions and hypothesis were not thoroughly addressed as there was no discussion section and that was not addressed in the results section.

### Summary assessment
* decent level of thoroughness

## Further comments

Overall, I think this a great project so far. I would just add in the discussion and tie everything together. Great job!

### 2023-04-28 reviews/Lattner_Review.qmd
Project Review - Lattner

---
title: Wu/Chen Project Review 
author: Seth Lattner
date: "`r file.mtime(knitr::current_input())`"
format: 
  html:
    toc: true
    toc-depth: 3
    number-sections: true
---

# Overview

Title of project: Foodborne-Outbreak Project

Name of project author(s): Weifan Wu and Kai Chen

Name of project reviewer: Seth Lattner

# Specific project content evaluation
Evaluate the different parts of the project by filling in the sections below.

## Background, Context and Motivation
How well is the context of the project described? Is a comprehensive background, including summary of previous/related work given? Is the project well placed into the context of existing work (including proper referencing of existing work). Is it clear why the project was undertaken and what new information it hopes to provide?

### Feedback and Comments

The background does a good job of contextualizing the project. The literature referenced in the introduction supports the importance of the study and guides the reader towards the objectives of the report.

### Summary assessment

* strong contextualization and motivation

## Question description

How well and clear are the question(s)/hypotheses the project aims to address described? Is it clear how the questions relate to the data?

### Feedback and Comments

The questions addressed in this manuscript are clearly defined, as are the hypotheses. Both the questions and the hypotheses seem relevant to the data, and can be answered feasibly.

### Summary assessment

* question/hypotheses fully clear

## Data description
How well is the data overall described? Is the source provided? Is a codebook or other meta-information available that makes it clear what the data is?

### Feedback and Comments

The data source is clearly described and referenced throughout the project. I was unable to find a codebook defining each variable used in the analysis. While most of the variables are easy to understand, a codebook would be useful in defining terms. However, the authors do a good job of describing the data throughout the project, especially in the processing script.

### Summary assessment

* source and overall structure of data somewhat explained

## Data wrangling and exploratory analysis
How well is the data cleaned/processed and explored? Are all steps reasonable and well explained? Are alternatives discussed and considered? Are meaningful exploratory results shown (e.g. in the supplementary materials)?

### Feedback and Comments

Data wrangling was clearly explained and done very well. The authors were very thorough in cleaning and preparing the data for analysis and did a great job. The figures created in the exploratory analysis were impressive, and the authors clearly did their due diligence exploring and preparing the data. My only suggestion would be to add in a supplementary materials file where a lot of these exploratory results can be stored and shared.

### Summary assessment

* essentially no weaknesses in wrangling and exploratory component

## Appropriateness of Analysis

Were the analysis methods appropriate for the data? Was the analysis done properly? Were different components of the analysis (e.g. performance measure, variable selection, data pre-processing, model evaluation) done in the best way possible and explained well?

### Feedback and Comments

I found the statistical analysis to be appropriate based on the data at hand. The response and predictor variables seemed appropriate for the questions being answered. Multiple modeling approaches were implemented and compared. Overall, the analysis was done very well.

### Summary assessment

* strong and reasonable analysis

## Presentation
How well are results presented? Are tables and figures easy to read and understand? Are the main figures/tables publication level quality?

### Feedback and Comments

Results are presented clearly and effectively. Figures and tables are professional and informative. Numbering and captioning all figures would be helpful in keeping up with what is found in the written results, however.

### Summary assessment

* results are very well presented

## Discussion/Conclusions
Are the study findings properly discussed? Are strengths and limitations acknowledged? Are findings interpreted properly?

### Feedback and Comments

The discussion and interpretation of the results is very thorough. Implications of the findings are addressed, and conclusions are drawn appropriately. More detail could be given in limitations resulting from the study design, available data, etc. Discussion about the modeling approaches was especially interesting.

### Summary assessment

* strong, complete and clear discussion

## Further comments

_Add any other comments regarding the different aspects of the project here. Write anything you think can help your classmate improve their project._

Great work!

# Overall project content evaluation
Evaluate overall features of the project  by filling in the sections below.

## Structure
Is the project well structured? Are files in well labeled folders? Do files have reasonable names? Are all "junk" files not needed for analysis/reproduction removed? By just looking at files and folders, can you get an idea of how things fit together?

### Feedback and Comments

For the most part the structure is straightforward and clear. However, there were a few things that I found to be confusing. In the analysis code folder, there are various R, Rmd, and Qmd files that appear to do similar things, but not quite the same thing. Several "junk" files are still found in the project folders. The manuscript files are also confusing to me. There are 2 differnt manuscript Qmd files that have slightly different output and material. Neither currently renders into the Word doc that is in the manuscript folder. I am basing this review of the .html output, as that seems to be the most up-to-date.

### Summary assessment

* mostly clear, but some confusing parts (e.g. useless files, things in the wrong folders)

## Documentation 
How well is the project documented? Are you able to understand each step of the whole analysis, each decision that was made, and each line of code? Is enough information provided as comments in code or as part of Rmd files?

### Feedback and Comments

Comments are descriptive and helpful. The documentation is clear and explains each line of code, as well as justifying decisions in data processing and analysis.

### Summary assessment

* fully and well documented

## Reproducibility
Are all results fully reproducible? Is documentation provided which clearly explains how to reproduce things, and does it work without the need for any manual intervention? Are you able to re-run the whole analysis without having to do manual interventions/edits?

### Feedback and Comments

I was able to fully reproduce the analysis without any issues. Removing some of the superfluous files would make it even easier, however, the documentation was sufficient for me to follow their process.

### Summary assessment

* fully reproducible without issues

## Thoroughness
How thorough was the overall study? Were alternatives (e.g. different ways of processing the data or different models) considered? Were alternatives discussed? Were the questions/hypotheses fully and thoroughly addressed?

### Feedback and Comments

This study was very thoroughly conducted throughout the whole process. Ample documentation, description, and discussion was provided both in the code and the manuscript. Several modeling approaches were taken and compared, and the questions and hypotheses were fully addressed.

### Summary assessment

* strong level of thorougness

## Further comments

_Add any other comments regarding the overall project here. Write anything you think can help your classmate improve their project._

Overall a really good job! Tidying up the directory would be helpful, but otherwise it is in great shape.

### 2023-04-28 reviews/Project_Review_DeannaL.qmd
combine two reviews into review folder

--- 2023-04-28 Project_Review_DeannaL.qmd
@@
 
-The background does a good job of describing your topic. I would have liked to see more on previous or related work, or a statement saying that there is no previous work prior to this study.
+The background does a good job of describing your topic and the previous related studies.
+There a relevant references and it is well documented why the project was undertaken and the new information it will provide
 
 ### Summary assessment (PICK ONE, DELETE THE OTHERS)
-* some contextualization and motivation
+* strong contextualization and motivation
 
@@
 I especially liked the Pareto Chart with Missing Values figures in your processing script.
-While I do think there could have been more documentation for the steps you took to clean with some rational, I think all of the steps were reasonable.
+I think all of the steps were reasonable.
 For the exploratory analysis, the figures look great and all the results presented are related to the overarching study.
-I would like to have seen some sentences to explain the exploration and the figures provided.
-There were no 'takeaways' from the exploratory analysis just by reading the manuscript.
-The sentences found at the end of your exploratory analysis code would be a great addition to the manuscript to just give some general overview of what you are visualizing and the importance.
+The meaningful results from the exploration are explained and documented.
 
@@
 Great analysis!
+You started with explain the findings from our exploratory analysis and linked that to the rational behind your statistical analysis which was very helpful.
 You this statistical analysis was very appropriate for the questions and hypothesis and you utilized multiple models and included an great model interpretation with some performance metrics.
-I would help to include some model interpretation that relates back to the questions and hypothesis.
+The model results were also explained well and additional preprocessing steps were described.
 
@@
 
-The figures are off to a great start.
-For all of you figures, I would increase the size of the text for the title, axis labels, axis titles and legend.
-There are too small to read without having to zoom in.
-Additionally, for figures 2 and 3, I would make the axis labels at an angle so that they are not overlapping.
-THe overlap makes the figure difficult to read.
+The figures are well presented.
+They include legible titles, descriptions, and are relevant to the analysis.
+The axis titles could be slightly larger but overall, they are great quality.
+I would recommend making the titles of your figure have consistent formatting.
+Some include the figure number and others do not and the last figure did not include a title at all.
+For the tables, I would recommend having all of the captions and formatting done in a different .qmd or .r file so that we do not see the code you used to add the caption and section criteria.
 
@@
 
-The discussion was not included in the manuscript so I am unable to give feedback.
+The findings are proper discussed.
+The hypothesis stated earlier in the manuscript were mentioned in the discussion as to whether the analysis supported or did not support them.
+I also liked that other findings from previous studies were included and related to the results from this project.
+There was a dedicated section for strengths and limitations, however it seemed to focus more on future directions of the study.
+You should definitely keep the future directions, but add the strengths and limitations.
+The conclusion was concise and all of the findings seem to be interpreted properly.
 
 ### Summary assessment
-* major parts of discussion missing or wrong
+* strong, complete and clear discussion
 
@@
 Overall I think the project is great and I love the topic and dataset you all chose.
-You are asking great questions and have a great start to your analysis.
-I would only recommend making those edits to the figures and tying everything together in the discussion and results section so we can see the findings of your analysis.
+This is a great analysis and it seems like a lot of consideration was put into the methods of analysis and the model interpretation.
+Great work!
 
@@
 
-I think this is a thorough study up to this point and there were different models used.
-The questions and hypothesis were not thoroughly addressed as there was no discussion section and that was not addressed in the results section.
+This was a thorough study overall.
+Four different models were compared for the analysis and they used the performance metrics to determine the best model for analysis.
+The questions and hypotheses were all thoroughly addressed in the discussion and conclusion section.
+They found that 2 of the 3 hypotheses were not supported after the analysis and all of the questions were addressed through out the project.
 
 ### Summary assessment
-* decent level of thoroughness
+* strong level of thorougness
 
@@
 
-Overall, I think this a great project so far. I would just add in the discussion and tie everything together. Great job!
+Overall this is a great analysis and very comprehensive.
+My only suggestions would be to adjust your axis labels and title sizes and to incorporate more descriptive README.md files to make reproducing your results more easier.
+Great work!

### 2023-05-05 code/analysis_code/exploratory_analysis.qmd
cleanup

@@
   labs(x = "Pathogen Type", y = "Hospitalizations", 
-       title = "Bar Chart of Hospitalizations by Pathogen type") +
+       title = "Bar Chart of Hospitalizations by Pathogen Type") +
   theme_classic() + 

### 2023-05-05 products/manuscript/Manuscript2.qmd
cleanup

@@
 
-![Figure 3.
-Pathogen types and Hospitalizations.](images/pathogen_type_v_hospitalizations.png) ![Figure 4.
-Common bacterial etiology causing hospitalizations from 1998-2015.](images/bacterial_etiology_v_hospitalization_frequency.png) ![Figure 5.
-Top 3 bacterial pathogens and their common food carrier month from 1998-2015.](images/top3_bacterial_etiology_v_food_hospitalization_frequency.png) ![Figure 5.
-Leading locations that result in hospitalizations](images/Location_v_hospitalization.png) ![Figure 6.
-Illness over time by pathogen type (2009-2020).](images/time_v_illness_pathogen_type.png) ![Figure 7.
-Illnesses by outbreak location (2009-2020).](images/Location_v_Illness.png) ![Figure 8.
-Bacterial etiology by outbreak frequency (2009-2020).](images/bacterial_etiology_v_outbreak_frequency.png) ![Figure 9.
-Food by outbreak frequency (2009-2020).](images/food_type_v_illness_frequency.png)
+![Figure 3. Pathogen types and Hospitalizations.](images/pathogen_type_v_hospitalizations.png)
+
+![Figure 4. Common bacterial etiology causing hospitalizations from 1998-2015.](images/bacterial_etiology_v_hospitalization_frequency.png)
+
+![Figure 5. Top 3 bacterial pathogens and their common food carrier month from 1998-2015.](images/top3_bacterial_etiology_v_food_hospitalization_frequency.png)
+
+![Figure 5. Leading locations that result in hospitalizations](images/Location_v_hospitalization.png)
+
+![Figure 6. Illness over time by pathogen type (2009-2020).](images/time_v_illness_pathogen_type.png)
+
+![Figure 7. Illnesses by outbreak location (2009-2020).](images/Location_v_Illness.png)
+
+![Figure 8. Bacterial etiology by outbreak frequency (2009-2020).](images/bacterial_etiology_v_outbreak_frequency.png)
+
+![Figure 9. Food by outbreak frequency (2009-2020).](images/food_type_v_illness_frequency.png)
+
 ![Figure 10. Variable importance in determining hospitalization predictions using random forest.](images/variable importance in predicting hospitalizations by random forest model.png)

### 2023-05-05 code/analysis_code/statistical_analysis.R

@@ -1,55 +1,0 @@
-###############################
-# analysis script
-#
-#this script loads the processed, cleaned data, does a simple analysis
-#and saves the results to the results folder
-
-#load needed packages. make sure they are installed.
-library(ggplot2) #for plotting
-library(broom) #for cleaning up output from lm()
-library(here) #for data loading/saving
-
-#path to data
-#note the use of the here() package and not absolute paths
-data_location <- here::here("data","processed_data","processeddata.rds")
-
-#load data. 
-mydata <- readRDS(data_location)
-
-
-######################################
-#Data fitting/statistical analysis
-######################################
-
-############################
-#### First model fit
-# fit linear model using height as outcome, weight as predictor
-
-lmfit1 <- lm(Height ~ Weight, mydata)  
-
-# place results from fit into a data frame with the tidy function
-lmtable1 <- broom::tidy(lmfit1)
-
-#look at fit results
-print(lmtable1)
-
-# save fit results table  
-table_file1 = here("results", "resulttable1.rds")
-saveRDS(lmtable1, file = table_file1)
-
-############################
-#### Second model fit
-# fit linear model using height as outcome, weight and sex as predictor
-
-
-# place results from fit into a data frame with the tidy function
-lmtable2 <- broom::tidy(lmfit2)
-
-#look at fit results
-print(lmtable2)
-
-# save fit results table  
-table_file2 = here("results", "resulttable2.rds")
-saveRDS(lmtable2, file = table_file2)
-
-  
